---
title: "qPCR data QC"
author: Laura Vermeren & Symul
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: refresh # true refresh false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---


```{r}
#| warning: false

library(tidyverse)
library(magrittr)
library(gt)
library(patchwork)
library(SummarizedExperiment)
library(tidySummarizedExperiment)

tmp <- fs::dir_map("R scripts/", source)

theme_set(theme_light())

```

## Loading the data


```{r}

data_source <- "real"
qpcr_dir <- str_c(get_data_dir(data_source), "00 Raw/03 qPCR/")
file <- "VIBRANT_qPCR_data_merged_250416.csv"
cat("We load the file", file, "\n\n")
qpcr <- read_csv(str_c(qpcr_dir, file))

```


The data is provided in long format:

```{r}

qpcr |> glimpse()

```


- `Data_row` is the row number
- `Well` is the well number from `r min(qpcr$Well)` to `r max(qpcr$Well)`
- `Fluor` is the fluorescent dye used (`r qpcr$Fluor |> unique() |> str_c(collapse = ", ")`)
- `Content` ???
- `Replicate` is the replicate number from `r min(qpcr$Replicate, na.rm = TRUE)` to `r max(qpcr$Replicate, na.rm = TRUE)` (with `r sum(is.na(qpcr$Replicate))` missing values). Values are repeated 495, or 990 times

- `Sample` is a sample ID --- ?

- `Cq` is the quantification cycle (Cq) value.
- `Starting Quantity (SQ)` is ???
- `Cq Mean` is the mean Cq value for the replicate.


- `plate_id` is the qPCR plate ID; there are `r qpcr$plate_id |> unique() |> length()` unique plate IDs in total.

- `Plate_number` is the **extraction** plate number; there are `r qpcr$Plate_number |> unique() |> length()` unique plate numbers in total (5 identical `plate_id` per `Plate_number`).

- `gDNA_Plate_ID` is the gDNA plate ID; there are `r qpcr$gDNA_Plate_ID |> unique() |> length()` unique gDNA plate IDs in total (match the `Plate_number`).

- `VMRC_Group` is the VMRC group; there are `r qpcr$VMRC_Group |> unique() |> length()` unique VMRC groups in total (`r qpcr$VMRC_Group |> unique()  |> sort() |> str_c(collapse = ", ")`).

- `Group_Fluor` combines the `Fluor` and `VMRC_Group` columns

- `Target` is the target "gene" (taxa/strain); there are `r qpcr$Target |> unique() |> length()` unique targets in total (`r qpcr$Target |> unique()  |> sort() |> str_c(collapse = ", ")`).


```{r}

qpcr |> 
  dplyr::count(Fluor, VMRC_Group, Group_Fluor, Target)

```

For now, we only have data for the 15 LBP strains.

- `gDNA_Plate_Well` is the gDNA plate well; there are `r qpcr$gDNA_Plate_Well |> unique() |> length()` unique gDNA plate wells in total in the format `[gDNA_Plate_ID]_[Well]` (*e.g.*, `r qpcr$gDNA_Plate_Well[10]`).

- `VIBR_Sample_ID` is the VIBRANT sample ID; there are `r qpcr$VIBR_Sample_ID |> unique() |> length()` unique VIBRANT sample IDs in total.

- `Dilution_Factor` is the dilution factor; there are `r qpcr$Dilution_Factor |> unique() |> length()` unique dilution factors in total (`r qpcr$Dilution_Factor |> unique()  |> sort() |> str_c(collapse = ", ")`).

- `Quant_Adjusted` is the adjusted quantification value; they range from `r min(qpcr$Quant_Adjusted)` to `r max(qpcr$Quant_Adjusted)`.

- `Copies_per_swab` is the number of copies per swab; they range from `r min(qpcr$Copies_per_swab)` to `r max(qpcr$Copies_per_swab)`.

### Tidy names

We tidy the column names for consistency and readability.

```{r}

qpcr <- qpcr |> janitor::clean_names()
qpcr <- qpcr |> 
  dplyr::rename(
    pcr_plate_id = plate_id,
    ext_lib_plate_nb = plate_number,
    ext_lib_plate_id = g_dna_plate_id
  )

colnames(qpcr)
```


### Sample types and unique sample IDs

We define a `sample_type` variable to summarize information from `vibr_sample_id` and `sample` columns.

```{r}

qpcr <-  
  qpcr |> 
  mutate(
    sample_type = 
      case_when(
        str_detect(vibr_sample_id, "#N/A") & str_detect(sample, "std") ~ "Standard",
        str_detect(vibr_sample_id, "#N/A") & str_detect(sample, "water") ~ "Water",
        str_detect(vibr_sample_id, "EQ") ~ "Control sample",
        str_detect(vibr_sample_id, "empty") ~ "Empty",
        TRUE ~ "VIBRANT clinical sample"
      )
  )

```

```{r}

qpcr |> dplyr::count(sample_type) |> arrange(-n)

```



```{r}

tmp <- 
  qpcr |> 
  filter(sample_type == "VIBRANT clinical sample", target == target[1]) |>
  group_by(vibr_sample_id) |> 
  summarize(
    n_replicates = n(),
    n_plates = ext_lib_plate_nb |> unique() |> length()
  )

# tmp |> dplyr::count(n_plates) # All samples's replicates are on a single plate

# tmp |> dplyr::count(n_replicates) # All samples have 3 replicates

```

We note that all samples's replicates are on a single extraction plate (`ext_lib_plate_nb` or `ext_lib_plate_id`); and all samples have the same number of replicates (3).


The plate layout is as follows:

```{r}

qpcr <- 
  qpcr |> 
  mutate(
    well_col = well |> str_sub(1, 1), 
    well_row = well |> str_sub(2,3) |> as.integer()
    ) |> 
  relocate(well_col, well_row, .after = well)

```

```{r}
#| fig-width: 17
#| fig-height: 7

qpcr |> 
  select(ext_lib_plate_nb, well_col, well_row, sample_type, sample) |>
  distinct() |>
  ggplot() +
  aes(x =  well_row |> factor(), y = well_col |> factor() |> fct_rev(), fill = sample_type) +
  geom_tile() +
  geom_path(aes(group = sample), col = "black", alpha = 0.5) +
  geom_point() +
  facet_wrap(~ ext_lib_plate_nb) +
  xlab("Well column") + ylab("Well row") +
  labs(caption = "Black lines connect replicates") 
  
```

We also define a unique sample ID according to each sample type.

```{r}

qpcr <- 
  qpcr |> 
  mutate(
    sample_id = 
      case_when(
        (sample_type == "VIBRANT clinical sample") ~ vibr_sample_id,
        (sample_type == "Control sample") ~ vibr_sample_id,
        (sample_type == "Empty") ~ 
          str_c(vibr_sample_id,"_plate_", 
                ext_lib_plate_nb |> str_pad(width = 2, pad = 0),"_", sample),
        (sample_type == "Water") ~ 
          str_c("water_plate", 
                ext_lib_plate_nb |> str_pad(width = 2, pad = 0)),
        (sample_type == "Standard") ~ 
          str_c("std_", content |> str_remove("Std-"), 
                "_plate_", ext_lib_plate_nb |> str_pad(width = 2, pad = 0)),
        TRUE ~ NA_character_
      )
  ) |> 
  group_by(sample_id, target) |>
  arrange(well_col, well_row) |>
  mutate(replicate_nb = row_number()) |> 
  ungroup() |> 
  mutate(uid = str_c(sample_id, "_r", replicate_nb))

```

### Add metagenomics manifest to link sample ID to participant ID and visit code.

> TODO





### Target's plate layout and fluorescent probes

The plate x Fluorescence x Target layout is as follows:

```{r}
#| fig-width: 10
#| fig-height: 4

qpcr |> 
  dplyr::count(fluor, target, vmrc_group, pcr_plate_id, ext_lib_plate_nb) |> 
  arrange(ext_lib_plate_nb, vmrc_group, pcr_plate_id, fluor) |> 
  mutate(
    target = target |> fct_inorder(), 
    plate_nb = str_c("plate nb: ", ext_lib_plate_nb) |> fct_inorder(),
    pcr_plate_id = pcr_plate_id |> fct_inorder()
    ) |> 
  ggplot() +
  aes(
    x = pcr_plate_id,
    y = target |> factor() |> fct_rev(),
    fill = fluor
  ) +
  geom_tile() +
  facet_grid(vmrc_group ~ plate_nb, scales = "free", space = "free") +
  xlab("Plate ID") +
  ylab("Target") +
  scale_fill_manual(
    name = "Fluor.",
    values = c("FAM" = "dodgerblue1", "HEX" = "green3", "Cy5" = "#F8766D")
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  )

```
Potential batch effects due to extraction should be checked for `ext_lib_plate_id` and those due to the qPCR itself using the `pcr_plate_id`.

### Dilution factors ?

```{r}

qpcr |> 
  dplyr::count(target, dilution_factor) 

```


```{r}
#| fig-height: 7
#| fig-width: 15

qpcr |> 
  select(
    starts_with("well"), 
    pcr_plate_id, 
    vmrc_group, 
    ext_lib_plate_nb, 
    dilution_factor
    ) |> 
  distinct() |> 
  ggplot() +
  aes(
    x = well_row |> factor(), 
    y = well_col |> fct_rev(), 
    fill = dilution_factor |> factor()
    ) +
  geom_tile() +
  facet_grid(vmrc_group ~ ext_lib_plate_nb, scales = "free", space = "free") +
  scale_fill_manual(
    "Dilution factor", 
    values = colorRampPalette(c("black", "dodgerblue1"))(4)
    ) +
  xlab("Well column") + ylab("Well row") 

```



```{r}
#| fig-height: 15
#| fig-width: 15

qpcr |> 
  select(
    starts_with("well"), 
    pcr_plate_id, 
    target,
    vmrc_group, 
    ext_lib_plate_nb, 
    copies_per_swab
    ) |> 
  distinct() |> 
  ggplot() +
  aes(
    x = well_row |> factor(), 
    y = well_col |> fct_rev(), 
    fill = copies_per_swab |> asinh()
    ) +
  geom_tile() +
  facet_grid(vmrc_group + target ~ ext_lib_plate_nb, scales = "free", space = "free") +
  xlab("Well column") + ylab("Well row") 

```

```{r}
#| fig-height: 15
#| fig-width: 15

qpcr |> 
  select(
    starts_with("well"), 
    pcr_plate_id, 
    target,
    vmrc_group, 
    ext_lib_plate_nb, 
    cq
    ) |> 
  distinct() |> 
  ggplot() +
  aes(
    x = well_row |> factor(), 
    y = well_col |> fct_rev(), 
    fill = cq
    ) +
  geom_tile() +
  facet_grid(vmrc_group + target ~ ext_lib_plate_nb, scales = "free", space = "free") +
  xlab("Well column") + ylab("Well row") 

```


```{r}
#| fig-height: 3.5
#| fig-width: 12

tmp <- 
  qpcr |> 
  select(
    pcr_plate_id, 
    target,
    vmrc_group, 
    ext_lib_plate_nb, 
    cq,
    copies_per_swab,
    sample_type,
    sample_id,
    dilution_factor,
    quant_adjusted
    ) |> 
  filter(sample_type == "Standard", vmrc_group %in% c("VMRC_1","VMRC_2")) |>
  mutate(
    std = 
      sample_id |> 
      str_remove("_plate_([0-9]+)") |> 
      str_remove("std_") |> 
      as.integer()
  ) |> 
  distinct() 

tmp |> 
  ggplot() +
  aes(
    x = std, 
    y = cq, 
    col = ext_lib_plate_nb %in% c(1:3)
    ) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ vmrc_group + target, scales = "free", space = "free")


tmp |> 
  ggplot() +
  aes(
    x = std, 
    y = cq, 
    col = dilution_factor |> factor()
    ) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ vmrc_group + target, scales = "free", space = "free")

```

```{r}

tmp |> 
  ggplot() +
  aes(
    x = std, 
    y = copies_per_swab, 
    col = ext_lib_plate_nb %in% c(1:3)
    ) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ vmrc_group + target, scales = "free", space = "free") +
  scale_y_log10()


tmp |> 
  ggplot() +
  aes(
    x = std, 
    y = copies_per_swab, 
    col = dilution_factor |> factor()
    ) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ vmrc_group + target, scales = "free", space = "free") +
  scale_y_log10()

```
```{r}

tmp |> dplyr::count(std, dilution_factor, copies_per_swab)

```



```{r}

tmp |> 
  ggplot() +
  aes(
    x = std, 
    y = quant_adjusted, 
    col = ext_lib_plate_nb %in% c(1:3)
    ) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ vmrc_group + target, scales = "free", space = "free") +
  scale_y_log10()


tmp |> 
  ggplot() +
  aes(
    x = std, 
    y = quant_adjusted, 
    col = dilution_factor |> factor()
    ) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ vmrc_group + target, scales = "free", space = "free") +
  scale_y_log10()

```
```{r}

tmp |> dplyr::count(std, dilution_factor, copies_per_swab, quant_adjusted)

```

```{r}

tmp |> 
  ggplot() +
  aes(
    x = cq, 
    y = copies_per_swab, 
    col = ext_lib_plate_nb %in% c(1:3)
    ) +
  geom_text(aes(label = std), alpha = 0.5) +
  facet_grid(. ~ vmrc_group + target) +
  scale_y_log10()


tmp |> 
  ggplot() +
  aes(
    x = cq, 
    y = copies_per_swab, 
    col = dilution_factor |> factor()
    ) +
  geom_text(aes(label = std), alpha = 0.5) +
  facet_grid(. ~ vmrc_group + target) +
  scale_y_log10()

```


## Exploratory & QC analyses

### Total number of copies per swab per sample type


```{r}
#| fig-width: 8
#| fig-height: 5

g <- 
  qpcr |> 
  ggplot() +
  aes(x = target, y = copies_per_swab, col = sample_type, fill = sample_type) +
  geom_boxplot(alpha = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

g 

g + scale_y_log10()

```


```{r}
#| fig-width: 8
#| fig-height: 5

g <- 
  qpcr |> 
  ggplot() +
  aes(x = target, y = copies_per_swab, col = sample_type, fill = sample_type) +
  geom_boxplot(alpha = 0.5) +
  facet_grid(sample_type ~ ., scales = "free") +
  guides(fill = "none", color = "none") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  )

g 

g + scale_y_log10()

```


```{r}
#| fig-width: 8
#| fig-height: 5


g <- 
  qpcr |> 
  ggplot() +
  aes(x = ext_lib_plate_nb |> factor(), y = copies_per_swab, col = target, fill = target) +
  geom_boxplot(alpha = 0.5) +
  facet_grid(sample_type ~ ., scales = "free") +
  xlab("Extraction plate number") +
  theme(
    strip.text.y = element_text(angle = 0),
    strip.text.x = element_text(angle = 90, hjust = 0)
  )

g 

g + scale_y_log10() 

```


### Empty and water samples

From the plot above, it looks like the empty and water samples have non-zero values on a few plates.


```{r}

plot_qpcr_empties <- function(qpcr, color_by = "sample_type") {
  qpcr |> 
    filter(sample_type %in% c("Empty", "Water")) |> 
    group_by(sample, pcr_plate_id, ext_lib_plate_nb, target) |>
    mutate(replicate_nb = row_number()) |> 
    ungroup() |> 
    ggplot() +
    aes(x = sample, y = copies_per_swab, label = replicate_nb, col = !!sym(color_by)) +
    geom_text(size = 3) +
    facet_grid(vmrc_group + target ~ ext_lib_plate_nb, scales = "free_x", space = "free_x") +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text.y = element_text(angle = 0)
    ) 
}

```


```{r}
#| fig-width: 10
#| fig-height: 6

plot_qpcr_empties(qpcr, color_by = "sample_type") 

plot_qpcr_empties(qpcr |> mutate(zero_copies = (copies_per_swab == 0)), color_by = "zero_copies") 

```

```{r}

prob_target <-
  qpcr |> 
  select(target) |> 
  distinct() |> 
  mutate(prob_target = (target %in% c("C0022A1", "C0059E1", "C0175A1")))
  
```


::: callout-warning
We may have to check the values for targets `r prob_target$target[prob_target$prob_target] |> str_c(collapse = ", ")`.
:::

```{r}

prob_target <- 
  prob_target |> 
  mutate(prob_target = if_else(prob_target, "x", "v")) 

qpcr <- 
  qpcr |> 
  left_join(prob_target, by = "target") 
  
```




### Control samples


```{r}
#| fig-width: 10
#| fig-height: 6

qpcr |> 
  filter(sample_type %in% c("Control sample")) |> 
  group_by(sample, pcr_plate_id, ext_lib_plate_nb, target) |>
  mutate(replicate_nb = row_number()) |> 
  ungroup() |> 
  ggplot() +
  aes(x = vibr_sample_id, y = copies_per_swab, label = replicate_nb, col = replicate_nb |> factor()) +
  geom_text(size = 3) +
  scale_y_log10() +
  scale_color_discrete("replicate") +
  facet_grid(prob_target + target ~ ext_lib_plate_nb, scales = "free_x", space = "free_x") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  ) 

```

From these plots, I assume that some of these controls are negative controls and others are positive controls.

> TODO: check if they match the sampleID from the metagenomics data or if we need a manifest.

Replicability looks good, but we note the same issues with the problematic targets in the (likely) negative controls as in the empty samples.



### Standards

> TODO

### Clinical samples


```{r}
#| fig-width: 15
#| fig-height: 10

qpcr |> 
  filter(sample_type %in% c("VIBRANT clinical sample")) |> 
  group_by(vibr_sample_id, sample, pcr_plate_id, ext_lib_plate_nb, target) |>
  mutate(
    replicate_nb = row_number(),
    median_cps = median(copies_per_swab, na.rm = TRUE),
    mean_cps = mean(copies_per_swab, na.rm = TRUE)
    ) |> 
  ungroup() |> 
  mutate(vibr_sample_id = vibr_sample_id |> fct_reorder(mean_cps)) |>
  ggplot() +
  aes(x = vibr_sample_id, y = copies_per_swab, label = replicate_nb, col = replicate_nb |> factor()) +
  geom_line(aes(group = vibr_sample_id), col = "black", linewidth = 0.1) +
  geom_text(size = 3) +
  scale_y_log10() +
  scale_color_discrete("replicate") +
  scale_x_discrete("Samples", breaks = NULL) +
  facet_grid(prob_target + target ~ ext_lib_plate_nb, scales = "free_x", space = "free_x") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  )

```
We notice the same issues with the problematic targets in plate 1-3 (bottom, left).




These three targets are from the same plates (`plate_id` column):

```{r}

qpcr |> 
  filter(prob_target == "x", ext_lib_plate_nb %in% 1:3) |> 
  select(target, fluor, pcr_plate_id, ext_lib_plate_nb) |> 
  distinct() |> 
  gt()

```

```{r}
#| fig-width: 15
#| fig-height: 10

qpcr |> 
  filter(prob_target == "x", ext_lib_plate_nb %in% 1:3) |> 
  select(target, fluor, pcr_plate_id, ext_lib_plate_nb, starts_with("well"), sample, vibr_sample_id, sample_type, copies_per_swab) |> 
  arrange(sample_type) |>
  ggplot() +
  aes(x = well_row |> factor(), y = well_col |> fct_rev(), fill = sample_type) +
  geom_tile() +
  geom_point(aes(size = copies_per_swab |> log10())) +
  geom_line(aes(group = sample), col = "black", alpha = 0.5) +
  facet_grid(target ~ ext_lib_plate_nb) +
  coord_fixed() +
  xlab("Well column") + ylab("Well row") +
  labs(caption = "Black lines connect replicates") 

```



```{r}
#| fig-width: 15
#| fig-height: 10

qpcr |> 
  filter(prob_target == "x", ext_lib_plate_nb %in% 1:4) |> 
  select(target, fluor, pcr_plate_id, ext_lib_plate_nb, starts_with("well"), sample, vibr_sample_id, sample_type, dilution_factor, copies_per_swab) |> 
  arrange(sample_type) |>
  ggplot() +
  aes(x = well_row |> factor(), y = well_col |> fct_rev(), fill = sample_type) +
  geom_tile(aes(alpha = dilution_factor)) +
  geom_point(aes(size = copies_per_swab |> log10())) +
  geom_line(aes(group = sample), col = "black", alpha = 0.5) +
  facet_grid(target ~ ext_lib_plate_nb) +
  coord_fixed() +
  xlab("Well column") + ylab("Well row") +
  labs(caption = "Black lines connect replicates") 

```



There isn't a clear "geometrical" pattern on the plate, nor a clear dilution factor association. Let's just assume that these three plates failed.


::: callout-warning
We could simply discard these values (exclude these targets x plates) or take the minimum when aggregating across replicates (probably not a good idea for the remaining good plates).
:::




```{r}
#| fig-width: 15
#| fig-height: 10

qpcr |> 
  filter(prob_target == "x", ext_lib_plate_nb %in% 1:4) |> 
  select(target, fluor, pcr_plate_id, ext_lib_plate_nb, starts_with("well"), sample, vibr_sample_id, sample_type, dilution_factor, copies_per_swab, cq) |> 
  arrange(sample_type) |>
  ggplot() +
  aes(x = well_row |> factor(), y = well_col |> fct_rev(), fill = sample_type) +
  geom_tile(aes(alpha = dilution_factor)) +
  geom_point(aes(size = cq |> log10())) +
  geom_line(aes(group = sample), col = "black", alpha = 0.5) +
  facet_grid(target ~ ext_lib_plate_nb) +
  coord_fixed() +
  xlab("Well column") + ylab("Well row") +
  labs(caption = "Black lines connect replicates") 

```


## Creating `SummarizedExperiment` objects

We create two `Summarized Experiment` objects:

- one with all values as provided in the original file, where each "sample" is a plate well and each feature is a target (LBP taxa), with the following assays

    + cq values (`cq`)
    + `starting_quantity_sq`
    + `cq_mean`
    + `quant_adjusted`
    + `copies_per_swab_raw` (current `copies_per_swab` column)
    + `copies_per_swab` (where "problematic" targets x plates are set to NA)
    
    The `colData` contains the following columns: 
        
        + `uid`
        + `sample_id`
        + `sample`
        + `vibr_sample_id`
        + `sample_type`
        + `ext_lib_plate_nb` 
        + `ext_lib_plate_id`
        + (`dilution_factor`)
        
    The `rowData` contains the following columns:
    
        + `fluor`
        + `pcr_plate_id` (the concatenation of `pcr_plate_id` for each `ext_lib_plate_nb`)
        + `valid_plates` (the `ext_lib_plate_nb` where signal looked good)
        
        
- one with summarized values (aggregated across replicates) where each "sample" is a VIBRANT sample ID (including control samples) and each feature is a target (LBP taxa), with the following assays

    + `copies_per_swab_med` (the median `copies_per_swab` across replicates)
    + `copies_per_swab_mean` (the mean `copies_per_swab` across replicates)
    + `copies_per_swab_cv` (the coefficient of variation of the `copies_per_swab` across replicates)

    The `colData` contains the following columns: 
        
        + `sample_type`
        + `vibr_sample_id`
        + `sample`
        + `ext_lib_plate_nb` 
        + `ext_lib_plate_id`
        + `dilution_factor`
        
    The `rowData` contains the following columns:
    
        + `fluor`
        + `vmrc_group`
        + `pcr_plate_id` (the concatenation of `pcr_plate_id` for each `ext_lib_plate_nb`)
        + `valid_plates` (the `ext_lib_plate_nb` where signal looked good)

        
### Raw `SummarizedExperiment` object


We first identify potential issues with qPCR plates if the `copies_per_swab` values are not zero for the empty and water wells.

```{r}
#| fig-height: 4
#| fig-width: 10

map(
  qpcr$vmrc_group |> unique(),
  ~{
    qpcr |> 
      filter(vmrc_group == .x) |> 
      filter(
        sample_type %in% c("Empty", "Water"), 
        str_detect(sample_id, "G12")|str_detect(sample_id, "H12")|str_detect(sample_id, "water")
        ) |>
      ggplot() +
      aes(x = target, y = copies_per_swab, col = target, fill = target) +
      # geom_boxplot(alpha = 0.5, outlier.shape = NA) +
      geom_point(aes(shape = sample_type), alpha = 0.3) +
      # ggbeeswarm::geom_quasirandom(alpha = 0.5) +
      scale_y_log10() + expand_limits(y = 1) +
      facet_wrap(. ~ ext_lib_plate_nb + pcr_plate_id, nrow = 2, scales = "free_x") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      guides(col = "none", fill = "none") +
      ggtitle(str_c("qPCR plates for targets of group ", .x)) 
  }
)

```
Plates are flagged as non-valid if there are more than 25% of the empty or water samples with non-zero values.

```{r}

valid_plates <- 
  qpcr |> 
  filter(
    sample_type %in% c("Empty", "Water"), 
    str_detect(sample_id, "G12")|str_detect(sample_id, "H12")|str_detect(sample_id, "water")
  ) |>
  group_by(ext_lib_plate_nb, pcr_plate_id, vmrc_group) |>
  summarize(
    n = n(),
    n_non_zero = sum(copies_per_swab > 0, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    f_non_zero = n_non_zero / n,
    valid = (f_non_zero <= 1/4) & (n_non_zero <= 2)
  ) 
  
valid_plates |> filter(!valid) |> gt(caption = "qPCR plates flagged as non-valid.")
  
```

```{r}

qpcr <- 
  qpcr |> 
  left_join(
    valid_plates |> 
      dplyr::rename(valid_qpcr_plate = valid) |>
      select(pcr_plate_id, valid_qpcr_plate) |> 
      distinct(),
    by = join_by(pcr_plate_id)
  )

```


```{r}

make_raw_qpcr_SE <- function(qpcr){
  
  qpcr <- 
    qpcr |> 
    arrange(ext_lib_plate_nb, well_col, well_row) |>
    mutate(uid = uid |> fct_inorder()) |> 
    arrange(vmrc_group, fluor) |> 
    mutate(target = target |> fct_inorder()) |> 
    arrange(uid, target)
    
  
  cq_assay <- 
    qpcr |>
    select(uid, target, cq) |> 
    arrange() |> 
    pivot_wider(names_from = uid, values_from = cq) |> 
    as.data.frame() |> 
    column_to_rownames("target") 
  
  starting_quantity_sq_assay <- 
    qpcr |>
    select(uid, target, starting_quantity_sq) |> 
    arrange() |> 
    pivot_wider(names_from = uid, values_from = starting_quantity_sq) |> 
    as.data.frame() |> 
    column_to_rownames("target")
  
   cq_mean_assay <- 
    qpcr |>
    select(uid, target, cq_mean) |> 
    arrange() |> 
    pivot_wider(names_from = uid, values_from = cq_mean) |> 
    as.data.frame() |> 
    column_to_rownames("target")
   
   quant_adjusted_assay <- 
     qpcr |>
     select(uid, target, quant_adjusted) |> 
     arrange() |> 
     pivot_wider(names_from = uid, values_from = quant_adjusted) |> 
     as.data.frame() |> 
     column_to_rownames("target")
   
   copies_per_swab_raw_assay <- 
     qpcr |>
     select(uid, target, copies_per_swab) |> 
     arrange() |> 
     pivot_wider(names_from = uid, values_from = copies_per_swab) |> 
     as.data.frame() |> 
     column_to_rownames("target")
   
  copies_per_swab_assay <- 
    qpcr |>
    mutate(copies_per_swab_valid = ifelse(valid_qpcr_plate, copies_per_swab, NA)) |> 
    select(uid, target, copies_per_swab_valid) |> 
    arrange() |> 
    pivot_wider(names_from = uid, values_from = copies_per_swab_valid) |> 
    as.data.frame() |> 
    column_to_rownames("target")
  
   coldata <- 
     qpcr |> 
     select(uid, sample_id, replicate_nb, sample, vibr_sample_id, sample_type, ext_lib_plate_nb, ext_lib_plate_id) |> 
     distinct() |> 
     arrange(uid) |> 
     as.data.frame() |> 
     mutate(rownames = uid) |> 
     column_to_rownames("rownames")
   
   rowdata <-
     qpcr |> 
     select(target, fluor, vmrc_group, pcr_plate_id, valid_qpcr_plate, ext_lib_plate_nb) |> 
     distinct() |> 
     arrange(target, ext_lib_plate_nb) |> 
     group_by(target) |> 
     mutate(
       pcr_plate_id = str_c(pcr_plate_id, " (extr. plate ", ext_lib_plate_nb |> str_pad(width = 2, pad = "0"), ")", sep = ""),
     ) |> 
     summarize(
       fluor = fluor |> unique() |> str_c(collapse = ", "),
       vmrc_group = vmrc_group |> unique() |> str_c(collapse = ", "),
       pcr_plate_ids = pcr_plate_id |> unique() |> str_c(collapse = ", "),
       non_valdid_plates = pcr_plate_id[!valid_qpcr_plate] |> unique() |> str_c(collapse = ", ")
     ) |> 
     ungroup() |> 
     as.data.frame() |> 
     mutate(rownames = target) |> 
     column_to_rownames("rownames")
   
   SE <- SummarizedExperiment(
     assays = list(
       cq = cq_assay |> as.matrix(),
       starting_quantity_sq = starting_quantity_sq_assay |> as.matrix(),
       cq_mean = cq_mean_assay |> as.matrix(),
       quant_adjusted = quant_adjusted_assay |> as.matrix(),
       copies_per_swab_raw = copies_per_swab_raw_assay |>  as.matrix(),
       copies_per_swab = copies_per_swab_assay |> as.matrix()
     ),
     colData = coldata,
     rowData = rowdata
   )
   
   SE
  
}

```


```{r}

SE_qPCR_raw <- make_raw_qpcr_SE(qpcr)

```



### Aggregated `SummarizedExperiment` object

      
- one with summarized values (aggregated across replicates) where each "sample" is a VIBRANT sample ID (including control samples) and each feature is a target (LBP taxa), with the following assays

    + `copies_per_swab_med` (the median `copies_per_swab` across replicates)
    + `copies_per_swab_mean` (the mean `copies_per_swab` across replicates)
    + `copies_per_swab_cv` (the coefficient of variation of the `copies_per_swab` across replicates)

    The `colData` contains the following columns: 
        
        + `sample_type`
        + `vibr_sample_id`
        + `sample`
        + `ext_lib_plate_nb` 
        + `ext_lib_plate_id`
        + ? `dilution_factor`
        
    The `rowData` contains the following columns:
    
        + `fluor`
        + `vmrc_group`
        + `pcr_plate_ids` (the concatenation of `pcr_plate_id` for each `ext_lib_plate_nb`)
        + `valid_plates` (the `ext_lib_plate_nb` where signal looked good)



```{r}

make_agg_qpcr_SE <- function(qpcr){
  
  qpcr <- 
    qpcr |> 
    filter(sample_type %in% c("VIBRANT clinical sample", "Control sample")) |>
    arrange(ext_lib_plate_nb, well_col, well_row) |>
    mutate(uid = uid |> fct_inorder(), sample_id = sample_id |> fct_inorder()) |> 
    arrange(vmrc_group, fluor) |> 
    mutate(target = target |> fct_inorder()) |> 
    arrange(uid, target)
    
  summarized_qpcr <- 
    qpcr |>
    group_by(sample_id, target) |> 
    summarize(
      copies_per_swab_med = median(copies_per_swab),
      copies_per_swab_mean = mean(copies_per_swab),
      copies_per_swab_sd = sd(copies_per_swab),
      .groups = "drop"
    ) |> 
    mutate(
      copies_per_swab_cv = ifelse(copies_per_swab_sd == 0, 0, copies_per_swab_sd / copies_per_swab_mean)
    ) 
  
  coldata <- 
    qpcr |> 
    select(sample_id, vibr_sample_id, sample, sample_type, ext_lib_plate_nb, ext_lib_plate_id) |>
    distinct() |> 
    arrange(sample_id) |> 
    left_join(
      qpcr |> 
        select(sample_id, pcr_plate_id, valid_qpcr_plate) |> 
        group_by(sample_id) |> 
        summarize(
          pcr_plate_ids = str_c(pcr_plate_id, collapse = ", "),
          non_valid_plate_ids = str_c(pcr_plate_id[!valid_qpcr_plate], collapse = ", ")
        ),
      by = join_by(sample_id)
    ) |> 
    as.data.frame() |> 
    mutate(rownames = sample_id) |> 
    column_to_rownames("rownames")
   
   rowdata <-
     qpcr |> 
     select(target, fluor, vmrc_group, pcr_plate_id, ext_lib_plate_nb) |> 
     distinct() |> 
     arrange(target, ext_lib_plate_nb) |> 
     group_by(target) |> 
     mutate(
       pcr_plate_id = str_c(pcr_plate_id, " (extr. plate ", ext_lib_plate_nb |> str_pad(width = 2, pad = "0"), ")", sep = ""),
     ) |> 
     summarize(
       fluor = fluor |> unique() |> str_c(collapse = ", "),
       vmrc_group = vmrc_group |> unique() |> str_c(collapse = ", "),
       pcr_plate_ids = pcr_plate_id |> unique() |> str_c(collapse = ", ")
     ) |> 
     ungroup() |> 
     as.data.frame() |> 
     mutate(rownames = target) |> 
     column_to_rownames("rownames") 
   
   SE <- SummarizedExperiment(
     assays = list(
       copies_per_swab_med = 
         summarized_qpcr |> 
         pivot_wider(id_cols = target, names_from = sample_id, values_from = copies_per_swab_med) |> 
         as.data.frame() |> 
         column_to_rownames("target") |> 
         as.matrix(),
       copies_per_swab_mean = 
         summarized_qpcr |> 
         pivot_wider(id_cols = target, names_from = sample_id, values_from = copies_per_swab_mean) |> 
         as.data.frame() |> 
         column_to_rownames("target") |> 
         as.matrix(),
       copies_per_swab_cv = 
         summarized_qpcr |> 
         pivot_wider(id_cols = target, names_from = sample_id, values_from = copies_per_swab_cv) |> 
         as.data.frame() |> 
         column_to_rownames("target") |> 
         as.matrix()
     ),
     colData = coldata,
     rowData = rowdata
   )
   
   SE
     
}

```


```{r}

SE_qPCR_agg <- make_agg_qpcr_SE(qpcr)

```


## Save `SummarizedExperiment` objects

Save the `SE` objects to disk

```{r}

saveRDS(
  SE_qPCR_raw, 
  str_c(
    get_output_dir(data_source = data_source),  
    "03_se_pcr_raw_", today() |> str_remove_all("-"), ".rds"
    )
  )


saveRDS(
  SE_qPCR_agg, 
  str_c(
    get_output_dir(data_source = data_source),  
    "03_se_pcr_agg_", today() |> str_remove_all("-"), ".rds"
    )
  )

```


