---
title: "Assembling assays into a master `MultiAssayExperiment` object"
author: Laura Symul, Laura Vermeren
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: refresh # true refresh false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| warning: false
#| cache: false

library(tidyverse)
library(magrittr)
library(gt)
library(patchwork)
library(SummarizedExperiment)
library(tidySummarizedExperiment)
library(MultiAssayExperiment)

tmp <- fs::dir_map("R scripts/", source)

theme_set(theme_light())

```


The purpose of this document is to create a `MultiAssayExperiment` object containing all assays and clinical data generated in the context of the VIBRANT trial at the participant x visit resolution (which means that duplicates were aggregated in previous steps such that for a given assay/variable, there is a single value for each participant and visit combination).

The document is organised as follow:

- First, we load and minimally transform the "clinical" data (*i.e.*, data collected through CRFs).

- Next, we load and minimally transform the `SummarizedExperiment` objects for each assay (metagenomics, qPCR, 16S amplicon sequencing, Luminex, and flow cytometry).

- Finally, we create the `MultiAssayExperiment` object by combining the clinical data and assay data. This entails several steps, including creating a `mae_colData` table that combines the clinical data and assay sample information, ensuring that the `uid`s (unique identifiers) are consistent across all data tables.

At the end, we document which assays are available (= have data) for which participants and visits.



# Clinical data (CRF data)


## CRF data (raw and QCed)

```{r}

crf_files <- 
  get_01_output_dir() |> 
  fs::dir_ls() |> 
  str_subset("/01_")

crf_clean_file <- crf_files |> str_subset("01_crf_clean") |> sort(decreasing = TRUE) |> magrittr::extract(1)
crf_raw_file <- crf_files |> str_subset("01_crf_raw") |> sort(decreasing = TRUE) |> magrittr::extract(1)
crf_dict_file <- crf_files |> str_subset("01_crf_plates_dictionary") |> sort(decreasing = TRUE) |> magrittr::extract(1)

load(crf_clean_file, verbose = TRUE)
load(crf_raw_file, verbose = TRUE)
load(crf_dict_file, verbose = TRUE)

rm(crf_clean_file, crf_raw_file, crf_dict_file)

```


## `Participants` table

```{r}

load(
  crf_files |> str_subset("01_participants_2025") |> sort(decreasing = TRUE) |> magrittr::extract(1),
  verbose = TRUE
  )
load(
  crf_files |> str_subset("01_participants_variable_dictionary") |> sort(decreasing = TRUE) |> magrittr::extract(1),
  verbose = TRUE
  )
load(
  crf_files |> str_subset("01_participant_crfs_merged") |> sort(decreasing = TRUE) |> magrittr::extract(1),
  verbose = TRUE
  )

```

## `Visits` table

```{r}

load(
  crf_files |> str_subset("01_visits_2025") |> sort(decreasing = TRUE) |> magrittr::extract(1),
  verbose = TRUE
  )

load(
  crf_files |> str_subset("01_visits_long_2025") |> sort(decreasing = TRUE) |> magrittr::extract(1),
  verbose = TRUE
  )

load(
  crf_files |> str_subset("01_visits_crfs_merged") |> sort(decreasing = TRUE) |> magrittr::extract(1),
  verbose = TRUE
  )

```


```{r}
#| eval: false
#| fig-width: 25

visits |> 
  ggplot() +
  aes(x = study_day, y = pid |> factor(), color = visit_code |> factor()) +
  geom_point() 

```

We define the `uid` (unique identifier):


```{r}

visits <- 
  visits |> 
  mutate(
    uid = str_c(pid, visit_code, sep = "_")
  ) |> 
  select(uid, everything())


```


## `Exposures` table

```{r}

load(
  crf_files |> str_subset("01_exposures_") |> sort(decreasing = TRUE) |> magrittr::extract(1),
  verbose = TRUE
  )

```



# Assays: loading the `SummarizedExperiment` objects 


```{r}

data_dir <- get_01_output_dir()
se_files <- fs::dir_ls(data_dir, regexp = "se_.*\\.rds$") |> sort(decreasing = TRUE)

```

## Metagenomics data

```{r}

mg_file <- se_files[grepl("mg", se_files)][1]
cat("Loading metagenomics data from file: ", mg_file |> basename(), "\n")
se_mg <- readRDS(mg_file)
se_mg <- check_se(se_mg)
se_mg

```

## qPCR data

```{r}

qPCR_file <- se_files[grepl("03_se_pcr_agg", se_files)][1]
cat("Loading qPCR data from file: ", qPCR_file |> basename(), "\n")
se_qPRC <- readRDS(qPCR_file)
se_qPRC <- check_se(se_qPRC)
se_qPRC

```

## qPCR data for daily swabs (generated post-unblinding)

```{r}

daily_qPCR_file <- se_files[grepl("30_se_pcr_agg", se_files)][1]
cat("Loading qPCR data for daily swabs from file: ", daily_qPCR_file |> basename(), "\n")
se_daily_qPRC <- readRDS(daily_qPCR_file)
se_daily_qPRC <- check_se(se_daily_qPRC)
se_daily_qPRC

```


## 16S rRNA amplicon sequencing

We add both the ASV-level data and the data aggregated by taxonomic assignment.

```{r}

amplicon_ASV_file <- se_files[grepl("16S_ASV_2", se_files)][1]
cat("Loading 16S rRNA amplicon sequencing ASV-level data from file: ", amplicon_ASV_file |> basename(), "\n")
se_16S_ASV <- readRDS(amplicon_ASV_file)
se_16S_ASV <- check_se(se_16S_ASV)
se_16S_ASV

```


```{r}

amplicon_file <- se_files[grepl("16S_agg_2", se_files)][1]
cat("Loading 16S rRNA amplicon sequencing data aggregated by taxonomic assignment from file: ", amplicon_file |> basename(), "\n")
se_16S <- readRDS(amplicon_file)
se_16S <- check_se(se_16S)
se_16S

```



## Luminex data


```{r}

luminex_file <- se_files[grepl("luminex_agg", se_files)][1]
cat("Loading Luminex data from file: ", luminex_file |> basename(), "\n")
se_luminex <- readRDS(luminex_file)
se_luminex <- check_se(se_luminex)
se_luminex

```


## Flow cytometry data


```{r}

flow_file <- se_files[grepl("flow", se_files)][1]
cat("Loading flow cytometry data from file: ", flow_file |> basename(), "\n")
se_flow <- readRDS(flow_file)
se_flow <- check_se(se_flow)
se_flow

```

# Creating the `MultiAssayExperiment` object

## Assay list

```{r}

assay_list <- 
  bind_rows(
    tibble(
      mae_assay_name = "mg", 
      se_object_name = "se_mg", 
      assay_label = "Metagenomic"
      ),
    tibble(
      mae_assay_name = "qPCR", 
      se_object_name = "se_qPRC", 
      assay_label = "qPCR weekly swabs"
      ),
    tibble(
      mae_assay_name = "qPCR_daily", 
      se_object_name = "se_daily_qPRC", 
      assay_label = "qPCR home swabs"
      ),
    tibble(
      mae_assay_name = "amplicon_ASV", 
      se_object_name = "se_16S_ASV", 
      assay_label = "16S rRNA amplicon seq. (ASV-level)"
    ),
    tibble(
      mae_assay_name = "amplicon", 
      se_object_name = "se_16S", 
      assay_label = "16S rRNA amplicon seq."
    ),
    tibble(
      mae_assay_name = "luminex", 
      se_object_name = "se_luminex", 
      assay_label = "Luminex"
      ),
    tibble(
      mae_assay_name = "flow", 
      se_object_name = "se_flow", 
      assay_label = "Flow cytometry"
      )
  ) |> 
  mutate(
    assay_label = assay_label |> fct_inorder()
  )

```

We will assemble the following tables (`SE` objects):

```{r}
assay_list |> gt()
```


## `mae_coldata_assays` tables


```{r}

available_samples <- 
  map(
    1:nrow(assay_list),
    function(i, assay_list){
      se <- get(assay_list$se_object_name[i])
      tibble(
        assay = assay_list$mae_assay_name[i],
        se_colname = se |> colnames(), 
        uid = se$uid, 
        sample_type = se$sample_type,
        control_type = se$control_type,
        )
    },
    assay_list = assay_list
  ) |> 
  bind_rows()

if (!all(available_samples$se_colname == available_samples$uid, na.rm = TRUE)) {
  stop("The uid column in the SE objects does not match their colnames")
}

mae_coldata_assays <-
  available_samples |>
  group_by(uid) |>
  summarize(
    n_sample_type = n_distinct(sample_type),
    n_control_type = n_distinct(control_type),
    sample_type = sample_type[1],
    control_type = control_type[1],
    assays = str_c(assay, collapse = ", ")
  )

if (any(mae_coldata_assays$n_sample_type > 1)) {
  warning("Some samples have multiple sample types")
}

if (any(mae_coldata_assays$n_control_type > 1)) {
  warning("Some samples have multiple control types")
}

mae_coldata_assays <- 
  mae_coldata_assays |> 
  select(uid, sample_type, control_type, assays) 

```

The `mae_coldata_assays` table contains the following columns:

```{r}

mae_coldata_assays |> colnames()
mae_coldata_assays |> str()

```


```{r}

mae_coldata_assays |> 
  dplyr::count(sample_type, control_type) |> 
  gt(caption = "Number of samples per sample type and control type") 
  
```

We also remove the columns `sample_type` and `control_type` from the `colData` of the individual assays to avoid conflicts when joining with the `mae@colData` table in downstream analyses.

```{r}

for (i in 1:nrow(assay_list)) {
  se <- get(assay_list$se_object_name[i])
  if (("sample_type" %in% colnames(colData(se))) | ("control_type" %in% colnames(colData(se)))) {
    colData(se) <- colData(se)[, !colnames(colData(se)) %in% c("sample_type", "control_type")]
  }
  assign(assay_list$se_object_name[i], se)
}

```


## `MAE@colData`: joining the `visits` and `mae_coldata_assays` tables



We first do a full join of the `visits` and `mae_coldata_assays` tables by `uid`. This will allow us to keep all visits, even those that do not have assay data, and to identify which visits are in the CRF data and which are in the assay data.


```{r}

mae_coldata <- 
  # we first do a full join of the CRF visits and mae_coldata_assays by uids;
  dplyr::full_join(
    visits |> select(-starts_with("specimen_collection")) |> mutate(in_CRF = TRUE),
    mae_coldata_assays |> mutate(in_assays = TRUE) |> select(in_assays, everything()),
    by = join_by(uid)
  ) |> 
  mutate(
    in_CRF = in_CRF |> replace_na(FALSE),
    in_assays = in_assays |> replace_na(FALSE)
  )

```


```{r}
#| eval: false

# We notice that there are 5 "Clinical sample" entries for which we do have assay data but that did not appear in the CRF data:

mae_coldata |> 
  filter(is.na(visit_code), sample_type == "Clinical sample") 
  
# we fix these below

```


```{r}

mae_coldata <- 
  mae_coldata |> 
  mutate(
    needs_fixing = is.na(pid) & (sample_type == "Clinical sample") & str_detect(uid, "[0-9]{9}_[0-9]{4}"),
    pid = 
      case_when(
        needs_fixing ~ str_remove(uid, "_[0-9]{4}"),
        TRUE ~ pid
      ),
    visit_code =
      case_when(
        needs_fixing ~ str_remove(uid, "[0-9]{9}_"),
        TRUE ~ visit_code
      ),
    study_day = 
      case_when(
        needs_fixing & (visit_code %in% c("0001", "0011")) ~ -7,
        needs_fixing & in_assays & (visit_code == "1000") ~ 0,
        TRUE ~ study_day
      ),
    visit_attended = 
      case_when(
        is.na(visit_attended) & needs_fixing ~ TRUE,
        TRUE ~ visit_attended
      ),
    visit_planned = 
      case_when(
        is.na(visit_planned) & needs_fixing ~ "Unplanned visit",
        TRUE ~ visit_planned
      ),
    visit_type = 
      case_when(
        is.na(visit_type) & needs_fixing ~ "Clinic",
        TRUE ~ visit_type
      )
  ) 

```

We then merge that table with the `participants` table to add the participant-invariant information (e.g., `randomized`, `mITT`, etc.) to the `mae_coldata` table.


```{r}


# we add the participant-invariant information from the `participants` table
mae_coldata <- 
  mae_coldata |> 
  dplyr::left_join(
    participants,
    by = join_by(pid)
  )

```

```{r}
mae_coldata <- mae_coldata |> select(-needs_fixing)
```




```{r}
#| eval: true

mae_coldata |> 
  dplyr::count(in_CRF, randomized, in_assays, sample_type, assays) |> 
  gt()

```



### Imputing `study_day` for daily `visit_code` that did not have CRF data

There were a few entries with daily `visit_code` that did not have a date in the CRF data. Consequently they also did not have study days (`study_day`) assigned.

```{r}

mae_coldata |> 
  filter(
    is.na(study_day), 
    is.na(sample_type) | (sample_type == "Clinical sample")
    ) |> 
  dplyr::count(visit_type, assays, in_CRF, visit_attended, sample_type)

```



We impute the missing values using the `study_days` from "surrounding" visits.

```{r}

# first, we create a "dictionary" with the most common study_days for each daily swab visit code
study_days_for_daily_codes <- 
  mae_coldata |> 
  dplyr::filter(visit_type == "Home") |> 
  dplyr::count(visit_code, study_day) |> 
  arrange(visit_code, -n) |> 
  group_by(visit_code) |> 
  slice_head(n = 1) |> 
  ungroup() |> 
  dplyr::rename(exp_study_day = study_day) |> 
  select(-n)

```

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| eval: false

study_days_for_daily_codes |> 
  ggplot(aes(x = exp_study_day, y = visit_code)) +
  geom_point()

```

```{r}

# In the plot above, we notice that visits 1501:1503 did not have "expected" study day, so we create it by expanding the time series
study_days_for_daily_codes <-
  study_days_for_daily_codes |> 
  mutate(
    exp_study_day = 
      case_when(
        is.na(exp_study_day) & (visit_code %in% c(1501:1503)) ~ 
          as.numeric(visit_code) - 1501 + 36,
        TRUE ~ exp_study_day
      )
  )

```




```{r}

# visits from the participants with missing study_days
pid_with_missing_daily_study_days <- 
  # we first filter to retrieve the pids
  mae_coldata |> 
  filter(is.na(study_day)) |>  # , in_assays
  select(pid) |> distinct() |> 
  # we join with the coldata
  left_join(mae_coldata, by = join_by(pid)) |> 
  arrange(pid, visit_code) |> 
  # we join the study_day dictionary
  left_join(study_days_for_daily_codes) |> 
  mutate(diff = study_day - exp_study_day) |> # the typical difference for a participant between their study_day and the "expected" study_day (some participant started sampling at home one day early or late)
  group_by(pid) |> 
  fill(diff, .direction = "downup") |>  # impute when missing
  mutate(diff = ifelse(all(is.na(diff)), 0, diff)) |> # if still missing,we set to 0
  ungroup() |> 
  # we fix the study_day
  mutate(
    fixed_study_day = 
      case_when(
        is.na(study_day) & !is.na(exp_study_day) ~ exp_study_day + diff,
        TRUE ~ study_day
      )
  ) |> 
  select(uid, pid, visit_code, study_day, fixed_study_day, exp_study_day, diff, in_CRF, in_assays, crf_plates, assays)

# pid_with_missing_daily_study_days |> View()

```

```{r}

mae_coldata <- 
  mae_coldata |> 
  left_join(
    pid_with_missing_daily_study_days |> select(uid, fixed_study_day), 
    by = join_by(uid)
    ) |> 
  mutate(
    study_day =
      case_when(
        is.na(study_day) ~ fixed_study_day,
        TRUE ~ study_day
      )
  ) |> 
  select(-fixed_study_day)

```

```{r}
#| eval: false

mae_coldata |> 
  filter(is.na(study_day), in_assays) |> 
  View()

# all is fixed

```





### Creating the `visit`, `visit_number`, and `study_week` columns


```{r}

visit_dict <- 
  bind_rows(
    tibble(
      visit = "Screening", 
      definition = "Screening visit for which we have the most assay data. If two screening visits have the same number of assays, we take the last one.", 
      possible_visit_codes = "0000, 0001, 0010, 0011",
      visit_number = "V0",
      study_week = -1
    ),
    tibble(
      visit = "Add. screening", 
      definition = 'If a participant has data for two screening visits, one is the "main" screening visit, and the other is labelled as an additional screening visit.', 
      possible_visit_codes = "0000, 0001, 0010, 0011",
      visit_number = "V0b",
      study_week = -2
      ),
    tibble(
      visit = "Day 0 (pre-MTZ)", 
      definition = "Visit before the metronidazole treatment (MTZ) starts.", 
      possible_visit_codes = "1000, 1011",
      visit_number = "V1",
      study_week = 0
      ),
    tibble(
      visit = "Week 1 (post-MTZ)", 
      definition = "Visit after the metronidazole treatment (MTZ) ends.", 
      possible_visit_codes = "1100",
      visit_number = "V2",
      study_week = 1
      ),
    tibble(
      visit = "Week 2 (post-SP)", 
      definition = "Visit after the study product (SP) ends.", 
      possible_visit_codes = "1200, 1211",
      visit_number = "V3",
      study_week = 2
      ),
    tibble(
      visit = "Week 3", 
      definition = "Visit one week after the study product (SP) ends.", 
      possible_visit_codes = "1300",
      visit_number = "V4",
      study_week = 3
      ),
    tibble(
      visit = "Week 4", 
      definition = "Visit two weeks after the study product (SP) ends.", 
      possible_visit_codes = "1400",
      visit_number = "V5",
      study_week = 4
      ),
    tibble(
      visit = "Week 5", 
      definition = "Visit three weeks after the study product (SP) ends.", 
      possible_visit_codes = "1500, 1511, 1512",
      visit_number = "V6",
      study_week = 5
      ),
    tibble(
      visit = "Week 7", 
      definition = "Visit five weeks after the study product (SP) ends.", 
      possible_visit_codes = "1700, 1711",
      visit_number = "V7",
      study_week = 7
      ),
    tibble(
      visit = "Week 9", 
      definition = "Visit seven weeks after the study product (SP) ends.", 
      possible_visit_codes = "1900, 1911",
      visit_number = "V8",
      study_week = 9
      ),
    tibble(
      visit = "Week 12", 
      definition = "Visit ten weeks after the study product (SP) ends.", 
      possible_visit_codes = "2120, 2121, 2122",
      visit_number = "V9",
      study_week = 12
    )
  ) |> 
  mutate(
    visit = visit |> fct_reorder(study_week)
  )

```

```{r}

visit_dict |> 
  gt() |> 
  tab_header(
    title = "Visit codes and definitions"
  ) |> 
  cols_label(
    visit = "Visit",
    definition = "Definition",
    possible_visit_codes = "Possible visit codes",
    visit_number = "Visit number",
    study_week = "Study week"
  ) |> 
  cols_align(
    align = "left", columns = c(visit, definition, possible_visit_codes)
  )

```

We first determine, for each participant, which is their main *vs.* their additional screening visit.

```{r}

mae_coldata <- 
  mae_coldata |> 
  group_by(pid) |> 
  mutate(
    is_screening_visit = 
      case_when(
        visit_code %in% c("0000", "0001", "0010", "0011") ~ TRUE,
        TRUE ~ FALSE
      ),
    n_assays = ifelse(in_assays, str_split(assays, ", ") |> lengths(), 0),
    screening_visit_score = is_screening_visit * (n_assays + study_day/10)
  ) |> 
  arrange(-screening_visit_score) |> 
  mutate(
    screening_visit = 
      case_when(
        is_screening_visit & (row_number() == 1) ~ "main screening",
        is_screening_visit & (row_number() > 1) ~ "additional screening",
        TRUE ~ NA_character_
      )
  ) |> 
  ungroup() |> 
  select(-is_screening_visit, -n_assays, -screening_visit_score)

```

```{r}
#| eval: false

mae_coldata |> 
  dplyr::filter(screening_visit == "additional screening") |> 
  select(pid) |> distinct() |> 
  left_join(mae_coldata) |> 
  arrange(randomized |> desc(), pid, visit_code) |> 
  select(randomized, pid, visit_code, study_day, crf_plates, assays, screening_visit) |> 
  View()

mae_coldata |> 
  dplyr::filter(screening_visit == "additional screening") |> 
  select(pid) |> distinct() |> 
  left_join(mae_coldata) |> 
  arrange(randomized |> desc(), pid, visit_code) |> 
  select(randomized, pid, visit_code, study_day, crf_plates, assays, screening_visit) |> 
  dplyr::filter(randomized, (screening_visit == "main screening"), study_day == 0) |> 
  select(pid) |> distinct() |> 
  left_join(mae_coldata) |> 
  arrange(randomized |> desc(), pid, visit_code) |> 
  select(randomized, mITT, pid, visit_code, study_day, crf_plates, assays, screening_visit) |> 
  View()


# > check 068200020
# unclear - her main screening visit could be at day -3, but it might be at day 0, which would mean that her screening visit and pre-MTZ visit samples were collected on the same day
crf_raw$crf35 |> dplyr::filter(pid == "068200020") |> View()

```

```{r}

mae_coldata |> 
  filter(!is.na(screening_visit)) |> 
  dplyr::count(screening_visit) 

```

We next join the `mae_coldata` table with the `visit_dict` table to add the `visit`, `visit_number`, and `study_week` columns.

```{r}

mae_coldata <- 
  mae_coldata |> 
  select(-any_of(c("PIPV", "visit", "visit_number", "study_week"))) |>
  left_join(
    visit_dict |> 
      mutate(visit_code = possible_visit_codes |> str_split(", ")) |>
      unnest(visit_code) |>
      mutate(
        screening_visit = 
          case_when(
            visit_number == "V0b" ~ "additional screening",
            visit_number == "V0" ~ "main screening",
            TRUE ~ NA_character_
            ),
        PIPV = visit_number != "V0b"
        ) |> 
      select(visit_code, screening_visit, PIPV, visit, visit_number, study_week)
  ) |> 
  mutate(PIPV = PIPV |> replace_na(FALSE)) |> 
  relocate(
    PIPV, visit, visit_number, study_week, .after = "visit_code"
  ) 

```


We check that `study_day`s are unique for each `visit_number` at which assay samples were collected for each participant.

```{r}

mae_coldata |> 
  group_by(pid, visit_number) |> 
  mutate(n_study_day = n_distinct(study_day[!is.na(assays)])) |>
  ungroup() |> 
  filter(PIPV, n_study_day > 1) |> 
  select(uid, pid, visit_code, study_day, assays) |> 
  group_by(pid) |> 
  gt(row_group_as_column = TRUE)

```

There are just two participants for which the dates don't match and the different assays were collected at different dates.
This may be important for integrative analyses later, so we leave as is for now.

### Creating the dictionary for the mae_coldata table

We create a dictionary to document the variables in the `mae_coldata` table, which will be useful for downstream analyses and for the data sharing.

We already have a dictionnary for the `participants` table (`participants_variable_dictionary`), we use this and add the variables from the `visits` table and the `mae_coldata_assays` table, as well as the variables we created in the previous steps (e.g., `visit`, `visit_number`, `study_week`, etc.).

```{r}

coldata_dictionary <- 
  bind_rows(
    participants_variable_dictionary |> select(variable, description),
    tibble(
      variable = setdiff(colnames(mae_coldata), colnames(participants)),
      description = case_when(
        variable == "uid" ~ "Unique identifier for each participant x visit combination, in the format `pid`_`visit_code`",
        variable == "PIPV" ~ "Planned in-person visit (PIPV) flag",
        variable == "visit_code" ~ "Visit code as in the CRF data",
        variable == "visit" ~ "Visit name (e.g., Screening, Week 1 (post-MTZ), etc.)",
        variable == "visit_number" ~ "Visit number (e.g., V0, V1, etc.)",
        variable == "study_week" ~ "Study week number (from -2 to 12, with 0 being the day of the pre-MTZ visit)",
        variable == "study_day" ~ "Relative study day number with 0 being the day of the pre-MTZ visit",
        variable == "n_distinct_study_days" ~ "Number of distinct study days for a given participant and visit number",
        variable == "crf_plates" ~ "CRF plates for which we have data for this participant x visit combination",
        variable == "visit_attended" ~ "Whether the participant attended the visit according to the CRF data",
        variable == "visit_planned" ~ "Whether the visit was planned according to the CRF data",
        variable == "visit_type" ~ "Visit type (e.g., Clinic, Home, Unclear or NA)",
        variable == "in_CRF" ~ "Whether this participant x visit combination is in the CRF data", 
        variable == "in_assays" ~ "Whether this participant x visit combination is in an assay data",
        variable == "sample_type" ~ "Sample type for this participant x visit combination in the assay data (Clinical sample, Clinical sample (other study), Negative/Positive control), Test sample or NA)",
        variable == "control_type" ~ "Control type for the Negative and Positive control samples in the assay data (Nuclease-free water, Unused swab + C2, Mock 1 and Mock 2)",
        variable == "assays" ~ "Names of the assays for which we have data for this participant x visit combination (mg, qPCR, amplicon, luminex, flow, etc.)",
        variable == "screening_visit" ~ "Whether the visit is the main screening visit, an additional screening visit, or not a screening visit"
      )
  )
  )

```

```{r}
coldata_dictionary |> 
  arrange(variable) |>
  gt() |> 
  tab_header(
    title = "Dictionary for the mae_coldata table"
  ) |> 
  cols_label(
    variable = "Variable",
    description = "Description"
  ) |> 
  cols_align(
    align = "left", columns = variable
  )
```


## Creating an assay to document specimen collection, and available data


- `specimen_collection_swab` (`collected`, `not collected`, `unclear`)
- `specimen_collection_softcup` (`collected`, `not collected`, `unclear`)
- `specimen_collection_cytobrush` (`collected`, `not collected`, `unclear`)
- `mg` 
- `qPCR`
- `amplicon`
- `luminex`
- `flow`

```{r}

sample_and_visit_info_assay <- 
  mae_coldata |> 
  select(uid, sample_type, in_CRF, in_assays, randomized) 

sample_and_visit_info_assay <- 
  sample_and_visit_info_assay |> 
  dplyr::left_join(
    visits |> select(uid, starts_with("specimen_collection")),
    by = join_by(uid)
  ) 

sample_and_visit_info_assay <-
  sample_and_visit_info_assay |>
  expand_grid(assay = assay_list$mae_assay_name) |>
  dplyr::left_join(
    available_samples |>
      select(uid, assay) |>
      mutate(value = TRUE),
    by = join_by(uid, assay)
  ) |>
  mutate(value = value |> replace_na(FALSE)) |>
  pivot_wider(names_from = assay, values_from = value)

# sample_and_visit_info_assay <- 
#   sample_and_visit_info_assay |> 
#   mutate(
#     mg_category = 
#       case_when(
#         !str_detect(sample_type, "Clinical sample") & mg ~ "Not a clinical sample",
#         is.na(randomized) & mg ~ "???",
#         !is.na(randomized) & !randomized & mg ~ "Additional sample from non-randomized participants",
#         !is.na(specimen_collection_swab) & (specimen_collection_swab > 0) & mg ~ "Expected data",
#         (is.na(specimen_collection_swab) | (specimen_collection_swab == 0)) & mg ~ "Unexpected data",
#         !is.na(specimen_collection_swab) & (specimen_collection_swab > 0) & !mg ~ "Missing data",
#         (is.na(specimen_collection_swab) | (specimen_collection_swab == 0)) & !mg ~ "Sample not collected",
#         TRUE ~ "ERROR"
#       ),
#     qPCR_category = 
#       case_when(
#         !str_detect(sample_type, "Clinical sample") & qPCR ~ "Not a clinical sample",
#         is.na(randomized) & qPCR ~ "???",
#         !is.na(randomized) & !randomized & qPCR ~ "Additional sample from non-randomized participants",
#         !is.na(specimen_collection_swab) & (specimen_collection_swab > 0) & qPCR ~ "Expected data",
#         (is.na(specimen_collection_swab) | (specimen_collection_swab == 0)) & qPCR ~ "Unexpected data",
#         !is.na(specimen_collection_swab) & (specimen_collection_swab > 0) & !qPCR ~ "Missing data",
#         (is.na(specimen_collection_swab) | (specimen_collection_swab == 0)) & !qPCR ~ "Sample not collected",
#         TRUE ~ "ERROR"
#       ),
#     amplicon_category = 
#       case_when(
#         !str_detect(sample_type, "Clinical sample") & amplicon ~ "Not a clinical sample",
#         is.na(randomized) & amplicon ~ "???",
#         !is.na(randomized) & !randomized & amplicon ~ "Additional sample from non-randomized participants",
#         !is.na(specimen_collection_swab) & (specimen_collection_swab > 0) & amplicon ~ "Expected data",
#         (is.na(specimen_collection_swab) | (specimen_collection_swab == 0)) & amplicon ~ "Unexpected data",
#         !is.na(specimen_collection_swab) & (specimen_collection_swab > 0) & !amplicon ~ "Missing data",
#         (is.na(specimen_collection_swab) | (specimen_collection_swab == 0)) & !amplicon ~ "Sample not collected",
#         TRUE ~ "ERROR"
#       ),
#     luminex_category = 
#       case_when(
#         !str_detect(sample_type, "Clinical sample") & luminex ~ "Not a clinical sample",
#         is.na(randomized) & luminex ~ "???",
#         !is.na(randomized) & !randomized & luminex ~ "Additional sample from non-randomized participants",
#         !is.na(specimen_collection_softcup) & (specimen_collection_softcup != "not collected") & luminex ~ "Expected data",
#         (is.na(specimen_collection_softcup) | (specimen_collection_softcup == "not collected")) & luminex ~ "Unexpected data",
#         !is.na(specimen_collection_softcup) & (specimen_collection_softcup == "collected") & !luminex ~ "Missing data",
#         (is.na(specimen_collection_softcup) | (specimen_collection_softcup == "not collected")) & !luminex ~ "Sample not collected",
#         TRUE ~ "ERROR"
#       )
#   )

# 
# The last 5 columns can take the following values:
# 
# - `expected` (swab collected and available data), 
# - `unexpected` (swab not collected but available data), 
# - `missing` (swab collected but no available data),
# - `maybe missing` (unclear if swab was collected and no available data),
# - `not collected` (swab not collected), 
# - `additional` (swab collected from a non-randomized participant), 
# - `NA` (not collected, no available data)
# 


```

```{r}

se_sample_and_visit_info_assay <- 
  SummarizedExperiment(
    assays = list(
      sample_and_visit_info = 
        sample_and_visit_info_assay |> 
        select(-c(sample_type, in_CRF, in_assays, randomized)) |> 
        as.data.frame() |> 
        column_to_rownames("uid") |> 
        t()
    ),
    colData = 
      sample_and_visit_info_assay |> select(uid) |> 
      mutate(rownames = uid) |> as.data.frame() |> 
      column_to_rownames("rownames")
  )
```

## Creating a sample manifest linking `uid` to sample barcodes

In this section, we create a `sample_manifest` table which links each `uid` (unique identifier of the VIBRANT study, concatenation of study number, study site number, participant screening number, and visit code) to the corresponding biological sample `barcode` (the unique sample ID) for each assay. 

This will be useful to catch any mislabeling of sample and to do cross-assay QC as some assays (metagenomics and 16S amplicon sequencing and qPCR) were performed on the same sample, so they should have the same barcode.

We will also add to this manifest the DNA extraction related data (*e.g.*, extraction library plate nb) as they are shared across sequencing and qPCR assays.


```{r}

sample_manifest <- 
  bind_rows(
    se_mg |> colData() |> as.data.frame() |> as_tibble() |> 
      mutate(ext_lib_plate_nb = ext_lib_plate_nb |> as.character()) |> 
      select(
        uid, barcode = swab_barcode, 
             ext_lib_plate_nb, ext_lib_plate_id, ext_lib_position
        ) |> 
      mutate(source = "mg", sample_type = "swab"),
    se_qPRC |> colData() |> as.data.frame() |> as_tibble() |> 
      select(uid, barcode = vibr_sample_id, ext_lib_plate_nb) |> 
      mutate(source = "qPCR", sample_type = "swab"),
    se_daily_qPRC |> colData() |> as.data.frame() |> as_tibble() |> 
      select(
        uid, barcode = vibr_sample_id, 
             ext_lib_plate_nb, pcr_plate_id, pcr_plate_nb, ext_lib_plate_nb
        ) |>
      mutate(source = "qPCR_daily", sample_type = "swab"),
    se_16S_ASV |> colData() |> as.data.frame() |> as_tibble() |> 
      mutate(ext_lib_plate_nb = ext_lib_plate_nb |> as.character())  |> 
      select(
        uid, barcode = sample_id_barcode, 
        ext_lib_plate_type, ext_lib_plate_nb
        ) |>
      mutate(source = "16S_ASV", sample_type = "swab")
    # TODO: in the Luminex data processing, add the barcode information to the SE,
    # Same with the flow data (less critical)
  )

```


```{r}
sample_manifest 
```


### Consistency check: entries without barcodes

Some entries have no barcode:

```{r}

sample_manifest |> 
  dplyr::count(is.na(barcode), name = "n entries") |> 
  gt()

```

This is not unexpected as our data also contain data from positive/negative controls.


We check that all clinical samples have a barcode:

```{r}

sample_manifest <- 
  sample_manifest |> 
  select(-any_of("sample_type")) |> 
  left_join(
    mae_coldata |> select(uid, sample_type),
    by = join_by(uid)
  )

```

```{r}

sample_manifest |> 
  dplyr::count(is.na(barcode), sample_type, name = "n entries") |> 
  gt()

```

There is one clinical sample without barcode, which is problematic:

```{r}

sample_manifest |> 
  filter(sample_type == "Clinical sample", is.na(barcode)) |> 
  select(uid, barcode, source)


```

We check the barcodes for the "surrounding" `uid`s, which are from the same participant and previous/next visits:

```{r}

sample_manifest |> 
  filter(str_detect(uid, "068200278_140[1-5]")) |> 
  arrange(uid, source) |> 
  select(uid, barcode, source)

```


The sample with a missing barcode is a daily swab for which we do not have qPCR data.

Following the barcode order, the expected barcode would be `"2293841"`.

There are however no entries with that barcode.

```{r}


sample_manifest |> 
  filter(barcode == "2293841") |> 
  select(uid, barcode, source)


```

But that `uid` has data in the 16S assay, with compositions that appear to be consistent with that of surrounding visits:

```{r}


se_16S[, "068200278_1404"] |> assay("rel_ab") |> head()


```


```{r}

uids_to_check <- 
  sample_manifest |> 
  filter(str_detect(uid, "068200278_140[1-7]")) |> 
  pull(uid) |> sort() |> unique()


```

```{r}
#| fig-width: 8
#| fig-height: 2.5

se_16S[, uids_to_check] |>
  as_tibble() |> 
  filter(counts > 0, rel_ab > 1/100) |> 
  ggplot() +
  aes(y = uid |> fct_rev(), x = rel_ab, fill = .feature) +
  geom_col() +
  ylab("") +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.key.size = unit(0.2, "cm")
  )

```

:::callout-important
Check with the 16S amplicon data generator what the barcode for that sample is.
:::

### Consistency check: barcodes with different assignments across assays

```{r}

sample_manifest_summary_by_barcode <- 
  sample_manifest |> 
  filter(!is.na(barcode)) |>
  group_by(barcode) |> 
  summarize(
    n = n(), 
    n_uids = n_distinct(uid),
    uids = str_c(unique(uid), collapse = ", "),
    sources = str_c(unique(source), collapse = ", "),
    sample_types = str_c(unique(sample_type), collapse = ", "),
    .groups = "drop"
    ) 

sample_manifest_summary_by_barcode |> 
  arrange(-n_uids) |> 
  slice_head(n = 10)

```
There are 2 barcodes that have been assigned to different `uid`s, which is unexpected and may indicate a mislabeling of samples.


#### 2286665

```{r}

sample_manifest |> 
  filter(str_detect(uid, "068200385_120[2-7]")) |> 
  arrange(barcode, source) |> 
  select(barcode, uid, source)

```

```{r}

se_16S[, c("068200385_1204","068200385_1205", "068200385_1206", "068200385_1207")] |>
  assay("rel_ab") |> head() |> round(4)

```
The 16S data looks legit and distinct for these two samples... 

:::callout-important
Check with the 16S amplicon data generator what the barcode for that sample is.
:::

Until we have figured this out, we set the barcode of sample `068200385_1206` to NA as it is the one that looks more likely to be mislabelled.

```{r}
sample_manifest <- 
  sample_manifest |> 
  mutate(
    barcode = case_when(
      uid == "068200385_1206" ~ NA_character_,
      TRUE ~ barcode
    )
  )
```



#### 2292179


```{r}

sample_manifest |> 
  filter(barcode == "2292179") |>  
  arrange(barcode, source) |> 
  select(barcode, uid, source)

```
Again it appears to be something unexpected in the 16S amplicon data...


```{r}

sample_manifest |> 
  filter(str_detect(uid, "068200356_130[4-6]|068200439_110[5-7]")) |>  
  arrange(barcode, source) |> 
  select(barcode, uid, source)

```

From this data, it looks like the barcode "2292179" should be assigned to "068200439_1107" and not to "068200356_1305".

The barcode of "068200356_1305" should likely be "2292083" or "2292085" to follow the logical sequence in barcode numbers, but these two barcodes do not exist:

```{r}

sample_manifest |> 
  filter(barcode %in% c("2292083", "2292085")) |>  
  arrange(barcode, source) |> 
  select(barcode, uid, source)

```

We check the 16S data just to be sure that, besides a barcode mislabeling, there is no other issue with the data for these two samples:

```{r}

uids_to_check <- 
  sample_manifest |> 
  filter(str_detect(uid, "068200356_130[4-6]|068200439_110[5-7]")) |> 
  pull(uid) |> sort() |> unique()

se_16S[, uids_to_check] |>
  assay("rel_ab") |> head() |> round(4)

```

```{r}
#| fig-width: 8
#| fig-height: 2.5

se_16S[, uids_to_check] |>
  as_tibble() |> 
  filter(counts > 0, rel_ab > 1/100) |> 
  ggplot() +
  facet_grid(sample_id_barcode ~ ., scales = "free", space = "free") +
  aes(y = uid, x = rel_ab, fill = .feature) +
  geom_col() +
  ylab("") +
  theme(
    strip.text.y = element_text(angle = 0),
    legend.key.size = unit(0.2, "cm")
  )

```

Fortunately, the `uid` assignments look correct and we can likely consider that the barcode is missing for the `068200356_1305` sample.

:::callout-important
Check with the 16S amplicon data generator what the barcode for that sample is.
:::

Until we have figured this out, we set the barcode of sample `068200356_1305` to NA as it is the one that looks more likely to be mislabelled.

```{r}

sample_manifest <- 
  sample_manifest |> 
  mutate(
    barcode = case_when(
      uid == "068200356_1305" ~ NA_character_,
      TRUE ~ barcode
    )
  )
```





### Consistency check: extraction data


We add the `visit_type` from the `mae_coldata` table to the `sample_manifest` table to check if extraction library plate numbers are unique across all sample types or if they are numbered by sample type (weekly vs daily swabs).

```{r}

sample_manifest <- 
  sample_manifest |> 
  select(-any_of("visit_type")) |>
  left_join(
    mae_coldata |> select(uid, visit_type),
    by = join_by(uid)
  )

```



```{r}
#| fig-width: 6
#| fig-height: 10

sample_manifest |> 
  # filter(sample_type == "Clinical sample") |> 
  dplyr::count(source, sample_type, visit_type, ext_lib_plate_nb) |>
  distinct() |> 
  ggplot() +
  facet_grid(visit_type ~ source, scales = "free", space = "free") +
  aes(y = ext_lib_plate_nb |> fct_rev(), x = sample_type, fill = n |> log10()) +
  geom_tile() +
  ylab("Extraction library plate number\n(ext_lib_plate_nb)") +
  scale_fill_gradient("Number of samples\n(log10 scale)", low = "lightblue1", high = "steelblue4") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

First, it looks like numbers are not unique across sample types, so we have to specify the extraction batch (weekly or daily swabs) to differentiate between numbers. 

```{r}

sample_manifest <- 
  sample_manifest |>
  mutate(
    ext_lib_plate_batch = 
      case_when(
        source == "mg" ~ "Weekly swab",
        source == "qPCR" ~ "Weekly swab",
        source == "qPCR_daily" ~ "Daily swab",
        source == "16S_ASV" ~ "???"
      )
  )
```


In the 16S amplicon data, the extraction library plate numbers are only known for the control samples, but not for the clinical samples.

For the clinic samples (weekly swab), the extraction plate numbers are provided in both the `"mg"` and `"qPCR"` assays. We'll need to check that they are consistent between the two assays.

For the home samples (daily swabs), the extraction plate numbers are only provided in the `"qPCR_daily"` assay, so we won't be able to check their consistency across assays.

```{r}

sample_manifest_summary_by_uid <- 
  sample_manifest |> 
  filter(!is.na(ext_lib_plate_nb), sample_type == "Clinical sample") |> 
  group_by(uid) |> 
  summarize(
    n = n(), 
    n_ext_batch = n_distinct(ext_lib_plate_batch),
    ext_batches = str_c(unique(ext_lib_plate_batch) |> str_replace_na("NA"), collapse = " | "),
    n_plate_nb = n_distinct(ext_lib_plate_nb),
    plate_nbs = str_c(unique(ext_lib_plate_nb) |> str_replace_na("NA"), collapse = " | "),
    n_sources = n_distinct(source),
    sources = str_c(unique(source), collapse = " | "),
    .groups = "drop"
    ) 

```
We first check if any samples were extracted in the two different batches:

```{r}

sample_manifest_summary_by_uid |> 
  filter(n_ext_batch > 1)

sample_manifest |> 
  filter(uid %in% c("068200127_1100", "068200414_1500")) |> 
  select(uid, barcode, ext_lib_plate_nb, source, pcr_plate_nb, visit_type, ext_lib_plate_batch) |> 
  arrange(uid, barcode, source)

```

There are two "weekly" samples that were extracted in both batches. For both of these samples, the extraction plate number in the `qPCR` assay is 12, which I think it the plate with the re-extracted samples.

It is unclear if the amplicon sequencing was done on the original extraction or on the re-extracted samples.

:::callout-important
Ask Joseph about extraction plates for these samples...

& check whether these samples are among the duplicated samples in the original 16S data.
:::

For now, we will assume that the sequencing was done on the re-extracted samples...


Next, we check if there are any samples with more than one extraction plate number.


```{r}

sample_manifest_summary_by_uid |> 
  filter(n_ext_batch <= 1) |> 
  dplyr::count(n_plate_nb, sources)

sample_manifest_summary_by_uid |> 
  filter(n_ext_batch <= 1) |> 
  arrange(-n_plate_nb) |> 
  head()

```

There are a few samples with two different extraction plate numbers, but all of them are mismatch between the `"mg"` and `"qPCR"` assays, which is not unexpected because in the metagenomics data, Michael only provided the data for the retained extraction, while in the qPCR rate, the data from both extraction were provided. 


```{r}

sample_manifest |> 
  filter(uid %in% c("068100006_1400", "068100013_1100")) |> 
  select(uid, barcode, ext_lib_plate_nb, source, pcr_plate_nb, visit_type, ext_lib_plate_batch) |> 
  arrange(uid, barcode, source)

```


#### Imputing extraction plate numbers for amplicon data (and a few missing values in the metagenomics assay)

Since we didn't have the extraction library plate number for the 16S data, we will use the data from the qPCR assays to fill in the information for the amplicon data.

```{r}

ext_plate_amplicon <- 
  sample_manifest |> 
  filter(sample_type == "Clinical sample") |> 
  mutate(
    min_plate_nb = ext_lib_plate_nb |> str_remove(",.*") |> parse_number(),
  ) |> 
  filter(!is.na(min_plate_nb)) |> 
  group_by(uid) |>
  arrange(uid, ext_lib_plate_batch) |> 
  mutate(
    n_ext_batches = n_distinct(ext_lib_plate_batch),
    n_plate_nb = n_distinct(min_plate_nb)
    ) |> 
  select(uid, barcode, source, ext_lib_plate_batch, min_plate_nb, n_ext_batches, n_plate_nb) |> 
  arrange(-n_ext_batches, -n_plate_nb, uid, min_plate_nb) |> 
  slice_head(n = 1) |> 
  ungroup() |> 
  mutate(ext_lib_plate_nb = min_plate_nb) |> 
  select(uid, ext_lib_plate_batch, ext_lib_plate_nb)
  
```




```{r}

sample_manifest <- 
  sample_manifest |> 
  select(
    uid, barcode, source, sample_type, visit_type, 
    ext_lib_plate_batch, ext_lib_plate_nb
    ) |> 
  select(-any_of(ends_with("_new"))) |> 
  left_join(
    ext_plate_amplicon |> 
      select(
        uid,
        ext_lib_plate_batch_new = ext_lib_plate_batch, 
        ext_lib_plate_nb_new = ext_lib_plate_nb
        ),
    by = join_by(uid)
  ) 

```


```{r}

sample_manifest <- 
  sample_manifest |> 
  mutate(
    ext_lib_plate_batch =
      ifelse(
        (source == "16S_ASV") & !is.na(ext_lib_plate_batch_new), 
        ext_lib_plate_batch_new, ext_lib_plate_batch
      ),
    ext_lib_plate_nb_origin = ext_lib_plate_nb,
    ext_lib_plate_nb = 
      ifelse(
        ((source == "16S_ASV") & !is.na(ext_lib_plate_nb_new)) | 
          ((source == "mg") & is.na(ext_lib_plate_nb)), 
        ext_lib_plate_nb_new, ext_lib_plate_nb
      ),
    ext_lib_plate_id = 
      ifelse(
        ext_lib_plate_batch != "???",
        str_c(
          ext_lib_plate_batch, "s plate ", 
          str_pad(ext_lib_plate_nb, width = 2, pad = "0")
        ),
        NA_character_
      ),
    ext_lib_plate_nb_num = ext_lib_plate_nb |> str_remove(",.*") |> parse_number(),
    ext_lib_info_source = 
      case_when(
        !is.na(ext_lib_plate_nb) & (ext_lib_plate_nb == ext_lib_plate_nb_origin) ~ "colData of source assay",
        !is.na(ext_lib_plate_nb) ~ "imputed from qPCR or mg",
        TRUE ~ NA_character_
      )
  ) |> 
  arrange(ext_lib_plate_batch, ext_lib_plate_nb_num) |> 
  mutate(ext_lib_plate_id = ext_lib_plate_id |> fct_inorder()) 
  
```


```{r}

sample_manifest <- 
  sample_manifest |> 
  select(
    uid, barcode, source, sample_type, visit_type, 
    ext_lib_plate_batch, ext_lib_plate_nb, ext_lib_plate_id, ext_lib_info_source
  )

```

```{r}
sample_manifest |> 
  dplyr::count(ext_lib_info_source)
```







After imputation, there are still a few samples with missing extraction data:

```{r}

sample_manifest |> filter(is.na(ext_lib_plate_id), sample_type == "Clinical sample") 

```
We impute the plates from these samples from the surrounding samples of the same participant when possible:

```{r} 


sample_manifest |> 
  filter(uid |> str_detect("068200356_130[3-7]")) |> 
  arrange(uid) |> 
  select(uid, barcode, ext_lib_plate_nb, ext_lib_plate_id) |> 
  distinct()
# 19

sample_manifest |> 
  filter(uid |> str_detect("068200385_120[3-7]")) |> 
  arrange(uid) |> 
  select(uid, barcode, ext_lib_plate_nb, ext_lib_plate_id) |> 
  distinct()
# 17


sample_manifest |> 
  filter(uid |> str_detect("068200278_140[2-7]")) |> 
  arrange(uid) |> 
  select(uid, barcode, ext_lib_plate_nb, ext_lib_plate_id) |> 
  distinct()

# 21 or 22

```

```{r}
#| fig-width: 10
#| fig-height: 3
#| eval: false

sample_manifest |> 
  filter(ext_lib_plate_batch == "Daily swab", sample_type == "Clinical sample") |>
  select(uid, barcode, ext_lib_plate_nb) |> 
  distinct() |> 
  dplyr::count(ext_lib_plate_nb) |> 
  ggplot() +
  aes(x = ext_lib_plate_nb, y = n) +
  geom_label(aes(label = n), size = 3) 

```



```{r}

sample_manifest <- 
  sample_manifest |> 
  mutate(
    ext_lib_plate_batch =
      case_when(
        uid %in% c("068200278_1404", "068200356_1305", "068200385_1205", "068200385_1206")  ~ "Daily swab",
        TRUE ~ ext_lib_plate_batch
      ),
    ext_lib_plate_nb =
      case_when(
        uid == "068200278_1404" ~ "21", # not sure...
        uid == "068200356_1305" ~ "19", # not sure....
        uid == "068200385_1205" ~ "17",
        uid == "068200385_1206" ~ "17",
        TRUE ~ ext_lib_plate_nb
      ),
    ext_lib_plate_id =
      case_when(
        uid == "068200278_1404" ~ "Daily swabs plate 21" |> factor(levels = levels(ext_lib_plate_id)),
        uid == "068200356_1305" ~ "Daily swabs plate 19" |> factor(levels = levels(ext_lib_plate_id)),
        uid == "068200385_1205" ~ "Daily swabs plate 17" |> factor(levels = levels(ext_lib_plate_id)),
        uid == "068200385_1206" ~ "Daily swabs plate 17" |> factor(levels = levels(ext_lib_plate_id)),
        TRUE ~ ext_lib_plate_id
      ),
    ext_lib_info_source = 
      case_when(
        uid %in% c("068200278_1404", "068200356_1305", "068200385_1205", "068200385_1206")  ~ "imputed from surrounding samples of the same participant",
        TRUE ~ ext_lib_info_source
      )
  )

```


```{r}

sample_manifest <- 
  sample_manifest |> 
  select(
    uid, barcode, source, 
    ext_lib_plate_batch, ext_lib_plate_nb, ext_lib_plate_id, ext_lib_info_source
  ) |> 
  arrange(source, ext_lib_plate_id, uid)

```



```{r}
sample_manifest |> 
  dplyr::count(ext_lib_info_source)
```


## Creating the `MultiAssayExperiment` object




```{r}

SE_list <- 
  list(
    visit_and_sample_info = se_sample_and_visit_info_assay,
    mg = se_mg,
    qPCR = se_qPRC,
    qPCR_daily = se_daily_qPRC,
    amplicon_ASV = se_16S_ASV,
    amplicon = se_16S,
    luminex = se_luminex,
    flow = se_flow
  )


mae <- 
  MultiAssayExperiment::MultiAssayExperiment(
    experiments = MultiAssayExperiment::ExperimentList(SE_list),
    colData = 
      mae_coldata |> select(-in_CRF, -in_assays) |> 
      as.data.frame() |> mutate(rownames = uid) |> 
      column_to_rownames("rownames"),
    metadata = list(
      study = "VIBRANT",
      data_source = "actual study data (not simulated)",
      data_status = "partially QCed",
      date = today(),
      colData_dictionary = coldata_dictionary,
      crf_plates_description = crf_plates_dictionary,
      crf_data_raw = crf_raw,
      crf_data_clean = crf_clean, 
      participant_crfs_merged = participant_crfs_merged,
      visits_crfs_merged = visits_crfs_merged,
      exposures = exposures,
      sample_manifest = sample_manifest |> select(uid, barcode, source_assay = source, starts_with("ext_lib_"))
    )
  )

```


```{r}
mae
```




# Available assays

```{r}

mae_sample_map <- 
  mae@sampleMap |> as_tibble() |> dplyr::filter(assay %in% assay_list$mae_assay_name) |>
  select(uid = primary, assay) |> 
  arrange(uid) |> 
  dplyr::left_join(assay_list |> dplyr::rename(assay = mae_assay_name), by = join_by(assay)) |> 
  dplyr::right_join(
    mae@colData |> as.data.frame() |> as_tibble() |> 
      select(uid, pid, visit_code, PIPV, visit, visit_type, sample_type, location, randomized, arm), 
    by = join_by(uid)
    )

```

All samples

```{r}
#| fig-height: 4
#| fig-width: 12

mae_sample_map |> 
  dplyr::filter(!is.na(assay), assay != "amplicon_ASV") |> 
  group_by(uid) |> mutate(n = n()) |> ungroup() |> 
  mutate(uid = uid |> fct_reorder(-n)) |>
  ggplot() +
  aes(x = uid, y = assay_label |> fct_rev(), fill = assay_label) +
  geom_tile() +
  facet_grid(. ~ sample_type, scales = "free_x", space = "free_x") +
  scale_x_discrete("Sample IDs", breaks = NULL) +
  ylab("") +
  scale_fill_brewer("Assay", type = "qual", palette = 7) + # 3
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.x = element_text(angle = 90),
    legend.position = "bottom"
    )

```


Longitudinal availability for all visits


```{r}
#| fig-height: 17
#| fig-width: 15

mae_sample_map |> 
  dplyr::filter(!is.na(assay), assay != "amplicon_ASV") |> 
  # filter(visit_code %in% c("0000", seq(1000, 1900, by = 100), "2120")) |> 
  ggplot() +
  aes(x = assay_label, y = pid, fill = assay_label) +
  geom_tile() +
  facet_grid(
    ifelse(randomized, "Randomized", "Not randomized") + location ~ visit_code, 
    scales = "free_y", space = "free_y"
    ) + # sample_category
  scale_y_discrete("Participants IDs") + #, breaks = "NULL")  +
  xlab("") +
  scale_fill_brewer("Assay", type = "qual", palette = 7) + # 3
  theme(
    axis.text.x = element_blank(), #element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0),
    strip.text.x = element_text(angle = 90),
    legend.position = "bottom"
    )

```
For the weekly visits of all participants

```{r}
#| fig-height: 12
#| fig-width: 12

mae_sample_map |> 
  dplyr::filter(!is.na(assay), assay != "amplicon_ASV") |> 
  dplyr::filter(visit_type == "Clinic") |> 
  ggplot() +
  aes(x = assay_label, y = pid, fill = assay_label) +
  geom_tile() +
  facet_grid(
    location + ifelse(randomized, "Randomized", "Not randomized")  + arm  ~ visit_code, 
    scales = "free_y", space = "free_y"
    ) + # sample_category
  scale_y_discrete("Participants IDs") + #, breaks = "NULL")  +
  xlab("") +
  scale_fill_brewer("Assay", type = "qual", palette = 7) + # 3
  theme(
    axis.text.x = element_blank(), #element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0),
    strip.text.x = element_text(angle = 90),
    legend.position = "bottom"
    )

```

For the weekly visits of randomized participants

```{r}
#| fig-height: 12
#| fig-width: 12

mae_sample_map |> 
  dplyr::filter(randomized) |> 
  dplyr::filter(!is.na(assay), assay != "amplicon_ASV") |> 
  dplyr::filter(visit_type == "Clinic") |> 
  ggplot() +
  aes(x = assay_label, y = pid, fill = assay_label) +
  geom_tile() +
  facet_grid(
    location + ifelse(randomized, "Randomized", "Not randomized")  + arm  ~ visit_code, 
    scales = "free_y", space = "free_y"
    ) + # sample_category
  scale_y_discrete("Participants IDs") + #, breaks = "NULL")  +
  xlab("") +
  scale_fill_brewer("Assay", type = "qual", palette = 7) + # 3
  theme(
    axis.text.x = element_blank(), #element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0),
    strip.text.x = element_text(angle = 90),
    legend.position = "bottom"
    )

```


```{r}
#| fig-height: 8
#| fig-width: 8

mae_sample_map |> 
  dplyr::filter(randomized) |> 
  dplyr::filter(!is.na(assay), assay != "amplicon_ASV") |> 
  dplyr::filter(PIPV) |> 
  ggplot() +
  aes(x = assay_label, y = pid, fill = assay_label) +
  geom_tile() +
  facet_grid(
    location + arm  ~ visit, 
    scales = "free_y", space = "free_y"
    ) + # sample_category
  scale_y_discrete("Participants IDs") + #, breaks = "NULL")  +
  xlab("") +
  scale_fill_brewer("Assay", type = "qual", palette = 7) + # 3
  theme(
    axis.text.x = element_blank(), #element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0),
    strip.text.x = element_text(angle = 90),
    legend.position = "bottom"
    )

```



# Saving the `MultiAssayExperiment` objects

We export the full `MultiAssayExperiment` object as an `.rds` file for further analyses.

```{r}

saveRDS(
  mae, 
  str_c(
    get_02_output_dir(),
    "mae_full_", today() |> str_remove_all("-"), ".rds"
  )
)

```



