---
title: "MAE augmentation (primary and secondary outcome definitions)"
author: Laura Symul, Laura Vermeren
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: refresh # true refresh false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| warning: false
#| cache: false

library(tidyverse)
library(magrittr)
library(gt)
library(patchwork)
library(SummarizedExperiment)
library(tidySummarizedExperiment)
library(MultiAssayExperiment)

tmp <- fs::dir_map("R scripts/", source)
tmp <- fs::dir_map("../VIBRANT-99-utils/R/", source)
rm(tmp)
theme_set(theme_light())
```


```{r}

mae_files <- 
  fs::dir_ls(
    get_02_output_dir(),
    regexp = "mae_full_.*\\.rds$"
  ) |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)

mae <- readRDS(mae_files)

rm(mae_files)

```


## Introduction





> TODO: change the title of document, and edit the description of what this document does

The primary purpose of this document is to define and create additional assays to the `MultiAssayExperiment` object that contain the primary and secondary outcomes of the VIBRANT trial.





## Comparison of the total 16S in weekly *vs* daily samples

```{r}

qpcr_16S <- 
  bind_rows(
    mae[["qPCR_daily"]] |> 
      as_tibble() |>
      dplyr::filter(.feature == "16S", !is.na(copies_per_swab_med)) |>
      select(uid, dilution, copies_per_swab_med) |> 
      mutate(assay = "daily qPCR") |> 
      expand_grid(calibration = c("orig", "LL3", "LM")),
    mae[["qPCR"]] |> 
      as_tibble() |>
      dplyr::filter(.feature == "16S") |>
      select(uid, dilution, copies_per_swab_med, copies_per_swab_orig_med, copies_per_swab_LM_med) |> 
      pivot_longer(
        cols = c(copies_per_swab_med, copies_per_swab_orig_med, copies_per_swab_LM_med),
        names_to = "calibration",
        names_pattern = "copies_per_swab_(.*)_med",
        values_to = "copies_per_swab_med"
      ) |>
      mutate(
        calibration = calibration |> str_replace_na("LL3"),
        assay = "weekly qPCR"
        )
  ) |> 
  dplyr::left_join(
    mae@colData |> 
      as_tibble() |> 
      dplyr::select(
        uid, pid, visit_code, visit, study_day, sample_type, 
        site, randomized, arm, ITT, mITT, PP),
    by = join_by(uid)
    ) |>
  dplyr::filter(sample_type == "Clinical sample") |> 
  mutate(calibration = calibration |> fct_inorder())

```


```{r}
#| fig-width: 13
#| fig-height: 8
  
qpcr_16S |>
  filter(randomized, study_day > -10) |> 
  ggplot() +
  aes(x = study_day, y = copies_per_swab_med, color = assay) +
  facet_grid(site ~ calibration, 
    scales = "free_y", space = "free_y", labeller = label_both
    ) +
  geom_line(aes(group = interaction(pid, assay)), alpha = 0.2) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_y_log10() +
  theme(
    strip.text.y = element_text(angle = 0)
  )
```


```{r}


selected_time_points <- 
  qpcr_16S |> 
  filter(randomized) |> 
  group_by(pid, study_day) |> 
  mutate(n_assays = n_distinct(assay)) |>
  ungroup() |> 
  dplyr::filter(n_assays == 2) 

tmp <- 
 dplyr::full_join(
   selected_time_points |> 
     filter(assay == "weekly qPCR") |> 
     dplyr::select(calibration, copies_per_swab_med, pid, study_day, site) |> 
     dplyr::rename(copies_per_swab_med_weekly = copies_per_swab_med),
   selected_time_points |> 
     filter(assay == "daily qPCR")  |> 
     dplyr::select(calibration, copies_per_swab_med, pid, study_day, site) |> 
     distinct() |> 
     dplyr::rename(copies_per_swab_med_daily = copies_per_swab_med)
 )
  

```


```{r}
#| fig-width: 8
#| fig-height: 6

tmp |> 
  ggplot() +
  aes(x = copies_per_swab_med_daily, y = copies_per_swab_med_weekly, col = calibration) +
  facet_grid(
    site ~ calibration, labeller = label_both
    ) +
  geom_abline(slope = 1, intercept = 0, color = "black") +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_x_log10("16S copies/swabs - Daily swabs") +
  scale_y_log10("16S copies/swabs - Weekly swabs") +
  coord_fixed() +
  ggtitle("Swabs collected the same day from the same participant")

```


## Comparison of LBP strains in weekly *vs* daily samples

```{r}

qpcr_LBP <- 
  bind_rows(
    mae[["qPCR_daily"]] |> 
      as_tibble() |>
      dplyr::filter(.feature != "16S", !is.na(copies_per_swab_med)) |>
      select(uid, dilution, target, copies_per_swab_med) |> 
      mutate(assay = "daily qPCR"),
    mae[["qPCR"]] |> 
      as_tibble() |>
      dplyr::filter(.feature != "16S") |>
      select(uid, dilution, target, copies_per_swab_med) |>
      mutate(assay = "weekly qPCR")
  ) |> 
  dplyr::left_join(
    mae@colData |> 
      as_tibble() |> 
      dplyr::select(
        uid, pid, visit_code, visit, study_day, sample_type, 
        site, randomized, arm, ITT, mITT, PP),
    by = join_by(uid)
    ) |>
  dplyr::filter(sample_type == "Clinical sample")

```


```{r}
#| fig-width: 18
#| fig-height: 8
  
qpcr_LBP |>
  filter(randomized, study_day > -10) |> 
  ggplot() +
  aes(x = study_day, y = copies_per_swab_med, color = assay) +
  facet_grid(site ~ target, 
    scales = "free_y", space = "free_y", labeller = label_both
    ) +
  geom_line(aes(group = interaction(pid, assay)), alpha = 0.2) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_y_log10() +
  theme(
    strip.text.y = element_text(angle = 0)
  )
```

```{r}
#| fig-width: 18
#| fig-height: 20
  
qpcr_LBP |>
  filter(mITT, study_day > -10, site == "MGH") |> 
  ggplot() +
  aes(x = study_day, y = copies_per_swab_med, color = assay) +
  facet_grid(site + pid ~ target, labeller = label_both) +
  geom_line(alpha = 0.2) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_y_log10() +
  theme(
    strip.text.y = element_text(angle = 0)
  )
```

```{r}
#| eval: false
#| fig-width: 18
#| fig-height: 20
  
qpcr_LBP |>
  filter(mITT, study_day > -10, site == "CAP", arm == "LC-106-o") |> 
  ggplot() +
  aes(x = study_day, y = copies_per_swab_med, color = assay) +
  ggplot2::facet_grid(pid ~ target) + #    , labeller = label_both
  geom_line(alpha = 0.2) +
  geom_point(alpha = 0.5, size = 0.5) +
  scale_y_log10() +
  theme(
    strip.text.y = element_text(angle = 0)
  )

```


```{r}
#| fig-width: 12
#| fig-height: 8
  
qpcr_LBP |>
  filter(randomized, study_day > -10) |> 
  ggplot() +
  aes(x = assay, y = copies_per_swab_med, color = assay, fill = assay) +
  facet_grid(site ~ target, 
    scales = "free_y", space = "free_y", labeller = label_both
    ) +
  geom_boxplot(alpha = 0.5) +
  scale_y_log10() +
  theme(
    strip.text.y = element_text(angle = 0),
    axis.text.x = element_blank(),
    legend.position = "bottom"
  )
```


```{r}


selected_time_points <- 
  qpcr_LBP |> 
  filter(randomized) |> 
  group_by(pid, study_day) |> 
  mutate(n_assays = n_distinct(assay)) |>
  ungroup() |> 
  dplyr::filter(n_assays == 2) 

tmp <- 
 dplyr::full_join(
   selected_time_points |> 
     filter(assay == "weekly qPCR") |> 
     dplyr::select(target, copies_per_swab_med, pid, study_day, site) |> 
     dplyr::rename(copies_per_swab_med_weekly = copies_per_swab_med),
   selected_time_points |> 
     filter(assay == "daily qPCR")  |> 
     dplyr::select(target, copies_per_swab_med, pid, study_day, site) |> 
     distinct() |> 
     dplyr::rename(copies_per_swab_med_daily = copies_per_swab_med)
 )
  
```


```{r}
#| fig-width: 18
#| fig-height: 3

tmp |> 
  mutate(
    study_phase = case_when(
      study_day <= 0 ~ "pre-MTZ",
      study_day > 0 & study_day <=7 ~ "Week 1",
      study_day <= 15 ~ "Week 2",
      TRUE ~ "Week 3+"
    ) |> factor(levels = c("pre-MTZ", "Week 1", "Week 2", "Week 3+"))
  ) |> 
  ggplot() +
  aes(x = copies_per_swab_med_daily, y = copies_per_swab_med_weekly, col = study_phase) +
  facet_grid(
    site ~ target, labeller = label_both
    ) +
  geom_abline(slope = 1, intercept = 0, color = "black") +
  geom_point(alpha = 0.5, size = 0.5) + # aes(col = study_day), 
  scale_x_log10("Copies/swabs - Daily swabs") +
  scale_y_log10("Copies/swabs - Weekly swabs") +
  # scale_color_gradient(low = "dodgerblue1", high = "red") +
  coord_fixed()

```


```{r}
#| fig-width: 12
#| fig-height: 6

tmp |> 
  ggplot() +
  aes(x = copies_per_swab_med_daily, y = copies_per_swab_med_weekly, col = site) +
  ggh4x::facet_nested_wrap(target + site ~ ., ncol = 10) +
  geom_abline(slope = 1, intercept = 0, color = "black") +
  geom_point(alpha = 0.5, size = 0.5) + # aes(col = study_day), 
  scale_x_log10("Copies/swabs - Daily swabs") +
  scale_y_log10("Copies/swabs - Weekly swabs") +
  scale_color_manual(values = site_colors) +  
  coord_fixed() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1))

```

## Estimation and corrections for the technical biases between the two assays (16S qPCR)


> TODO: include code and explanation from M&K's documents


## Saving the `MultiAssayExperiment` objects

We save the MAE for downstream analyses.

```{r}

saveRDS(
  mae, 
  str_c(
    get_03_output_dir(),  
    "mae_full_", today() |> str_remove_all("-"), ".rds"
  )
)

```



