---
title: "MAE augmentation - Calculated variables"
author: Laura Symul, Laura Vermeren
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: refresh # true refresh false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| warning: false
#| cache: false

library(tidyverse)
library(magrittr)
library(gt)
library(patchwork)
library(grid)
library(gridExtra)
library(factoextra)
library(SummarizedExperiment)
library(tidySummarizedExperiment)
library(MultiAssayExperiment)

tmp <- fs::dir_map("R scripts/", source)
tmp <- fs::dir_map("../VIBRANT-99-utils/R/", source)
rm(tmp)
theme_set(theme_light())
```



```{r}

mae_files <- 
  fs::dir_ls(
    get_03_output_dir(),
    regexp = "mae_full_02_.*\\.rds$"
  ) |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)

mae <- readRDS(mae_files)

rm(mae_files)

```



# Calculated variables
The objective of this section is to create new tables in the metadata of the `MultiAssayExperiment` object that contain calculated variables relevant for downstream analyses. We create two new tables:

- `calc_vars_pid` contains one row per participant, initialized with `pid`, `site`, and population flags (`ITT`, `mITT`, `PP`, `randomized`). Variables related to participant-level characteristics and outcomes will be progressively added to this table.

- `calc_vars_visits` contains one row per participant and study day. It is initialized from the `colData` of the `MultiAssayExperiment` object, which provides the list of observed study days alongside randomization and population flag information. When a clinical and a home visit fall on the same study day, the two rows are merged into a single row with `visit_type = "Clinic & Home"`. To ensure continuity over the main follow-up period, all missing study days between day 0 and day 35 are imputed using `tidyr::complete()`. For these imputed days, the `visit_code` is derived from the study day using the weekly visit structure (e.g., day 7 → `1100`, day 8 → `1101`), and `visit_type` is inferred from the `visit_code` — visits with a code divisible by 100 are classified as `Clinic`, all others as `Home`. Study days beyond day 35 (follow-up visits at weeks 7, 9, 12, and end of study) are not imputed and remain as observed. A `predicted` flag distinguishes imputed rows from true observations.

All variables are derived from the `participant_crfs_merged` and `visits_crfs_merged` tables available in the metadata of the `MultiAssayExperiment` object, which contain the harmonized CRF data, as well as from the `colData`, which provides randomization and treatment arm information.

```{r}

mae@metadata$calc_vars_pid <- 
  mae@colData |> as_tibble() |> select(pid, site, randomized, ITT, mITT, PP) |> distinct()

mae@metadata$calc_vars_visits <- 
  mae@colData |> 
  as_tibble() |> 
  select(uid, pid, visit_code, study_day, visit_type) |> 
  filter(!is.na(pid), !is.na(study_day)) |>

  # If there is duplicate (pid, study_day) (i.e. multiple visits on the same day), we keep the visit_code and visit_type of the Clinic visit if there is one, otherwise we keep the first one (which is arbitrary but it does not matter because we will predict visit_code and visit_type for all duplicated days in the next step)
  group_by(pid, study_day) |>
  mutate(
    visit_code = case_when(
        any(visit_type == "Clinic") ~ visit_code[visit_type == "Clinic"][1],
        TRUE ~ visit_code[1]
      ),
      visit_type = case_when(
        n() > 1 ~ "Clinic & Home",
        TRUE ~ visit_type[1]
      )
    ) |>
    dplyr::slice(1) |>
    ungroup() |>

    group_by(pid) |>
    complete(study_day = seq(max(0, min(study_day)), min(35, max(study_day)))) |>
    mutate(
      predicted = is.na(visit_code),  # flag pour les jours prédits
      
      # Estimate visit_code and visit_type for missing days
      # Logique : les jours manquants sont forcément des Home visits
      # visit_code prédit : base (ex: 1100) + position dans la semaine
      visit_code = case_when(
    !predicted ~ visit_code,
    TRUE ~ {
      week_block <- floor(study_day / 7)
      position   <- study_day - week_block * 7      # 0-6
      base       <- (week_block + 10) * 100         # 1000, 1100, 1200...
      sprintf("%04d", as.integer(base + position))
    }
  ),

  # visit_type déduit du visit_code :
  # divisible par 100 → Clinic, sinon → Home
  visit_type = case_when(
    !predicted ~ visit_type,
    as.integer(visit_code) %% 100 == 0 ~ "Clinic",
    TRUE ~ "Home"
  ), 
  uid = str_c(pid, visit_code, sep = "_")
  )

mae@metadata$calc_vars_visits <- 
  mae@metadata$calc_vars_visits |> 
  left_join(
    mae@colData |> as_tibble() |> select(pid, site, randomized, ITT, mITT, PP) |> distinct(),
    by = "pid"
  )

```


# Participant-level variables

## Education - can we harmonize between sites ?

We explore the education variable (from CRF2), which is collected in both sites but with different categories. We want to see if we can harmonize this variable between sites, and if not, we will keep it as a site-specific variable.

:::::: {.callout-warning}
For instance, we do not harmonized the education variable between sites because the categories are very different between sites, and there is no clear way to harmonize them. 
:::

```{r}

education <- 
  mae@metadata$participant_crfs_merged |> 
  select(pid, site, edu) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

```

```{r}
#| fig-width: 8
#| fig-height: 2.5
education|> 
  ggplot() +
  aes(y = edu, fill = randomized) +
  facet_wrap(. ~ site, scales = "free") +
  geom_bar() +
  xlab("Number of participants") +
  ylab("") +
  scale_fill_manual(
    name = "",
    values = c("TRUE" = "dodgerblue1", "FALSE" = "gray70"),
    labels = c("TRUE" = "Randomized", "FALSE" = "Not randomized")
  ) +
  theme(legend.position = "top")

```

```{r}
#| fig-width: 8
#| fig-height: 2

education |> 
  filter(randomized) |> 
  ggplot() +
  aes(y = edu) +
  facet_wrap(. ~ site, scales = "free") +
  geom_bar(fill = "dodgerblue1") +
  xlab("Number of participants") +
  ylab("")

```

We add the variable to `mae@metadata$calc_vars_pid`

```{r}
mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(
    education |> select(pid, education = edu) |> distinct(),
    by = "pid"
  )
```

## Economic worry

First, we retrieve the worry variables from CRF2 in the participant_crfs_merged_filtered dataset of the metadata. These variables represent the aspects of life for which participants reported having trouble paying for. The available choices were: *food*, *utilities*, *rent*, *student loans*, *education fees*, *mortgage*, *childcare expenses*, *medical bills*, and *none*. Participants were invited to check all options that applied to them.

We explore the difference between sites in the distribution of this variable, and we also explore the relationship between economic worry (CRF2) and food insecurity (CRF12), which is defined based on three variables food did not last, not afford balanced meals, and cut size meals from CRF12.

:::::: {.callout-warning}
Only three variable were used to define food insecurity (cut size meals, eat less, and hungry did not eat),but six variable are define in CRF12 for food insecuruity. Is it normal ? 
Variable not taken into account : food_did_not_last, not_afford_balanced_meals, often_cut_size_meals
:::

```{r}

worry <- 
  mae@metadata$participant_crfs_merged |> 
  select(pid, site, starts_with("worry_")) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

```

### Exploration of the worry variables

We plot the worry for each site for the randomized participants.

```{r}
#| fig-height: 5
#| fig-width: 6

# create an upset plot to see the combinations of worries
library(UpSetR)

# create a binary matrix for the worries
worry_matrix_CAP <-
  worry |>
  filter(randomized, site == "CAP", !is.na(worry_food)) |>
  select(starts_with("worry_")) |>
  # remove "worry_" prefix from column names
  rename_with(~ str_remove_all(., "worry_")) |>
  mutate(across(everything(), ~ ifelse(. == "Checked", 1, 0))) |>
  as.data.frame()

worry_matrix_MGH <-
  worry |>
  filter(randomized, site == "MGH", !is.na(worry_food)) |>
  select(starts_with("worry_")) |>
  # remove "worry_" prefix from column names
  rename_with(~ str_remove_all(., "worry_")) |>
  mutate(across(everything(), ~ ifelse(. == "Checked", 1, 0))) |>
  as.data.frame()

```

```{r}

# Capturer CAP sans afficher
pdf(NULL)
upset(worry_matrix_CAP,
  nsets = ncol(worry_matrix_CAP),
  nintersects = NA,
  keep.order = TRUE,
  order.by = "freq",
  main.bar.color = "dodgerblue1",
  sets.bar.color = "gray70",
  set_size.show = TRUE
)
grid.edit('arrange', name = 'arrange_CAP')
vp_CAP <- grid.grab()
dev.off()

# Capturer MGH sans afficher
pdf(NULL)
upset(worry_matrix_MGH,
  nsets = ncol(worry_matrix_MGH),
  nintersects = NA,
  keep.order = TRUE,
  order.by = "freq",
  main.bar.color = "coral1",
  sets.bar.color = "gray70",
  set_size.show = TRUE
)
grid.edit('arrange', name = 'arrange_MGH')
vp_MGH <- grid.grab()
dev.off()

# Seul ce plot apparaîtra dans l'output
dev.new()
grid.arrange(vp_CAP, vp_MGH, ncol = 2,
             top = "Worries by site (CAP and MGH)")

```


We explore the correlations between the different types of worries using a multiple correspondence analysis (MCA). We can see that all *unchecked* worries are correlated together, and *checked* worries are correlated together by cluster. We see a first cluster of worries with worry about medical bills, education fees, utilities and mortgage, and a second cluster of worries with worry about food, childcare expenses. Worry about rent and worry about student loans are less correlated with the other worries.

```{r}

worry_mca <- 
  worry |>
  select(-c("pid", "site")) |>
  rename_with(~ str_remove(.x, "^worry_")) |>
  mutate(across(everything(), ~ case_when(
    .x == "Checked" ~ 1,
    .x == "Unchecked" ~ 0,
    TRUE ~ 0
  )|> as.factor()
  )
  )
   
res.mca <- FactoMineR::MCA(worry_mca, graph = FALSE)

factoextra::fviz_screeplot(res.mca, addlabels = TRUE)

```

```{r}

factoextra::fviz_mca_var(
  res.mca,
  repel = TRUE,
  ggtheme = theme_minimal(),
  title = "MCA - Worry variables",
  col.var = case_when(
    str_detect(rownames(res.mca$var$coord), "_0") ~ "steelblue",
    str_detect(rownames(res.mca$var$coord), "_1") ~ "firebrick",
    TRUE ~ "grey60"  # au cas où il reste autre chose
  ),
  labelsize = 4    
) + 
  theme(legend.position = "none")

```


### Creation of a worry score

We will now create a worry_score variable to summarize the different worries reported by each participant. We attribuate more weight to worry about utilisties, rent and food.

```{r}
# weights for worry score food, utilities, rent
w <- 2

worry <- 
  worry |> 
  select(-c("worry_none")) |>
  mutate(
    across(contains("worry"), ~ case_when(
      .x == "Checked" ~ 1,
      .x == "Unchecked" ~ 0,
      TRUE ~ 0)
      ), 
      worry_score = worry_food * w + worry_utilities * w + worry_rent * w + worry_student_loans + worry_education_fees + worry_mortgage + worry_childcare + worry_medical_bills
    ) 

worry |>
  ggplot() +
  aes(x = worry_score, fill = site) +
  geom_histogram(binwidth = 1) +
  facet_wrap(. ~ site, scales = "free") +
  scale_fill_manual(values = c("CAP" = "dodgerblue1", "MGH" = "coral1")) +
  xlab("Worry score") +
  ylab("Number of participants") +
  theme(legend.position = "top")    

```

### Exploration of the relationship between economic worry and food insecurity

Now we explore the relationship between economic worry and food insecurity. We define food insecurity based on these three variables: cut size meals, eat less, and hungry did not eat. We define a participant as food insecure if they answered "Yes" to at least one of these three variables. We then plot the relationship between food insecurity and economic worry (worry_food) for each site.

Some participants report no food insecurity (answer No for cut size meals, eat less, and hungry did not eat) but report worry about food.We also see that the distribution of worry about food is different between sites, with more participants in CAP reporting worry about food compared to MGH.

```{r}

table1_data <- 
  metadata(mae)$participant_crfs_merged |> 
  dplyr::left_join(
    colData(mae) |> as_tibble() |> select(pid, site, arm, randomized, ITT, mITT, PP) |> distinct(), 
    by = join_by(pid, site)
    )  |> 
  filter(randomized) |> 
  select(pid, site, arm, cut_size_meals, eat_less, hungry_did_not_eat) |> 
  mutate(
    food_insecurity = ifelse(
      cut_size_meals == "Yes" | eat_less == "Yes" | hungry_did_not_eat == "Yes",
      "Yes",
      "No"
    ),
    food_insecurity = factor(food_insecurity, levels = c("No", "Yes"))
  ) |> 
  select(-c(eat_less, cut_size_meals, hungry_did_not_eat))

```

```{r}
#| fig-height: 4
#| fig-width: 8

table1_data |> 
  left_join(
    worry |> select(pid, worry_food) |> distinct(),
    by = join_by(pid)
  ) |> 
  mutate(worry_food = worry_food |> recode("Checked" = "Yes", "Unchecked" = "No")) |>
  dplyr::count(site, food_insecurity, worry_food) |> 
  ggplot() +
  aes(x = food_insecurity, y = worry_food, fill = n) +
  geom_tile() +
  geom_text(aes(label = n)) +
  facet_wrap(. ~ site, scales = "free") +
  scale_fill_gradient(low = "white", high = "dodgerblue1") 

```


We add the variable to `mae@metadata$calc_vars_pid`

```{r}
mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(
    worry |> select(pid, worry_score) |> distinct(),
    by = "pid"
  )
```

## Food security

We process food security data collected at the screening visit using the six-item short form of the USDA Household Food Security Survey Module (HFSSM) questionnaire. Six questions were asked to assess food security status over the past 12 months. Based on the number of affirmative responses, we classify participants into high/marginal, low, or very low food security categories.

Affirmative responses are defined as follows:
- Food did not last: "Often true" or "Sometimes true"
- Not afford balanced meals: "Often true" or "Sometimes true"
- Cut size meals: "Yes"
- Often cut size meals: "Almost every month" or "Some months but not every month"
- Eat less: "Yes"
- Hungry did not eat: "Yes"

```{r}

food_security <- 
  mae@metadata$participant_crfs_merged |> 
  select(pid, site, food_did_not_last, not_afford_balanced_meals, cut_size_meals, often_cut_size_meals, eat_less, hungry_did_not_eat) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, randomized) |> 
      distinct(),
    by = join_by(pid)
  ) |> 
  filter(randomized)

```

```{r}

food_security |>
  select(pid, site, 
         food_did_not_last, not_afford_balanced_meals, 
         cut_size_meals, often_cut_size_meals, 
         eat_less, hungry_did_not_eat) |>
  pivot_longer(-c(pid, site), names_to = "question", values_to = "response") |>
  mutate(
    question = factor(question, levels = c(
      "food_did_not_last", "not_afford_balanced_meals",
      "cut_size_meals", "often_cut_size_meals",
      "eat_less", "hungry_did_not_eat"
    ))
  ) |>
  ggplot(aes(x = response, fill = site)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ question, scales = "free_x", ncol = 2) +
  scale_fill_manual(values = c("CAP" = "dodgerblue1", "MGH" = "coral1")) +
  scale_x_discrete(labels = scales::label_wrap(12))
  labs(x = "", y = "Number of participants", title = "Food security items by site") +
  theme(
    legend.position = "top"
  )

```

```{r}

food_security <- 
  food_security |>
  mutate(
    # HH3: "Often true" or "Sometimes true" = affirmative
    aff_food_did_not_last = food_did_not_last %in% c("Often true", "Sometimes true"),
    # HH4: idem
    aff_not_afford_balanced_meals = not_afford_balanced_meals %in% c("Often true", "Sometimes true"),
    # AD1: "Yes" = affirmative
    aff_cut_size_meals = cut_size_meals == "Yes",
    # AD1a: "Almost every month" or "Some months but not every month" = affirmative
    # "Only 1 or 2 months" is NOT affirmative for AD1a (AD1 already counted above)
    aff_often_cut_size_meals = often_cut_size_meals %in% c("Almost every month", "Some months but not every month"),
    # AD2
    aff_eat_less = eat_less == "Yes",
    # AD3
    aff_hungry_did_not_eat = hungry_did_not_eat == "Yes",
    # Raw score = sum of 6 affirmatives (max = 6)
    food_security_score = aff_food_did_not_last + aff_not_afford_balanced_meals +
                        aff_cut_size_meals + aff_often_cut_size_meals +
                        aff_eat_less + aff_hungry_did_not_eat,
    # Categories per USDA guide
    food_security_cat = case_when(
      food_security_score <= 1 ~ "High/Marginal",
      food_security_score <= 4 ~ "Low",
      TRUE                   ~ "Very low"
    ) |> factor(levels = c("High/Marginal", "Low", "Very low"))
  )
```

```{r}

food_security |>
  ggplot(aes(x = food_security_score, fill = site)) +
  geom_bar(position = "dodge") +
  facet_grid(site ~ food_security_cat, scales = "free") +
  scale_fill_manual(values = c("CAP" = "dodgerblue1", "MGH" = "coral1")) +
  scale_x_continuous(breaks = 0:6) +
  labs(x = "Raw score (0–6)", y = "Number of participants",
       title = "Food security score by site") +
  theme(legend.position = "none")

```

We compare the food security score and the worry score

```{r}

food_security |>
  select(pid, site, food_security_score, food_security_cat) |>
  left_join(
    worry |> select(pid, worry_score),
    by = "pid"
  ) |>
  dplyr::count(site, food_security_cat, worry_score) |>
  ggplot() +
  aes(x = food_security_cat, y = worry_score, fill = n) +
  geom_tile() +
  geom_text(aes(label = n)) +
  facet_wrap(. ~ site, scales = "free") +
  scale_fill_gradient(low = "white", high = "dodgerblue1") +
  labs(x = "Food security category", y = "Worry score", title = "Food security category vs worry score by site") +
  theme(legend.position = "none")

```

We add the variable to `mae@metadata$calc_vars_pid`

```{r}
mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(
    food_security |> select(pid, food_security_score, food_security_cat) |> distinct(),
    by = "pid"
  )
```


## Pre-existing conditions

In CRF5 participants were asked about pre-existing conditions. We explore the distribution of this variable and the different types of pre-existing conditions reported by participants. We also explore the relationship between pre-existing conditions and diagnosis.

```{r}

pre_existing_conditions <- 
  mae@metadata$participant_crfs_merged |> 
  select(pid, site, no_pre_existing_condition, diagnosis_agg) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, randomized) |> 
      distinct(),
    by = join_by(pid)
  ) |> 
  filter(randomized)

```

```{r}
#| fig-height: 7

pre_existing_conditions |> 
  mutate(
    diagnosis_agg = case_when(
      no_pre_existing_condition ~ "No pre-existing condition",
      TRUE ~ diagnosis_agg
    ),
    diagnosis_agg = fct_reorder(
      diagnosis_agg, 
      str_count(diagnosis_agg, "&"),
      .fun = dplyr::first,
      .desc = TRUE 
    ),
    diagnosis_agg = fct_relevel(diagnosis_agg, "No pre-existing condition", after = Inf)
  ) |>
  ggplot() +
  aes(y = diagnosis_agg, fill = site) +
  geom_bar() +
  xlab("Number of participants") +
  ylab("") +
  scale_fill_manual(
    name = "Site",
    values = c("CAP" = "dodgerblue1", "MGH" = "coral1")
  ) +
  scale_y_discrete(labels = scales::label_wrap(30)) +
  theme(legend.position = "top")

```


We add the variable to `mae@metadata$calc_vars_pid`

```{r}
mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(
    pre_existing_conditions |> select(pid, no_pre_existing_condition, diagnosis_agg) |> distinct(),
    by = "pid"
  )
```

## Psychological wellbeing - Perceived Stress Scale (PSS)

We are interest in perceived stress of the participants. Questions about their feelings and thoughts during the last month were ask at the screening visit only. Therefore, we consider stress as a time-invariant variable.

The Perceived Stress scale used was developed by Sheldon Cohen and his colleagues in 1983. Nowadays, the PSS is most commonly implemented as the 10-item scale assessing the perceived stress degree based on that individual’s subjective interpretation of their reactions to stressors during a 1 month period.

The PSS-10 scale includes 6 negative items (upset, unable_to_control, nervous_stressed, could_not_cope, outside_your_control, difficulties_piling_up) and 4 positive items (confident, going_your_way, control_irritations, on_top_of_things). Each item is rated on a 5-point Likert scale ranging from 0 (never) to 4 (very often).



```{r}

negative_items <- c("upset", 
                  "unable_to_control", 
                  "nervous_stressed",
                  "could_not_cope",
                  "outside_your_control",
                  "difficulties_piling_up")

positive_items <- c("confident",
                    "going_your_way",
                   "control_irritations",
                   "on_top_of_things") 

stress <- 
  mae@metadata$participant_crfs_merged |>
  select(pid, site, all_of(negative_items), all_of(positive_items)) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, randomized) |> 
      distinct(),
    by = join_by(pid)
  ) |> 
  filter(randomized) |> 
  select(-randomized)



```

```{r}
#| fig-height: 10
#| fig-width: 8

stress |> 
  filter(!is.na(upset)) |>
  pivot_longer(
    cols = c(all_of(negative_items), all_of(positive_items)),
    names_to = "question",
    values_to = "answer"
  ) |> 
  group_by(site, pid) |> 
  mutate(total_score = sum(answer |> as.integer())) |>
  ungroup() |> 
  arrange(total_score) |> 
  mutate(
    pid = pid |> fct_inorder()
  ) |>
  ggplot() +
  aes(y = pid, x = question, fill = answer) +
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  geom_tile() +
  scale_fill_brewer(type = "div", palette = "PRGn") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  )

```

```{r}
#| eval: false

stress |> 
  pivot_longer(
    cols = -c(pid, site),
    names_to = "item",
    values_to = "response"
  ) |>
  mutate(
    response = factor(response, levels = c("0 = Never", "1 = Almost never", "2 = Sometimes", "3 = Fairly often", "4 = Very")), 
    group = case_when(
      item %in% negative_items ~ "Negative items",
      item %in% positive_items ~ "Positive items"
    ) |> factor(), 
    item = factor(item, levels = c(negative_items, positive_items))
  ) |>  
  ggplot() +
  aes(x = response, fill = group) +
  geom_bar() +
  facet_wrap( ~ item, ncol=2) + 
  scale_fill_manual(values = c("Negative items" = "tomato", "Positive items" = "steelblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}

stress |> 
  pivot_longer(
    cols = -c(pid, site),
    names_to = "item",
    values_to = "response"
  ) |>
  mutate(
    response = factor(response, levels = c("0 = Never", "1 = Almost never", "2 = Sometimes", "3 = Fairly often", "4 = Very")), 
    group = case_when(
      item %in% negative_items ~ "Negative items",
      item %in% positive_items ~ "Positive items"
    ) |> factor(), 
    item = factor(item, levels = c(negative_items, positive_items))
  ) |>  
  ggplot() +
  aes(x = response, fill = group) +
  geom_bar() +
  facet_grid(item ~ site, scales = "free_y") + 
  scale_fill_manual(values = c("Negative items" = "tomato", "Positive items" = "steelblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

We explore the relationship between the stress items using a PCA and a correlation matrix. For that we first need to convert the responses to numeric values (0, 1, 2, 3, 4) and reverse the scores for the positive items. The reverse-scoring of the four positively worded items (items 4, 5, 7 and 8: `confident`, `going_your_way`, `control_irritations`, `on_top_of_things`) is done by transforming 0 → 4, 1 → 3, 2 → 2, 3 → 1 and 4 → 0. The six negatively worded items (`upset`, `unable_to_control`, `nervous_stressed`, `could_not_cope`, `outside_your_control`, `difficulties_piling_up`) are scored directly. 

```{r}

stress <- 
  stress |>
  filter(!is.na(upset)) |>
  mutate(across(
    all_of(c(negative_items, positive_items)), 
    ~ as.integer(str_extract(as.character(.x), "^[0-4]"))
  )) |>
  # Inverser les 4 items positifs
  mutate(across(
    all_of(positive_items),
    ~ 4 - .x
  ))

```

```{r}

pca_stress <- 
  stress |>
  select(all_of(c(negative_items, positive_items))) |>
  prcomp(scale. = TRUE)

fviz_eig(pca_stress, addlabels = TRUE) +
  labs(title = "PSS-10 — Scree plot")


item_groups <- factor(
  ifelse(c(negative_items, positive_items) %in% positive_items, 
         "Positive items", 
         "Negative items")
)

fviz_pca_var(
  pca_stress,
  col.var   = item_groups,
  palette   = c("Negative items" = "tomato", "Positive items" = "steelblue"),
  legend.title = "Item type",
  repel     = TRUE,
  title     = "PSS-10 — Variable contributions"
)


fviz_pca_ind(
  pca_stress,
  geom = "point",
  habillage = stress$site,
  palette = c("CAP" = "dodgerblue1", "MGH" = "coral1"),
  addEllipses = TRUE,
  title = "PSS-10 — Participants by site"
)

```



Now we calculate the PSS-10 score by summing all 10 items, yielding a range of 0 to 40 where higher scores indicate greater perceived stress. Scores of 0–13, 14–26 and 27–40 are interpreted as low, moderate and high stress, respectively (Cohen & Janicki-Deverts, 2012). The PSS-10 comprises two subscales: *Perceived Helplessness* (items 1, 2, 3, 6, 9, 10) reflecting feelings of lack of control, and *Lack of Self-Efficacy* (items 4, 5, 7, 8) reflecting perceived inability to handle problems.



```{r}

stress <- 
  stress |> 
  rowwise() |>
  mutate(
    score_helplessness  = sum(c_across(all_of(negative_items))),
    score_self_efficacy = sum(c_across(all_of(positive_items))),
    pss_total           = score_helplessness + score_self_efficacy,
    stress_category     = case_when(
      pss_total <= 13 ~ "Low",
      pss_total <= 26 ~ "Moderate",
      TRUE            ~ "High"
    ) |> factor(levels = c("Low", "Moderate", "High"))
  ) |>
  ungroup()


stress |>
  ggplot(aes(x = pss_total, fill = site)) +
  geom_histogram(binwidth = 2) +
  geom_vline(xintercept = c(13, 26), linetype = "dashed", color = "red") +
  facet_wrap(. ~ site, scales = "free") +
  scale_fill_manual(values = c("CAP" = "dodgerblue1", "MGH" = "coral1")) +
  xlab("PSS total score") +
  ylab("Number of participants") +
  theme(legend.position = "top")

stress |> 
  ggplot(aes(x = stress_category, fill = site)) +
  geom_bar(position = "dodge") +
  facet_wrap(. ~ site, scales = "free") +
  scale_fill_manual(values = c("CAP" = "dodgerblue1", "MGH" = "coral1")) +
  xlab("Stress category") +
  ylab("Number of participants") +
  theme(legend.position = "top")

```


We add the variable to `mae@metadata$calc_vars_pid`

```{r}

mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(
    stress |> select(pid, stress_score = pss_total, stress_category) |> distinct(),
    by = "pid"
  )

```


# Visit-level variables

We create a new assay in the coldata of the `MultiAssayExperiment` object called `calc_vars_visit` that contains one row per participant and visit. We will add to this assay variables related to visit characteristics, exposures, and outcomes.

```{r}

visits_crfs_merged <-
  metadata(mae)$visits_crfs_merged |>
  left_join(
    mae@colData |> as_tibble() |> select(uid, site, visit_type, randomized, ITT, mITT, PP) |> distinct(),
    by = c("uid")
  ) |>
  select(uid, pid, visit_code, study_day, visit_type, everything())


```

## Sexual behavior

The participants’ sexual activity is measured in two different ways: via a prostate-specific antigen test (PSA) and through self-report. PSA are performed on all weekly vaginal swabs except visit 1300 and 1700, and they are used as a biomarker of recent semen exposure. PSA detect semen exposure (sex without condoms) within the past 24-48 hours. Self-reported is measured in the daily dairy and at the weekly visits. In the daily diary, participants report whether they had sex in the past 24 hours and if they used condoms or lubricants, and in the weekly visits, they report the type of partners (male/female/both and single/multiple), if it was a new partner, the frequency of condom and lubricant use, and the frequency of vaginal, oral, the type of sexual activity (vaginal, oral, anal) in the past week and finaly if they had sex in the past 24 hours.

In this study, we are interest in the semen exposure. We define semen exposure as penis-vagina sex without condom use. To capture this exposure as accurately as possible, we combine both sources of information — self-report and PSA — into a single continuous variable ranging from 0 (no exposure) to 1 (confirmed exposure). Self-reported data alone may underestimate exposure due to social desirability bias or recall issues, while PSA alone only captures exposure within the past 48 hours and cannot distinguish between condom-protected and unprotected sex. By integrating both sources, we aim to produce a more complete and robust measure of semen exposure at each study day.

Concretely, we proceed in three steps: (1) we harmonize the self-reported sexual behaviour between daily and weekly sources into consistent variables, (2) we create intermediate boolean variables (`SF_*`) that classify each study day according to the type and protection of sexual activity reported, and (3) we combine these self-reported variables with PSA results to produce the final semen exposure variables.

The goal of this exploratory analysis is to assess the consistency between self-reported sexual behaviour (also between daily responses and weekly responses) and PSA test results and to create new variables that summarize sexual activity based on both self-reported data and PSA results. 



```{r}

sex <- 
  visits_crfs_merged |> 
  select(uid, pid, study_day, visit_code, visit_type, randomized, site, sex_past_24hours.crf33, condom_use.crf33 = condom_use, lubricant_use, sex_type, past_week_sex_partner, sex_past_24hours.crf20, condom_use.crf20 = use_condoms, psa_results)

```

### PSA

PSA was performed on swabs collecetd at the clinic visits 1000, 1200, 1500 and 2120. Results are collected in CRF45. Possible values are "Positive", "Negative", "Invalid" and "not done". We explore the distribution of PSA results across visits and sites.


```{r}
#| fig-height: 8
#| fig-width: 6


sex |> 
  filter(randomized) |>
  filter(!is.na(psa_results)) |>
  ggplot() +
  aes(x = visit_code, y = pid, fill = psa_results) +
  geom_tile() + 
  facet_grid(site ~ ., scales = "free_y", space = "free") + 
  scale_fill_manual(
    values = c(
      "Negative" = "lightgray", 
      "Positive" = "steelblue", 
      "Not done" = "white", 
      "Invalid" = "black")
  )

```


### Self-reported sexual activity

As mentionned above, self-reported sexual activity is measured in the daily diary and at the weekly visits. We have the following source of information :

- From the daily CRF (CRF33) : 
  - `sex_past_24hours.crf33` : self-reported sexual behaviour in the past 24 hours (between visit 1000 and 1500). Participants answer the question "Did you had sex in the past 24 hours" every day. 
  - `condom_use.crf33` : if yes to the previous question, did you use a condom ? (Always, Sometimes, Never)
  - `lubricant_use` : if yes to the first question, did you use lubricant ? (Always, Sometimes, Never)
  - `sex_type` : if yes to the first question, what type of sex ? (Oral, Anal, Vaginal, Other)

- From the weekly CRF (CRF20) :
  - `past_week_sex_partner` : self-reported sexual behaviour in the past week (at visit 1000, 1100, 1200, 1300, 1400 and 1500). Participants answer the question "In the past week, did you have sex with any of the following : a single male partner, multiple male partners, a single female partner, multiple female partners, both, no sex in the past week. 
  - `sex_past_24hours.crf20` : self-reported sexual behaviour in the past 24 hours (each weekly visits). Participants answer the question "Did you had sex in the past 24 hours".
  - `use_condom` : self-reported condom use in the past week (at visit 1000, 1100, 1200, 1300, 1400 and 1500).

#### Daily responses

We first explore the daily responses from CRF33. We are interested in the sexual behaviour of the participants. We look at whether they reported having had sexual intercourse in the last 24 hours, and whether they used lubricant and condoms.

```{r}
#| fig.height: 8
#| fig.width: 10


p1 <- 
  sex |> 
  filter(sex_past_24hours.crf33 == "Yes") |> 
  filter(randomized) |>
  #mutate(pid = pid |> as.character() |> factor(levels = all_pids) |> fct_infreq() |> fct_rev()) |> 
  ggplot() +
  aes(
    x = visit_code |> factor(), y = pid, 
    col = condom_use.crf33, shape = sex_type
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  scale_color_manual(
    "Condom use", 
    values = c("Never" = "black", "Sometimes" = "deeppink4", "Always" = "deeppink1", "Unknown" = "gray70")
  ) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 22, 5, 2)) +
  scale_y_discrete(drop = FALSE) +
  labs(
    x = "Visit code", y = "Participants",
    shape = "Sex type", 
    title = "Condom use"
  ) +
  theme(
    axis.text.y = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), 
    plot.title = element_text(hjust = 0.5)
  )

p2 <- 
  sex |> 
  filter(sex_past_24hours.crf33 == "Yes") |> 
  filter(randomized) |>
  #mutate(pid = pid |> as.character() |> factor(levels = all_pids) |> fct_infreq() |> fct_rev()) |> 
  ggplot() +
  aes(
    x = visit_code |> factor(), y = pid, 
    col = lubricant_use, shape = sex_type
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  scale_color_manual(
    "Lubricant use", 
    values = c("Never" = "black", "Sometimes" = "gold3", "Always" = "gold", "Unknown" = "gray70")
  ) +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 22, 5, 2)) +
  scale_y_discrete(drop = FALSE) +
  labs(
    x = "Visit code", y = "Participants",
    shape = "Sex type", 
    title = "Lubricant use"
  ) +
  theme(
    axis.text.y = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), 
    plot.title = element_text(hjust = 0.5)
  ) 

(p1 + p2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")
```

We look at the relationship between condom use, lubricant use and sex. All participants that reported no sex in the past 24 hours reported "Blank" for condom use and lubricant use. Participants that reported having sex in the past 24 hours reported mainly "Never" for condom use and lubricant use.

```{r}
#| fig.height: 4
#| fig.width: 8

sex |>
  mutate(visit_code = visit_code |> as.numeric()) |>
  filter(visit_type == "Home") |> 
  filter(randomized) |>
  dplyr::count(site, sex_past_24hours.crf33, condom_use.crf33, lubricant_use) |> 
  ggplot() + 
  geom_tile(aes(x = condom_use.crf33, y = lubricant_use, fill = n)) +
  geom_text(aes(x = condom_use.crf33, y = lubricant_use, label = n)) + 
  scale_fill_gradient(low = "white", high = "steelblue1") + 
  facet_grid(site ~ sex_past_24hours.crf33, 
             labeller = as_labeller(c(
               "No" = "No sex in the past 24h", 
               "Yes" = "Sex in the past 24h", 
               "MGH" = "MGH",
               "CAP" = "CAP"
               ))
             ) +
  labs(x = "Condom use", y = "Lubricant use") +
  guides(fill = "none")
```

We look at the relationship between condom use, sex type and sex. All participants that reported no sex in the past 24 hours reported "Blank" for condom use and "NA" for sex type. Participants that reported having sex in the past 24 hours reported mainly "Never" for condom use and "Vaginal" for sex type.

```{r}
#| fig.height: 4
#| fig.width: 8

sex |>
  mutate(visit_code = visit_code |> as.numeric()) |>
  filter(visit_type == "Home") |> 
  filter(randomized) |>
  dplyr::count(site, sex_past_24hours.crf33, condom_use.crf33, sex_type) |> 
  ggplot() + 
  geom_tile(aes(x = condom_use.crf33, y = sex_type, fill = n)) +
  geom_text(aes(x = condom_use.crf33, y = sex_type, label = n)) + 
  scale_fill_gradient(low = "white", high = "steelblue1") + 
  facet_grid(site ~ sex_past_24hours.crf33, 
             labeller = as_labeller(c(
               "No" = "No sex in the past 24h", 
               "Yes" = "Sex in the past 24h", 
                "MGH" = "MGH",
                "CAP" = "CAP"
             ))
             ) +
  labs(x = "Condom use", y = "Sex type") +
  guides(fill = "none")
```

Three participants that reported having sex in the last 24 hours reported "Unknown" for condom use and lubricant use. For two of them, the type of sex was "mounth-vagina".

```{r}
 
sex |> 
  filter(!is.na(sex_past_24hours.crf33)) |>
  filter(randomized) |>
  filter(condom_use.crf33 == "Unknown" | lubricant_use == "Unknown") |>
  select(pid, visit_code, sex_past_24hours.crf33, sex_type, condom_use.crf33, lubricant_use) |> 
  gt(caption = "Zoom on unknown condom/lubricant use")

```


#### Weekly response

At the clinical visits, participants were ask about her sexual behavior in the last 24 hours (`sex_past_24hours.crf20`) and in the past week : her partner in the last week (`past_week_sex_partner`) and if there use use condom in the past week (`use_condom`). 

We first look at the past week sex partner and the condom use. 

```{r}
sex |> 
  filter(visit_type == "Clinic") |> 
  filter(randomized) |>
  select(pid, visit_code, past_week_sex_partner) |>
  distinct() |> 
  dplyr::count(visit_code, past_week_sex_partner) |> 
  pivot_wider(names_from = past_week_sex_partner, values_from = n, values_fill = 0) |>
  gt(caption = "Past week sex partner by visit")
```

```{r}

sex |> 
  filter(visit_type == "Clinic") |> 
  filter(randomized) |>
  select(pid, visit_code, past_week_sex_partner, condom_use.crf20) |> 
  distinct() |> 
  dplyr::count(past_week_sex_partner, condom_use.crf20) |> 
  ggplot() + 
  aes(x = past_week_sex_partner, y = condom_use.crf20, fill = n) +
  geom_tile() + 
  geom_text(aes(label=n)) + 
  scale_fill_gradient(low = "white", high = "steelblue1") + 
  guides(fill = "none") +
  labs(y = "Use condom", x = "Sexual partner(s) in the past week") + 
  theme(strip.text = element_text(size = 12), 
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed()

```


We look at the variable sex in the last 24hours (`sex_past_24hour.crf20`). We can see that the variable is not encoded as in the daily CRF.
Here the variable is encoded as the "sex.type" variable.

```{r}
#| fig.height: 8
#| fig.width: 10

sex |> 
  filter(visit_type == "Clinic") |> 
  filter(randomized) |>
  filter(sex_past_24hours.crf20 !="No") |> 
  ggplot() + 
  aes(x = visit_code |> as.factor(), y = pid, shape = sex_past_24hours.crf20) + 
  geom_point() + 
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  scale_shape_manual(values = c(16, 17, 15, 18, 8, 22, 5, 2)) + 
  labs(
        x = "Visit code", y = "Participants",
        shape = "Sex past 24h\n(=Sex type)"
    )
```

```{r}

sex |>
  filter(visit_type == "Clinic", visit_code != "0000", visit_code != "0010") |>
  filter(randomized) |>
  dplyr::count(site, visit_code, sex_past_24hours.crf20) |> 
  ggplot() +
  aes(y = sex_past_24hours.crf20, x = visit_code |> as_factor()) +
  geom_tile(aes(fill = n)) +
  geom_text(aes(label = n)) +
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  scale_fill_gradient(low = "white", high = "steelblue1") +
  labs(y = "Sex in past 24h", x = "Visit") +
  guides(fill = "none") 

```

We plot the relationship between `past_week_sex_partner`, `condom_use.crf20` and `sex_past_24hours.crf20`. We can see that participants

```{r}
#| fig.height: 8
#| fig.width: 10


sex |> 
  filter(visit_type == "Clinic") |> 
  filter(randomized) |>
  filter(!str_detect(past_week_sex_partner, "No")) |> 
  ggplot() + 
  aes(x = visit_code |> as.factor(), y = pid, shape = past_week_sex_partner, col = condom_use.crf20) + 
  geom_point() + 
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  scale_shape_manual(values = c(16, 17)) +
  scale_color_manual(
    "Condom use", 
    values = c("No" = "black", "Sometimes" = "deeppink4", "Every time" = "deeppink1", "Not applicable" = "gray70")
  ) + 
  labs(
        x = "Visit code", y = "Participants",
        shape = "Past week sexual partner(s)",
        col = "Condom use in the past week"
    )
```


Finaly we look at the consitency between `past_week_sex_partner` and `sex_past_24hours.crf20`. If participant report "Yes" to `sex_past_24hours.crf20`, they should not report "No sex in the past week" to `past_week_sex_partner`. All gray symbols in the plot below should be empty circles (if they reported Not having sex in the past week, they should report "No" to `sex_past_24hours.crf20`). We can see that there are some inconsistencies in the data. All colored symbols should be different from empty circles (if they reported having sex in the past week, they should report a type of sex to `sex_past_24hours.crf20`). We can see that there are some inconsistencies in the data.

:::::: {.callout-note}
Is it possible that the inconsistencies are due to the fact that the question "Did you have sex in the past 24 hours" refers to the 24 hours before the visit, while the question "In the past week, did you have sex with any of the following" refers to the past week and not the 7 days before the visit ? 
:::

```{r}
#| fig.height: 8
#| fig.width: 10


sex |> 
  filter(visit_type == "Clinic") |> 
  filter(randomized) |>
  #filter(sex_past_24hours.crf20 !="No") |> 
  ggplot() + 
  aes(x = visit_code |> as.factor(), y = pid, shape = sex_past_24hours.crf20, colour = past_week_sex_partner) + 
  geom_point() + 
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  scale_shape_manual(values = c(1, 17, 15, 18, 8, 22, 5, 2)) + 
  scale_color_manual(values = c("gray70", "orchid1", "orchid4")) +
  labs(
        x = "Visit code", y = "Participants",
        shape = "Did you have sex in the past 24 hour ?",
        col = "In the past week have you had sex with"
    )
```

```{r}
#| eval: false

sex |> 
  filter(visit_type == "Clinic") |> 
  group_by(visit_code) |> 
  dplyr::count(
    (sex_past_24hours.crf20 != "No" & !is.na(sex_past_24hours.crf20)) & past_week_sex_partner == "No sex in the past week"
  ) 

```



#### Harmonization between daily and weekly responses

We merged the information from sex_past_24hours from the daily CRF (CRF33) and the weekly crf (CRF20) in one single variable `sex_past_24hours`.

`sex_past_24hours` take the value of `sex_past_24hours.crf33` for the daily responses and take the value "Yes" if `sex_past_24hours.crf20` is not "No" or missing for the weekly responses. We also harmonized the value of `sex_type`for the clinical and home visits. Since we don't have information about condom use in the weekly CRF, if `sex_past_24hours.crf20` is not "No", we set `condom_use.crf33` to "Unknown" for those visits. Same for `lubricant_use`.



```{r}

# Harmonizing daily and weekly responses
# merged daily and weekly
# sex_past_24hours in the weekly crf20 is encoded as the sex_type
# we also complete the sex_type variable with this information

sex <- 
  sex |> 
  mutate(
    sex_past_24hours = case_when(
      !is.na(sex_past_24hours.crf33) ~ sex_past_24hours.crf33,
      !is.na(sex_past_24hours.crf20) & sex_past_24hours.crf20 != "No" ~ "Yes",
      !is.na(sex_past_24hours.crf20) & sex_past_24hours.crf20 == "No"  ~ "No",
      TRUE ~ NA_character_
    ),
    
    condom_use = case_when(
      !is.na(sex_past_24hours.crf20) & sex_past_24hours.crf20 != "No" ~ "Unknown",
      TRUE ~ condom_use.crf33
    ),
    
    lubricant_use = case_when(
      !is.na(sex_past_24hours.crf20) & sex_past_24hours.crf20 != "No" ~ "Unknown",
      TRUE ~ lubricant_use
    ),
    
    sex_type = case_when(
      sex_past_24hours.crf20 == "penis_vagina" ~ "penis-vagina", 
      sex_past_24hours.crf20 == "oral_sex" ~ "mouth-vagina",
      sex_past_24hours.crf20 == "fingering, penis_vagina" ~ "penis-vagina, finger-vagina", 
      sex_past_24hours.crf20 == "oral_sex, penis_vagina" ~ "penis-vagina, mouth-vagina",
      sex_past_24hours.crf20 == "fingering, oral_sex, penis_vagina" ~ "penis-vagina, finger-vagina, mouth-vagina",
      sex_past_24hours.crf20 == "anal_sex" ~ "anal",
      TRUE ~ sex_type
    )
  )

#check
sex |> 
  dplyr::count(sex_past_24hours, sex_past_24hours.crf33, sex_past_24hours.crf20, sex_type) |> 
  gt() |> 
  cols_label(
    sex_past_24hours = "sex_past_24hours (harmonized)",
    sex_past_24hours.crf33 = "sex_past_24hours (CRF33 - daily)",
    sex_past_24hours.crf20 = "sex_past_24hours (CRF20 - weekly)",
    sex_type = "sex_type",
    n = "Count"
  )
 
```

Some participants have the same `study_day` for both a clinic visit and a home (daily diary) visit. This occurs when a clinical visit and a daily diary entry fall on the same calendar day. In those cases, we have two rows for the same participant and study day, with potentially conflicting information on sexual behaviour. 

We identified `r sex |> dplyr::count(pid, study_day) |> filter(n > 1) |> nrow()` such duplicated `(pid, study_day)` pairs, involving `r sex |> dplyr::count(pid, study_day) |> filter(n > 1) |> distinct(pid) |> nrow()` participants.

To resolve these duplicates, we apply the following priority rules and keep one row per `(pid, study_day)` :

- **`sex_past_24hours`** : if the clinic visit reports `Yes`, it overrides a `No` from the daily diary (the clinical visit is considered more reliable in case of conflict). Otherwise, the daily diary value prevails.
- **`condom_use` / `sex_type`** : follow the same priority as `sex_past_24hours`.
- **`psa_results`** : always taken from the clinic visit.
- **`visit_code`** : kept from the clinic visit.
- **`visit_type`** : set to `"Clinic & Home"` when both exist.


```{r}
#| eval: false
# Display values of duplicate study_day for a participant

sex |>
  dplyr::count(pid, study_day) |>
  filter(n > 1) |>
  left_join(sex, by = c("pid", "study_day")) |>
  select(pid, study_day, sex_past_24hours, condom_use, sex_type, psa_results) |>
  arrange(pid, study_day) |>
  gt(caption = "Duplicated (pid, study_day) before harmonisation") 


```

```{r}

sex <- 
  sex |>
  group_by(pid, study_day) |>
  mutate(
    # Sex : Clinic prime si Clinic = Yes, sinon Home prime
    sex_past_24hours = case_when(
      any(visit_type == "Clinic" & sex_past_24hours == "Yes") ~ "Yes",
      any(visit_type == "Home")   ~ sex_past_24hours[visit_type == "Home"][1],
      TRUE ~ sex_past_24hours[1]
    ),
    
    # Condom : si Clinic a forcé sex = Yes → Unknown, sinon Home prime
    condom_use = case_when(
      any(visit_type == "Clinic" & sex_past_24hours == "Yes") & 
        any(visit_type == "Home" & sex_past_24hours == "No") ~ "Unknown", # Clinic a primé
      any(visit_type == "Home" & sex_past_24hours == "Yes") ~ condom_use[visit_type == "Home" & sex_past_24hours == "Yes"][1],
      TRUE ~ condom_use[1]
    ),
    
    # Sex type : même logique que condom
    sex_type = case_when(
      any(visit_type == "Clinic" & sex_past_24hours == "Yes") &
        any(visit_type == "Home" & sex_past_24hours == "No") ~ sex_type[visit_type == "Clinic"][1],
      any(visit_type == "Home" & sex_past_24hours == "Yes") ~ sex_type[visit_type == "Home" & sex_past_24hours == "Yes"][1],
      TRUE ~ sex_type[1]
    ),
    
    # PSA toujours depuis Clinic
    psa_results = case_when(
      any(visit_type == "Clinic") ~ psa_results[visit_type == "Clinic"][1],
      TRUE ~ psa_results[1]
    ),
    
    # Metadata : visit_code Clinic, visit_type concaténé si doublon
    visit_code = case_when(
      any(visit_type == "Clinic") ~ visit_code[visit_type == "Clinic"][1],
      TRUE ~ visit_code[1]
    ),
    visit_type = case_when(
      n() > 1 ~ "Clinic & Home",
      TRUE ~ visit_type[1]
    )
  ) |>
  dplyr::slice(1) |>  # une seule ligne par (pid, study_day)
  ungroup()


# Vérifier les cas Clinic & Home
sex |> 
  filter(visit_type == "Clinic & Home") |>
  dplyr::count(sex_past_24hours, condom_use) |>
  gt(caption = "Duplicated (pid, study_day) after harmonisation - Clinic & Home") |>
  cols_label(
    sex_past_24hours = "Sex in past 24h (harmonized)",
    condom_use = "Condom use (harmonized)",
    n = "Count"
  )

```


#### Comparing daily and weekly responses

::: {.callout-note}
For instance, the inconsistency are just reported but no action is taken to resolve them. 
:::

We first look at the consitency between the variable `sex_past_24hours` from CRF33 and the variable `past_week_sex_partner` from CRF20. If participant report "Yes" to `sex_past_24hours`, they should not report "No sex in the past week" to `past_week_sex_partner`.

```{r}
#| eval: false
#| fig-height: 10
#| fig-width: 10

sex |> 
  filter(!is.na(visit_code)) |> 
  ggplot() + 
  aes(x = visit_code, y = pid) +
  geom_point(data = sex |> 
               filter(sex_past_24hours.crf33 == "Yes"), 
             aes(fill = sex_past_24hours.crf33),
             size = 2.5, alpha = 0.5, shape = 21) +
  geom_point(data = sex |> 
               filter(!is.na(past_week_sex_partner)), 
             aes(color = past_week_sex_partner), 
             size = 2.5) +
  scale_fill_manual(
    name = "Sex in past 24h",
    values = c("Yes" = "orchid1", "No" = "white")
  ) +
  scale_color_manual(
    name = "Sex partner past week",
    values = c("No sex in the past week" = "grey80", 
               "A single male partner" = "orchid1", 
               "Multiple male partners" = "orchid3")
  ) +
  labs(x = "Visit", y = "Participant ID") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


```{r}

daily_summary <- 
  sex |>
  filter(visit_type %in% c("Home", "Clinic & Home")) |>
  mutate(period = str_sub(visit_code, 1, 2)) |>
  group_by(pid, period) |>
  summarise(
    any_daily_sex = any(sex_past_24hours.crf33 == "Yes", na.rm = TRUE),
    all_daily_no_sex = all(sex_past_24hours.crf33 == "No", na.rm = TRUE),
    n_daily_visits = n(),
    .groups = "drop"
  )


sex |>
  filter(visit_type == "Clinic") |>
  mutate(period = str_sub(visit_code, 1, 2)) |>
  select(pid, period, past_week_sex_partner) |> 
  left_join(daily_summary, by = c("pid", "period")) |>
  filter(period < 16 & period !="00") |> 
  mutate(
    inconsistency_1 = any_daily_sex & (past_week_sex_partner == "No sex in the past week"),
    inconsistency_2 = all_daily_no_sex & (past_week_sex_partner != "No sex in the past week")
  ) |> 
  dplyr::count(inconsistency_1, inconsistency_2) |> 
  gt(caption = "Table of inconsistencies between daily and weekly sexual behaviour reports between visits 1000 and 1500") |> 
  cols_label(
    inconsistency_1 = "Daily sex but no weekly",
    inconsistency_2 = "Weekly sex but no daily",
    n = "Count"
  ) 


```

We do the same with the condom use. At the weekly clinical visits, participants were ask if they used condom in the past week (In the past week, if you had male partners, did you use condoms? - `use_condom`). This should be consistent with the daily reports of condom use in the past 24 hours.

```{r}
#| eval: false
#| fig-height: 10
#| fig-width: 10
sex |> 
  filter(!is.na(visit_code)) |> 
  ggplot() + 
  aes(x = visit_code, y = pid) +
  geom_point(data = sex |> 
               mutate(condom_daily = ifelse(condom_use.crf33 %in% c("Always", "Sometimes"), "Yes", "No")) |>
               filter(condom_daily == "Yes"), 
             aes(fill = condom_daily),
             size = 2.5, alpha = 0.5, shape = 21) +
  geom_point(data = sex |> 
               mutate(condom_weekly = ifelse(condom_use.crf20 %in% c("Every time", "Sometimes"), "Yes", "No")) |>
               filter(condom_weekly == "Yes"),
             aes(color = condom_weekly), 
             size = 2.5) +
  labs(x = "Visit", y = "Participant ID") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```


```{r}

daily_summary <- 
  sex |>
  filter(visit_type %in% c("Home", "Clinic & Home")) |>
  mutate(period = str_sub(visit_code, 1, 2)) |>
  group_by(pid, period) |>
  summarise(
    any_daily_condom= any(condom_use.crf33 %in% c("Always", "Sometimes"), na.rm = TRUE),
    all_daily_no_condom = all(condom_use.crf33 %in% c("Never", "Unknown", "Blank"), na.rm = TRUE),
    n_daily_visits = n(),
    .groups = "drop"
  )


sex |>
  filter(visit_type == "Clinic") |>
  mutate(period = str_sub(visit_code, 1, 2)) |>
  select(pid, period, condom_use.crf20) |>
  mutate(
    past_week_condom = case_when(
      condom_use.crf20 %in% c("Every time", "Sometimes") ~ "Yes condom in past week",
      condom_use.crf20 %in% c("Never", "Not applicable") ~ "No condom in past week",
      TRUE ~ NA_character_
    )
  ) |>
  left_join(daily_summary, by = c("pid", "period")) |>
  filter(period < 16 & period !="00") |> 
  mutate(
    inconsistency_1 = any_daily_condom & (past_week_condom == "No condom in past week"),
    inconsistency_2 = all_daily_no_condom & (past_week_condom == "Yes condom in past week")
  ) |> 
  dplyr::count(inconsistency_1, inconsistency_2) |> 
  gt(caption = "Table of inconsistencies between daily and weekly condom use reports between visits 1000 and 1500") |> 
  cols_label(
    inconsistency_1 = "Any condom use daily but no reported weekly",
    inconsistency_2 = "No condom use daily but reported weekly",
    n = "Count"
  ) 

```

#### SF Semen Exposure Variables

Since our final goal is to assess semen exposure, we create boolean summary variables for the self - declared semen exposure.

*Note : pv is for penis-vagina*

-  `SF_pv_sex_w_semen_exposure` : TRUE if `sex_past_24hours` is "Yes", the type of sex is penis-vagina and `condom_use` is "Sometimes" or "Never".

- `SF_pv_sex_wo_semen_exposure` : TRUE if `sex_past_24hours` is "Yes", the type of sex is penis-vagina and the condom use is always.

- `SF_pv_sex_unknown_semen_exposure` : TRUE if `sex_past_24hours` is "Yes", the type of sex is penis-vagina and the condom use is unknonw, missing or blank

- `SF_other_sex_wo_semen_exposure` : TRUE if `sex_past_24hours` is "Yes", the type of sex is not penis-vagina. 

If a participant have FALSE values for all the SF variables, it means they did not have any sexual activity in the past 24 hours.

```{r}

# Creating SF_semen_exposure

sex <- 
  sex |> 
  mutate(
    SF_pv_sex_w_semen_exposure = sex_past_24hours == "Yes" & str_detect(sex_type, "penis-vagina") & (condom_use == "Never" | condom_use == "Sometimes"),
    SF_pv_sex_wo_semen_exposure = sex_past_24hours == "Yes" & str_detect(sex_type, "penis-vagina") & condom_use == "Always",
    SF_pv_sex_unknown_semen_exposure = sex_past_24hours == "Yes" & str_detect(sex_type, "penis-vagina") & (is.na(condom_use) | condom_use == "Unknown" | condom_use == "Blank"),
    SF_other_sex_wo_semen_exposure = sex_past_24hours == "Yes" & str_detect(sex_type, "finger-vagina|mouth-vagina|anal")
  )

# check 
sex |> 
  dplyr::count(sex_past_24hours, sex_type, condom_use, SF_pv_sex_w_semen_exposure, SF_pv_sex_wo_semen_exposure, SF_pv_sex_unknown_semen_exposure, SF_other_sex_wo_semen_exposure) |> 
  gt(caption = "Table of self-reported semen exposure variables") |> 
  cols_label(
    sex_past_24hours = "Sex in\npast 24h",
    condom_use = "Condom\nuse",
    sex_type = "Sex type",
    SF_pv_sex_w_semen_exposure = "SF pv sex w\nsemen exposure",
    SF_pv_sex_wo_semen_exposure = "SF pv sex wo\nsemen exposure",
    SF_pv_sex_unknown_semen_exposure = "SF pv sex unknown\nsemen exposure",
    SF_other_sex_wo_semen_exposure = "SF other sex wo\nsemen exposure",
    n = "Count"
  )

```

```{r}
#| fig.height: 8
#| fig.width: 10


sex |> 
  pivot_longer(cols = starts_with("SF_"), names_to = "semen_exposure", values_to = "value") |> 
  filter(value == TRUE) |> 
  mutate(semen_exposure = semen_exposure |> factor(levels = c("SF_pv_sex_w_semen_exposure", "SF_pv_sex_wo_semen_exposure", "SF_pv_sex_unknown_semen_exposure", "SF_other_sex_wo_semen_exposure"))) |> 
  ggplot() +
  aes(y = pid, x = visit_code, col = semen_exposure, shape = semen_exposure) +
  geom_point() + 
  facet_grid(site ~ semen_exposure, scales = "free_y") + 
  scale_color_manual(values = c("deeppink", "orange", "gray70", "black")) + 
  scale_shape_manual(values = c(16, 17, 18, 4)) 

sex |> 
  pivot_longer(cols = starts_with("SF_"), names_to = "semen_exposure", values_to = "value") |> 
  filter(value == TRUE) |> 
  mutate(semen_exposure = semen_exposure |> factor(levels = c("SF_pv_sex_w_semen_exposure", "SF_pv_sex_wo_semen_exposure", "SF_pv_sex_unknown_semen_exposure", "SF_other_sex_wo_semen_exposure"))) |> 
  ggplot() +
  aes(y = pid, x = visit_code, col = semen_exposure, shape = semen_exposure) +
  geom_point() + 
  facet_grid(site ~ ., scales = "free_y") + 
  scale_color_manual(values = c("deeppink", "orange", "gray70", "black")) + 
  scale_shape_manual(values = c(16, 17, 18, 4)) 

```


We created new variables to assess the time since last sexual intercourse (with or without condom) at each study day using the `sex_past_24hours` and `condom_use` variables.

- `day_since_last_sex_w_cndm` : count the number of day since the last self-report (penis-vagina) sexual intercourse with condom. 

- `days_since_last_sex_wo_cndm` : count the number of day since the last self-report (penis-vagina) sexual intercourse without condom. 

- `days_since_last_sex_unknown_cndm` : count the number of day since the last self-report (penis-vagina) sexual intercourse with unknown condom use. 

:::::: {.callout-note}
We don't have the information about the last sexual intercourse for the first visits until the first reported sexual intercourse. `day_since_last_sex` remains NA until the first reported sexual intercourse. 

Should I impute the last sexual intercourse for the first visit (1000) ? 
A solution could be to look at the variable `sexual_partners_past_month` and `sex_past_24hours` collected at the beginning of the study. If the participant reported no sexual partners in the last month and no sexual activity in the past 24 hours, we could impute the last sexual intercourse as 30 days before the visit. If the participant reported having a sexual partner in the last month but no sexual activity in the past 24 hours, we could impute the last sexual intercourse as 15 days before the visit.
:::


```{r}
# Approximate the last sexual intercourse before the study start

sex <- 
  sex |>
  arrange(pid, study_day) |>
  group_by(pid) |> 
  mutate(
  # with condom only
    last_sex_day_w_cndm = case_when(
      sex_past_24hours == "Yes" & str_detect(sex_type, "penis-vagina") & condom_use == "Always" ~ study_day,
      TRUE ~ NA_integer_
    ),
    last_sex_day_w_cndm = zoo::na.locf(last_sex_day_w_cndm, na.rm = FALSE), 
    days_since_last_sex_w_cndm = study_day - last_sex_day_w_cndm,
    
    # without condom only
    last_sex_day_wo_cndm = case_when(
      sex_past_24hours == "Yes" & str_detect(sex_type, "penis-vagina") & (condom_use == "Never"|condom_use == "Sometimes") ~ study_day,
      TRUE ~ NA_integer_
    ),
    last_sex_day_wo_cndm = zoo::na.locf(last_sex_day_wo_cndm, na.rm = FALSE),
    days_since_last_sex_wo_cndm = study_day - last_sex_day_wo_cndm, 

    # unknown condom use
    last_sex_day_unknown_cndm = case_when(
      sex_past_24hours == "Yes" & str_detect(sex_type, "penis-vagina") & (is.na(condom_use)|condom_use == "Unknown"|condom_use == "Blank") ~ study_day,
      TRUE ~ NA_integer_
    ),
    last_sex_day_unknown_cndm = zoo::na.locf(last_sex_day_unknown_cndm, na.rm = FALSE),
    days_since_last_sex_unknown_cndm = study_day - last_sex_day_unknown_cndm
  ) |> 
  ungroup() |> 
  select(-c(last_sex_day_wo_cndm, last_sex_day_w_cndm, last_sex_day_unknown_cndm))


```

Numbers and colors represent the number of day since the last sexual intercourse without condom


```{r}
#| fig.height: 10
#| fig.width: 10

sex |> 
  filter(randomized) |>
  ggplot() + 
  aes(x = visit_code |> as.factor(), y = pid, fill = days_since_last_sex_wo_cndm) +
  geom_tile() +
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  geom_text(aes(label = days_since_last_sex_wo_cndm), size = 2) + 
  scale_fill_gradient("Day since last sex\n(wo cndm)", low = "#EBFAFE", high =  "lightskyblue4") +
  labs(x = "Visit code", y="") +
  theme(
    #axis.text.y = element_blank(),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

```


### PSA and self-reported sexual activity


We compare the PSA results and the self-reported sexual behavior (`SF_pv_sex_w_semen_exposure` and `SF_pv_sex_unknown_semen_exposure`) to look at the consistency between those two variables.

```{r}
#| fig-height: 10
#| fig-width: 10

sex |> 
  filter(!is.na(visit_code)) |> 
  filter(randomized) |>
  ggplot() + 
  aes(x = visit_code, y = pid) +
  geom_point(data = sex |> 
               filter(SF_pv_sex_w_semen_exposure | SF_pv_sex_unknown_semen_exposure) |> 
               pivot_longer(cols = c(SF_pv_sex_w_semen_exposure, SF_pv_sex_unknown_semen_exposure), names_to = "semen_exposure", values_to = "value") |>
               filter(value == TRUE),
             aes(fill = semen_exposure),
             size = 2.5, alpha = 0.5, shape = 21) +
  geom_point(data = sex |> 
               filter(psa_results == "Positive"), 
             aes(shape = psa_results), 
             size = 2.5) +
  geom_vline(
    data = sex |>  select(visit_code, visit_type) |> filter(visit_type == "Clinic") |> distinct(),
    aes(xintercept = visit_code),
    color = "gray70",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  scale_fill_manual(
    name = "Sex in past 24h",
    values = c("gray70", "deeppink"),
  ) +
  scale_shape_manual(
    name = "PSA results", 
    values = c("Positive" = 4)
  )  +
  labs(x = "Visit", y = "Participant ID") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


#### Table of PSA and SF-sex

We take into account the time since last sexual intercourse and look at the consistency between PSA results and self-reported sexual behaviour depending on the time since last sexual intercourse.

We don't take into account the visit 0000 since we don't have information about their sexual behaviour : CRF33 and CRF20 where not fill at that visit. We just have information about PSA and we imputed the day since last sex using this information. 

:::callout-note
At the 1000th visit, the daily CRFs had not yet begun to be filled out: we therefore only have information from clinical visits, which did not provide details on condom use. The variable day_since_last_sex_wo_condom could not therefore be ‘updated’ since visit 0000 (at which this variable was initialised to 0 for participants with a positive PSA): it is therefore normal to have only values >4.
:::

Some participants that have a PSA positive but a day since last sex > 4 had report, at that weekly visit, a sexual intercourse in the last 24h BUT for the weekly visit we don't have the information about the condom use (so the variable `day_since_last_sex_wo_cndm`was not update)

```{r}

sex |> 
  #filter(randomized) |>
  filter(visit_type %in% c("Clinic", "Clinic & Home")) |>
  group_by(pid) |>
  mutate(last_sex_wo_cndm_cat = case_when(
    is.na(days_since_last_sex_wo_cndm) ~ "Unknown",
    days_since_last_sex_wo_cndm == 0 ~ "0 day",
    days_since_last_sex_wo_cndm == 1 ~ "1 day",
    days_since_last_sex_wo_cndm == 2 ~ "2 days",
    days_since_last_sex_wo_cndm == 3 ~ "3 days",
    days_since_last_sex_wo_cndm >= 4 ~ "4+ days",
    TRUE ~ NA_character_
  )) |>
  ungroup() |>
  select(pid, visit_code, last_sex_wo_cndm_cat, psa_results, visit_type) |>
  distinct() |>
  filter(visit_type %in% c("Clinic", "Clinic & Home"), visit_code > "0000", visit_code != "1000") |>
  filter(!is.na(psa_results)) |> 
  dplyr::count(visit_code, last_sex_wo_cndm_cat, psa_results) |>
  ggplot(aes(x = psa_results, y = last_sex_wo_cndm_cat, fill = n)) +
  geom_tile() +
  facet_grid(~ visit_code) +
  geom_text(aes(label = n)) +
  scale_fill_gradient(low = "white", high = "steelblue1") +
  ylab("Last sex without condom") +
  xlab("Psa result") + 
  guides(fill = "none") +
  theme(strip.text = element_text(size = 12), 
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed()


```

Let's do the same plot but taking the variable `day_since_last_sex_unknown_cndm` to account for the fact that at the weekly visit we don't have information about the condom use. If a participant reported a sexual intercourse in the last 24h at the weekly visit, we set the variable `day_since_last_sex_unknown_cndm` to 0 for that visit, even if we don't know if they used condom or not.

```{r}

sex |> 
  filter(randomized) |>
  filter(visit_type == "Clinic") |>
  group_by(pid) |>
  mutate(last_sex_wo_cndm_cat = case_when(
    is.na(days_since_last_sex_wo_cndm) ~ "Unknown",
    days_since_last_sex_wo_cndm == 0 | days_since_last_sex_unknown_cndm == 0 ~ "0 day",
    days_since_last_sex_wo_cndm == 1 | days_since_last_sex_unknown_cndm == 1 ~ "1 day",
    days_since_last_sex_wo_cndm == 2 | days_since_last_sex_unknown_cndm == 2 ~ "2 days",
    days_since_last_sex_wo_cndm == 3 | days_since_last_sex_unknown_cndm == 3 ~ "3 days",
    days_since_last_sex_wo_cndm >= 4 | days_since_last_sex_unknown_cndm >= 4 ~ "4+ days",
    TRUE ~ NA_character_
  )) |>
  ungroup() |>
  select(pid, visit_code, last_sex_wo_cndm_cat, psa_results, visit_type) |>
  distinct() |>
  filter(visit_type == "Clinic", visit_code > "0000", visit_code != "1000") |>
  filter(!is.na(psa_results)) |> 
  dplyr::count(visit_code, last_sex_wo_cndm_cat, psa_results) |>
  ggplot(aes(x = psa_results, y = last_sex_wo_cndm_cat, fill = n)) +
  geom_tile() +
  facet_grid(~ visit_code) +
  geom_text(aes(label = n)) +
  scale_fill_gradient(low = "white", high = "steelblue1") +
  ylab("Last sex without condom") +
  xlab("Psa result") + 
  guides(fill = "none") +
  theme(strip.text = element_text(size = 12), 
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_fixed()

```

### Create summary variable for semen exposure 

```{r}

psa_correction_days <- 
  sex |>
  filter(psa_results == "Positive") |>
  select(pid, study_day) |>
  reframe(correction_day = c(study_day, study_day - 1, study_day - 2), 
          .by = c(pid, study_day)) |>
  select(pid, correction_day) |>
  distinct() |>
  mutate(psa_positive_nearby = TRUE)

sex <- 
  sex |>
  mutate(
    # conversion booléen → numérique
    pv_sex_w_semen_exposure  = as.numeric(SF_pv_sex_w_semen_exposure  == TRUE),
    pv_sex_wo_semen_exposure = as.numeric(SF_pv_sex_wo_semen_exposure == TRUE),
    other_sex_wo_semen_exp   = as.numeric(SF_other_sex_wo_semen_exposure == TRUE),
    
    # Résolution du "unknown" aux visites weekly avec le PSA
    pv_sex_w_semen_exposure = case_when(
      SF_pv_sex_unknown_semen_exposure & psa_results == "Positive" ~ 0.5,
      TRUE ~ pv_sex_w_semen_exposure
    ),
    pv_sex_wo_semen_exposure = case_when(
      SF_pv_sex_unknown_semen_exposure & psa_results == "Negative" ~ 1,
      TRUE ~ pv_sex_wo_semen_exposure
    )
  ) |>
  
  # Correction PSA pour les jours entourant un PSA positif
  left_join(psa_correction_days, by = c("pid", "study_day" = "correction_day")) |>
  mutate(
    pv_sex_w_semen_exposure = case_when(
      psa_positive_nearby & pv_sex_w_semen_exposure == 0 ~ 0.5,
      TRUE ~ pv_sex_w_semen_exposure
    )
  ) |>
  select(-psa_positive_nearby)

```

```{r}
#| fig.height: 10
#| fig.width: 10


sex |>
  filter(randomized, study_day >= 0) |>
  pivot_longer(
    cols = c(pv_sex_w_semen_exposure, pv_sex_wo_semen_exposure, other_sex_wo_semen_exp),
    names_to = "variable", values_to = "value"
  ) |>
  ggplot(aes(x = study_day, y = pid, fill = value)) +
  geom_tile() +
  # Overlay PSA positif
  geom_point(
    data = sex |> filter(psa_results == "Positive", study_day >= 0),
    aes(x = study_day, y = pid), 
    inherit.aes = FALSE,
    shape = 4, size = 2, color = "red"
  ) +
  facet_grid(variable ~ ., scales = "free_y") +
  scale_fill_gradient("Value", low = "white", high = "steelblue", na.value = "grey90") +
  geom_vline(
    data = sex |> filter(visit_type == "Clinic", study_day >= 0) |> distinct(study_day),
    aes(xintercept = study_day), color = "gray50", linetype = "dashed", linewidth = 0.3
  ) + 
  scale_x_continuous(breaks = seq(0, 100, by = 7))

```

```{r}

sex |>
  filter(randomized, study_day >= 0) |>
  pivot_longer(
    cols = c(pv_sex_w_semen_exposure, pv_sex_wo_semen_exposure, other_sex_wo_semen_exp),
    names_to = "variable", values_to = "value"
  ) |>
  ggplot(aes(x = study_day, y = pid, col = value, shape = variable)) +
  geom_point()
  

```

We add the summary variables for sexual activity and semen exposure to the `calc_vars_visits` table in the metadata of the `mae` object.

```{r}
mae@metadata$calc_vars_visits <- 
  mae@metadata$calc_vars_visits |> 
  left_join(
    sex |> 
      select(pid, study_day, pv_sex_w_semen_exposure,  pv_sex_wo_semen_exposure, other_sex_wo_semen_exp),
    by = join_by(pid, study_day)
  )
```

## Metronidazole use

```{r}
#| eval: false

mtz_crf_raw <- 
  mae@metadata$crf_data_raw$crf22 |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, site, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

mtz_crf_raw |> 
  dplyr::count(dfseq)


mtz_crf_raw |> 
  filter(randomized) |> 
  dplyr::count(dfseq)


mtz_crf_raw |> 
  filter(randomized) |> 
  dplyr::filter(dfseq == 1200) 

mtz_crf_raw |> 
  filter(randomized) |> 
  dplyr::count(pid) |> 
  arrange(-n) |> 
  dplyr::count(n)

mtz_crf_raw |> 
  group_by(pid) |> 
  mutate(min_study_day = min(study_day)) |> 
  ungroup() |> 
  mutate(pid = pid |> fct_reorder(min_study_day)) |>
  ggplot() +
  aes(x = study_day, y = pid, col = dfseq |> as.character()) +
  geom_point()

mae@metadata$exposures |> 
  ggplot() +
  aes(x = study_day, y = pid, col = metronidazole |> factor()) + geom_point() +
  scale_color_manual(
    name = "Metronidazole use",
    na.value = "lightgray",
    values = c("0" = "red", "1" = "purple", "2" = "blue")
  ) +
  geom_point()
  

```

## Study product exposure related variables


### Number of self-reported study product doses taken


```{r}
#| eval: false

n_self_reported_doses <- 
  metadata(mae)$exposures |>
  group_by(pid) |>
  summarize(
    n_product_doses = sum(study_product |> as.integer(), na.rm = TRUE),
    n_product_doses_PP = sum(study_product_PP, na.rm = TRUE),
    comments_studyproduct = str_c(comments_studyproduct[!is.na(comments_studyproduct)], collapse = "; "),
    .groups = "drop"
  )


mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(n_self_reported_doses, by = join_by(pid))

rm(n_self_reported_doses)

```


### Time since last self-reported study product doses at Week 2 in-clinic visit


We document the time between the last dose taken prior to the in-person visit at week 2 and the swab collected at that visit. This is relevant because the primary outcome is detection of the LBP strain at the week 2 visit, and the time since last dose may influence the likelihood of detection.

Information on the self-reported time of last study product use comes from `exposures`, while the study day of the week 2 in-person visit is taken from `visits_crfs_merged.`

We define the following categories for the timing of the last study product use:
-   "same day": last dose taken on the same day as the visit (before 6 p.m.)
-   "day before": last dose taken the day before the visit
-   "more than 1 day": last dose taken more than one day before the visit


```{r}
#| eval: false

study_product <-
  mae@colData |>
  as_tibble() |>
  select(pid, site, randomized, arm, ITT, mITT, PP) |>
  distinct() |>
  filter(randomized, mITT) |>
  left_join(
    metadata(mae)$exposures |> select(pid, visit_code, study_product),
    by = "pid"
  ) |>
  left_join(
    metadata(mae)$visits_crfs_merged |> as_tibble() |> select(pid, visit_code, study_day, insert_study_product_time),
    by = (c("pid", "visit_code"))) |>
  dplyr::rename(
    study_day_product = study_day,
    visit_code_product = visit_code
  ) 

# study_day visit 2 in person
# ! pid 068100026 appear 2 times ! -> removed using filter(!is.na(study_day_visit2))
visit2_studyday <- 
  metadata(mae)$visits_crfs_merged |> 
  as_tibble() |> 
  select(pid, visit_code, study_day) |> 
  distinct() |> 
  filter(visit_code == "1200") |> 
  dplyr::rename(study_day_v2 = study_day) |> 
  dplyr::filter(!is.na(study_day_v2))

  
study_prod_time <- 
  study_product |> 
  left_join(visit2_studyday |> select(-visit_code), by = "pid") |> 
  filter(arm != "Pl") |> 
  group_by(pid) |> 
  filter(study_product == "1") |> 
  filter(study_day_product <= study_day_v2) |>  # only consider doses taken before v2
  filter(!((study_day_product == study_day_v2) & (insert_study_product_time %in% c("Late evening (9pm-midnight)", "Evening (6pm-9pm)")))) |> 
  arrange(study_day_product) |> 
  slice_tail(n = 1) |>              
  ungroup() 

# create category 
time_since_last_dose_week2 <- 
  study_prod_time |> 
  mutate(
    last_dose = 
      case_when(
        study_day_product == study_day_v2 ~ "same day", 
        study_day_product == study_day_v2-1 ~ "day before", 
        study_day_product < study_day_v2-1~ "more than one day"
      ), 
    last_dose_at_W2 = factor(last_dose, levels = c("same day", "day before", "more than one day"))
  ) |> 
  select(pid, last_dose_at_W2)


mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(time_since_last_dose_week2, by = join_by(pid))

rm(study_product, visit2_studyday, study_prod_time, time_since_last_dose_week2)

```

```{r}
#| eval: false

mae@metadata$calc_vars_pid |> 
  left_join(
    mae@colData |> as_tibble() |> select(pid, arm, mITT) |> distinct(),
    by = join_by(pid)
  ) |>
  filter(mITT) |> 
  dplyr::count(arm, last_dose_at_W2) |> 
  gt::gt()

```


## Bleeding


We explore the self-reported vaginal bleeding variable by first performing an exploratory analysis for both daily and weekly responses, checking for missing values.

We have available the following informations:

- From the daily CRF (CRF33):
  
  - `vaginal_bleeding.crf33`: daily self-reported vaginal bleeding in the past 24 hours (between visits 1000 and 1500), excluding weekly visit codes. Participants were asked the question: "in the past 24 hours, have you experienced any vaginal bleeding?". Possible responses were categorized into four levels:

    - `No - None`: no vaginal bleeding experienced

    - `Yes – Light`: light vaginal bleeding

    - `Yes – Medium`: moderate vaginal bleeding
    
    - `Yes – Heavy`: heavy vaginal bleeding
  
- From the weekly CRF (CRF19):

  - `vaginal_bleeding.crf19`: weekly self-reported vaginal bleeding. Participants were asked the question: "today, are you having any vaginal bleeding?". Possible responses were categorized into four levels:

    - `No-None`: no vaginal bleeding experienced

    - `Yes–Light`: light vaginal bleeding

    - `Yes–Medium`: moderate vaginal bleeding
    
    - `Yes–Heavy`: heavy vaginal bleeding


All those data are available in the `visits_crfs_merged` table stored in the metadata of the `mae` object.

```{r}

bleeding <- 
  visits_crfs_merged |> 
  select(uid, pid, visit_code, study_day, visit_type, site, randomized, vaginal_bleeding.crf33, vaginal_bleeding.crf19)

```


### Daily responses

We first look at the daily responses from CRF33.

```{r}
#| fig.height: 8
#| fig.width: 8

bleeding |> 
  filter(!is.na(vaginal_bleeding.crf33)) |>
  filter(randomized) |>
  ggplot() +
  aes(
    x = visit_code |> factor(), y = pid, 
    col = vaginal_bleeding.crf33
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Vaginal bleeding", 
    values = 
      c("Blank" = "gray80", "No - None" = "transparent", "Yes - Light" = "pink", "Yes - Medium" = "red", "Yes - Heavy" = "red4")) +
  scale_y_discrete(drop = FALSE) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Visit code", y = "Participants"
  ) 

```

All `Blank` responses are participants who did not report any vaginal bleeding in the past 24 hours. We assume (for now) that no vaginal bleeding occurred at those visits.

### Weekly response

Now, we look at the variable `vaginal_bleeding` from CRF19.

```{r}
#| fig.height: 8
#| fig.width: 5

bleeding |> 
  filter(!is.na(vaginal_bleeding.crf19)) |> 
  filter(randomized) |>
  ggplot() +
  aes(
    x = visit_code |> factor(), y = pid, 
    col = vaginal_bleeding.crf19
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Vaginal bleeding", 
    values = 
      c("Blank" = "gray80", "No-None" = "transparent", "Yes-Light" = "pink", "Yes-Medium" = "red", "Yes-Heavy" = "red4")) +
  scale_y_discrete(drop = FALSE) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Visit code", y = "Participants"
  ) 

```

From the plots above, we can see that the categorical variable `vaginal_bleeding` is not encoded in the same way in `CRF33` and `CRF19`. The correction will be carried out in the next section.

### Harmonization of daily and weekly reponses

In this section, we merge the information for `vaginal_bleeding` from the daily CRF33 and the weekly CRF19 into one single variable `vaginal_bleeding`.

```{r}
# merge daily and weekly
# harmonize variable names
# replace "Blank" by "No-None"

bleeding <- 
  bleeding |>
  mutate(
    vaginal_bleeding = coalesce(vaginal_bleeding.crf33, vaginal_bleeding.crf19),
    vaginal_bleeding = case_when(
      vaginal_bleeding %in% c("Blank") ~ "None",
      vaginal_bleeding == "No - None" | vaginal_bleeding == "No-None" ~ "None",
      vaginal_bleeding == "Yes - Light" | vaginal_bleeding == "Yes-Light" ~ "Light",
      vaginal_bleeding == "Yes - Medium" | vaginal_bleeding == "Yes-Medium" ~ "Medium",
      vaginal_bleeding == "Yes - Heavy" | vaginal_bleeding == "Yes-Heavy" ~ "Heavy",
      TRUE ~ vaginal_bleeding
    )
  )

```

The merged vaginal bleeding plot is displayed below.

```{r}
#| fig.height: 8
#| fig.width: 8

bleeding |>  
  filter(!is.na(vaginal_bleeding)) |> 
  filter(randomized) |>
  ggplot() +
  aes(
    x = visit_code |> factor(), y = pid, 
    col = vaginal_bleeding
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Vaginal bleeding", 
    values = 
      c("Blank" = "gray80", "None" = "transparent", "Light" = "pink", "Medium" = "red", "Heavy" = "red4")) +
  scale_y_discrete(drop = FALSE) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Visit code", y = "Participants"
  ) 

```



### Concatenation with study days

Multiple visit codes can correspond to the same study day when a clinical visit and a daily diary entry occur on the same calendar day. In such cases, duplicate `(pid, study_day)` pairs arise and must be resolved before analysis. For vaginal bleeding, we retain the most severe value reported across duplicate entries, following the ordinal scale: Blank < None < Light < Medium < Heavy. This conservative approach ensures that bleeding episodes are not underestimated.


We first inspect the distribution of duplicate study days across visit codes to identify the extent of the overlap.

```{r}
#| fig-height: 10
#| fig-width: 10

bleeding |>
  filter(randomized) |>
  ggplot() +
  aes(x = visit_code, y = pid, fill = vaginal_bleeding) +
  geom_tile() +
  geom_text(aes(label = study_day,
                col = ifelse(vaginal_bleeding == "Heavy", "white", "black")), size = 2.5) +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual("Vaginal bleeding", values =  c("None" = "grey90", "Light" = "pink", "Medium" = "red", "Heavy" = "red4")) +
  scale_color_identity() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    labs(x = "Visit code", y = "Participants")
  )

```

We identified `r bleeding |> dplyr::count(pid, study_day) |> filter(n > 1) |> nrow()` duplicated `(pid, study_day)` pairs involving `r bleeding |> dplyr::count(pid, study_day) |> filter(n > 1) |> distinct(pid) |> nrow()` participants. Among these, `r bleeding |> dplyr::count(pid, study_day) |> filter(n > 1) |> left_join(bleeding, by = c("pid", "study_day")) |> group_by(pid, study_day) |> filter(n_distinct(vaginal_bleeding, na.rm = TRUE) > 1) |> ungroup() |> distinct(pid, study_day) |> nrow()` pairs show inconsistent vaginal bleeding values across duplicate entries.

```{r}
#| eval: false

bleeding |>
  dplyr::count(pid, study_day) |>
  filter(n > 1) |>
  left_join(bleeding, by = c("pid", "study_day")) |>
  group_by(pid, study_day) |>
  filter(n_distinct(vaginal_bleeding, na.rm = TRUE) > 1) |>
  ungroup() |>
  select(pid, study_day, visit_code, visit_type, vaginal_bleeding) |>
  arrange(pid, study_day) |>
  gt(caption = "Duplicated (pid, study_day) with inconsistent vaginal bleeding values")

```

We notice that for several values, there is missing vaginal bleeding but a corresponding study day exists, so for these we will assign the value Blank for the vaginal bleeding.

We will create a confusion matrix to evaluate, when there are duplicates in `study_day`, to check if the values are mostly the same or different.

```{r}

bleeding <- 
  bleeding |>
  group_by(pid, study_day) |>
  filter(n() > 1) |>
  ungroup()

vb_pairs <- 
  bleeding |>
  group_by(pid, study_day) |>
  mutate(dup_id = row_number()) |>
  ungroup() |>
  select(pid, study_day, dup_id, vaginal_bleeding) |>
  tidyr::pivot_wider(
    names_from  = dup_id,
    values_from = vaginal_bleeding,
    names_prefix = "vb_"
  )

vb_counts <- vb_pairs |>
  dplyr::count(vb_1, vb_2)

vb_counts |>
  ggplot() +
  aes(x = vb_1, y = vb_2) +
  geom_tile(aes(fill = n)) +
  geom_text(aes(label = n)) +
  scale_fill_gradient(low = "white", high = "steelblue1") +
  labs(x = "Vaginal bleeding", y = "Vaginal bleeding")

```

Now to keep only one unique value of study day, we will always take the heaviest bleeding, i.e. we always take the highest value between the two: Blank < None < Light < Medium < Heavy.

Only participant 068200247 has NA values for both vaginal bleeding duplicates on the same study day. Since the corresponding `study_day` value is −11 and the analysis starts at day 0, this observation was added to the excluded values.

```{r}

bleeding <- 
  bleeding |>
  group_by(pid, study_day) |>
  mutate(
    vaginal_bleeding = case_when(
      
      "Heavy" %in% vaginal_bleeding ~ ifelse(vaginal_bleeding == "Heavy", "Heavy", "Excluded"),
      "Medium" %in% vaginal_bleeding ~ ifelse(vaginal_bleeding == "Medium", "Medium", "Excluded"),
      "Light" %in% vaginal_bleeding ~ ifelse(vaginal_bleeding == "Light", "Light", "Excluded"),
      "None" %in% vaginal_bleeding ~ ifelse(vaginal_bleeding == "None", "None", "Excluded"),
      "Blank" %in% vaginal_bleeding ~ ifelse(vaginal_bleeding == "Blank", "Blank", "Excluded"),
      
      TRUE ~ NA_character_
    ), 
    vaginal_bleeding = if_else(
      duplicated(vaginal_bleeding) & !is.na(vaginal_bleeding), "Excluded", vaginal_bleeding
    )
  ) |>
  ungroup()

```


```{r}
#| fig-height: 10
#| fig-width: 10

bleeding |>
  select(pid, study_day, visit_code, vaginal_bleeding) |>
  ggplot() +
  aes(x = visit_code, y = pid, fill = vaginal_bleeding) +
  geom_tile() +
  geom_text(aes(label = study_day,
                col = ifelse(vaginal_bleeding == "Heavy", "white", "black")), size = 2) +
  scale_fill_manual("Vaginal bleeding", 
                    values =  c("Excluded" = "yellow", "None" = "grey90", "Light" = "pink", "Medium" = "red", "Heavy" = "red4")) +
  scale_color_identity() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    labs(x = "Visit code", y = "Participants")
  )

```

We delete the rows of the duplicates (flagged by "Excluded"), keeping only the highest bleeding values.

```{r}

bleeding <- 
  bleeding |>
  filter(vaginal_bleeding != "Excluded")

```

We add the summary variables for sexual activity and semen exposure to the `calc_vars_visits` table in the metadata of the `mae` object.

```{r}
mae@metadata$calc_vars_visits <- 
  mae@metadata$calc_vars_visits |> 
  left_join(
    bleeding |> 
      select(pid, study_day, vaginal_bleeding),
    by = join_by(pid, study_day)
  )
```

## Vaginal insertions


We explore the self-reported vaginal insertions variable by first performing an exploratory analysis for both daily and weekly responses, checking for missing values.

We have available the following information:

- From the daily CRF (CRF33):

  - `put_in_vagina_type`: daily self-reported vaginal insertions in the past 24 hours (between visits 1000 and 1500), excluding weekly visit codes. Participants were asked the question: "In the past 24 hours, have you put anything inside your vagina?". 
 
  - `N-put_in_vagina`: variable summarizing the number of vaginal insertions, excluding "Nothing" when participant reported having inserted nothing in their vagina in the past 24 hours. 


- From the weekly CRF (CRF21): 

  - `put_in_vagina_type`: weekly self-reported vaginal insertions in the past 24 hours (between visits 1000 and 1500). Participants were asked the question: "Did you put anything in your vagina in the past 24 hours?".

> TODO: to complete texts

```{r}

vaginal_insertions <- 
  visits_crfs_merged |>
  select(uid, pid, visit_code, study_day, visit_type, site, randomized, n_put_in_vagina, put_in_vagina_type.crf21)

```


### Daily responses

We first look at the daily responses from CRF33.

```{r}
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8

vaginal_insertions |> 
  filter(!is.na(n_put_in_vagina)) |>
  filter(randomized) |>
  ggplot() +
  aes(
    x = visit_code |> factor(), y = pid, 
    col = n_put_in_vagina |> factor()
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Number of vaginal insertions", 
    values = 
      c("0" = "transparent", "1" = "black", "2" = "#E69F00", "3" = "#56B4E9")) +
  scale_y_discrete(drop = FALSE) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Visit code", y = "Participants"
  ) 

```

### Weekly responses

We also look at the weekly responses from CRF21. We can see that for this CRF, the variable `n_put_in_vagina` is not available. Thus, we will first create this variable.

```{r}
#| message: false
#| warning: false

vaginal_insertions <-
  vaginal_insertions |>
  mutate(
    put_in_vagina_type.crf21 = put_in_vagina_type.crf21 |>
      str_replace_all("none_of_the_above", "nothing") |>
      str_replace_all("put_other", "other") |>
      str_replace_all("_", " ") |>
      str_to_sentence(),
    
    n_put_in_vagina.crf21 = case_when(
      is.na(put_in_vagina_type.crf21) ~ NA_integer_,
      put_in_vagina_type.crf21 == "" ~ 0L,
      TRUE ~ put_in_vagina_type.crf21 |>
        str_split(",\\s*") |>
        map_int(~ sum(!str_to_lower(.x) %in% "nothing"))
    )
  ) |>
  dplyr::rename(n_put_in_vagina.crf33 = n_put_in_vagina)

```

The variable `n_put_in_vagina.crf22` is created by counting the number of inserted items separated by a comma, excluding the item `Nothing`.

```{r}
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 6

vaginal_insertions |> 
  filter(randomized) |>
  filter(!is.na(n_put_in_vagina.crf21)) |>
  ggplot() +
  aes(
    x = visit_code |> factor(), y = pid, 
    col = n_put_in_vagina.crf21 |> factor()
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Number of vaginal insertions",
    values =
      c("0" = "transparent", "1" = "black", "2" = "#E69F00", "3" = "#56B4E9")) +
  scale_y_discrete(drop = FALSE) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Visit code", y = "Participants"
  ) 

```

### Harmonization of weekly and daily responses

In this section, we merge the information for `n_put_in_vagina` from the daily CRF33 and the weekly CRF21 into one single variable `n_put_in_vagina`.

```{r}

vaginal_insertions <- 
  vaginal_insertions |>
  mutate(
    n_put_in_vagina = coalesce(n_put_in_vagina.crf33, n_put_in_vagina.crf21)
  )

```


```{r}
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 8

vaginal_insertions |> 
  filter(!is.na(n_put_in_vagina)) |>
  filter(randomized) |>
  ggplot() +
  aes(
    x = visit_code |> factor(), y = pid, 
    col = n_put_in_vagina |> factor()
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Number of vaginal insertions",
    values =
      c("0" = "transparent", "1" = "black", "2" = "#E69F00", "3" = "#56B4E9")) +
  scale_y_discrete(drop = FALSE) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Visit code", y = "Participants"
  ) 


```

There are some missing values in `n_put_in_vagina`, we replace them with 0.

```{r}

vaginal_insertions <- 
  vaginal_insertions |>
  mutate(
    n_put_in_vagina = case_when(
      as.character(n_put_in_vagina) %in% c("", NA) ~ "0",
      TRUE ~ as.character(n_put_in_vagina)
    )
  )

```

### Deleting duplicate values

As for the previous variable, some values are duplicated. We will remove these duplicates by keeping only the highest value for each participant on each study day.

```{r}
#| message: false
#| warning: false
#| fig-height: 10
#| fig-width: 13

vaginal_insertions |> 
  filter(!is.na(n_put_in_vagina)) |>
  filter(randomized) |>
  arrange(pid, study_day) |>
  group_by(pid, study_day) |>
  mutate(n_duplicates = n()) |>
  ggplot() +
  aes(
    x = study_day, y = pid, 
    col = n_duplicates |> as.character()
  ) +
  geom_point() +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Number of values per study_day",
    values =
      c("0" = "transparent", "1" = "#bdcedb", "2" = "green4", "3" = "orange")) +
  scale_y_discrete(drop = FALSE) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Study day", y = "Participants"
  ) 

```

The graph shows that most duplicates come from weekly samples, as expected from the study protocol. We'll first plot confusion matrices to see if these duplicates mostly agree with each other.

```{r}

duplicates <- 
  vaginal_insertions |> 
  select(pid, study_day, n_put_in_vagina) |>
  arrange(pid, study_day) |>
  group_by(pid, study_day) |>
  filter(n()==2) |>
  mutate(duplicate_nb = row_number()) |>
  ungroup() |>
  pivot_wider(
    names_from = duplicate_nb, 
    values_from = n_put_in_vagina, 
    names_prefix = "n_put_in_vagina_") |>
  select(n_put_in_vagina_1, n_put_in_vagina_2) 

tab <- table(duplicates$n_put_in_vagina_1, duplicates$n_put_in_vagina_2)

tab_df <- as.data.frame(tab)
names(tab_df) <- c("n_put_in_vagina_1", "n_put_in_vagina_2", "freq")

ggplot(tab_df, aes(x = n_put_in_vagina_1, y = n_put_in_vagina_2, fill = freq)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = freq), color = "black", size = 5) +
  scale_fill_gradient(low = "#f7f7f7", high = "#2166ac", name = "count") +
  labs(
    x = "n_put_in_vagina 1",
    y = "n_put_in_vagina 2"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text = element_text(size = 10),
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

We see that most duplicate values agreed with each other.

We’ll now remove duplicates using `slice_max()` (so we keep the highest `n_put_in_vagina` values).

```{r}
vaginal_insertions <- 
  vaginal_insertions |>
  select(uid, pid, study_day, visit_type, n_put_in_vagina) |>
  group_by(pid, study_day) |>
  slice_max(order_by = n_put_in_vagina, with_ties = FALSE)
```

We add the summary variables for sexual activity and semen exposure to the `calc_vars_visits` table in the metadata of the `mae` object.

```{r}
mae@metadata$calc_vars_visits <- 
  mae@metadata$calc_vars_visits |> 
  left_join(
    vaginal_insertions |> 
      select(pid, study_day, n_put_in_vagina),
    by = join_by(pid, study_day)
  )
```

## Supplementary antibiotics 

Antibiotic use was captured through two CRFs. CRF31 was completed at each clinical visit and recorded any concomitant medications reported by the participant. CRF20, part of the "Follow-up behaviour questionnaire", asked participants whether they had used any of the following in the past week: oral probiotic, vaginal probiotic, oral antibiotic, vaginal antibiotic, oral yeast medication, vaginal yeast medication, lubricant during sex, lubricant during masturbation or vibrator use, or none of the above.

### CRF 31 - Concomitant medications

CRF31 is not in the `visits_crfs_merged` or `participant_crfs_merged` datasets but in the `crf_data_clean` object.

```{r}


crf31 <- 
  metadata(mae)$crf_data_clean$crf31 |> 
  left_join(
    colData(mae) |> 
      as_tibble() |> 
      select(pid, ITT, mITT, PP, randomized, site) |> 
      distinct()
    ) |> 
  filter(randomized)

crf31 |> 
  dplyr::count(medication) |> 
  gt::gt(caption = "Concominant medication")

```

Some of those medications are antibiotics, which may have an impact on the vaginal microbiome and thus on the primary outcome. We look at the number of participants who took these antibiotics and when they took them. 


```{r}

antibiotics <- c("METRONIDAZOLE", "FLAGYL", "NITROFURANTOIN", "MACROBID", "AZITHROMYCIN", "PENICILLIN", "AMOXICILLIN", "AUGMENTIN", "CIPROBAY", "CIPROFLOXACIN", "CEFTRIAXONE", "DOXYCYCLINE","FOSFOMYCIN", "BENZANINE", "URIZONE")

# crf31 |> 
#   filter(str_detect(medication, paste(antibiotics, collapse = "|"))) |> 
#   dplyr::count(medication) |> 
#   gt::gt(caption="Supplementary antibiotics")


crf31 |> 
  filter(str_detect(medication, paste(antibiotics, collapse = "|"))) |>
  summarise(
    n = n(),
    n_PP = sum(PP, na.rm = TRUE),
    .by = medication
  ) |>
  arrange(desc(n)) |> 
  gt::gt(caption = "Supplementary antibiotics")

```

*Note : stars represent participants not included in the PP population*

```{r}

suppl_antibiotics_crf31 <- 
  crf31 |> 
  filter(str_detect(medication, paste(antibiotics, collapse = "|"))) |>
    mutate(
    antibiotics_cat = case_when(
      str_detect(medication, regex("METRONIDAZOLE|FLAGYL", ignore_case = TRUE)) ~ "Metronidazole",
      str_detect(medication, regex("NITROFURANTOIN|MACROBID", ignore_case = TRUE)) ~ "Nitrofurans",
      str_detect(medication, regex("AZITHRO", ignore_case = TRUE)) ~ "Macrolides",
      str_detect(medication, regex("PENICILLIN|AMOXICILLIN|AUGMENTIN", ignore_case = TRUE)) ~ "Penicillins",
      str_detect(medication, regex("CIPROFLOXACIN|CIPROBAY", ignore_case = TRUE)) ~ "Fluoroquinolones",
      str_detect(medication, regex("CEFTRIAXONE", ignore_case = TRUE)) ~ "Cephalosporins",
      str_detect(medication, regex("DOXYCYCLINE", ignore_case = TRUE)) ~ "Tetracyclines",
      str_detect(medication, regex("FOSFOMYCIN", ignore_case = TRUE)) ~ "Fosfomycin",
      str_detect(medication, regex("BENZANINE", ignore_case = TRUE)) ~ "Benzanines",
      str_detect(medication, regex("URIZONE", ignore_case = TRUE)) ~ "Urizone",
      TRUE ~ "Other_or_non_abx"
    ), 
    med_stop_relative_date = ifelse(is.na(med_stop_relative_date), med_start_relative_date, med_stop_relative_date), 
    pid_label = ifelse(!PP, paste("(*) ", pid), pid)
  ) |> 
    select(pid, medication, indication, med_start_relative_date, med_stop_relative_date, antibiotics_cat, PP, pid_label, site)

```

```{r}
#| fig.height: 8
#| fig.width: 8

suppl_antibiotics_crf31 |> 
  ggplot() +
  aes(
    x    = med_start_relative_date,
    xend = med_stop_relative_date,
    y    = pid_label,
    color = antibiotics_cat
  ) +
    annotate(
    "rect",
    xmin = -Inf, xmax = 35,
    ymin = -Inf, ymax = Inf,
    fill = "pink", alpha = 0.2
  ) + 
  geom_segment(linewidth = 1.2, lineend = "round") +
  geom_point(
    data = suppl_antibiotics_crf31 |> 
      filter(med_start_relative_date == med_stop_relative_date),
    aes(x = med_start_relative_date),
    size = 1
  ) +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  labs(
    x = "Study day",
    y = "Participant",
    color = "Antibiotic",
    title = "Periods of additional antibiotic use", 
    caption = "(*) Randomized participants not included in the per-protocol population"
  ) +
  scale_x_continuous(breaks = seq(0,100, by=7))

```

We have to re-structure the data to have one row per participant and per study day, with a variable indicating which supplementary antibiotics were taken on that day.If a participant took multiple supplementary antibiotics on the same day, we will concatenate the names of these antibiotics in the same cell, separated by a ";".


```{r}

suppl_antibiotics <- 
  suppl_antibiotics_crf31 |> 
  #filter(med_start_relative_date <= 35) |> 
  select(pid, med_start_relative_date, med_stop_relative_date, antibiotics_cat) |> 
  rowwise() |> 
  mutate(study_day = list(med_start_relative_date:med_stop_relative_date)) |> 
  unnest(study_day) |> 
  select(pid, study_day, supplementary_antibiotics = antibiotics_cat) |> 
  ungroup() 

suppl_antibiotics <- 
  suppl_antibiotics |> 
  group_by(pid, study_day) |>
  summarise(
    supplementary_antibiotics = paste(supplementary_antibiotics, collapse = "; "),
    .groups = "drop"
  )

```

### CRF 20 - Antibiotics

In the CRF20 it was ask wheter the participants took antibiotics (vaginal or oral) or any probiotics (vaginal or oral) or something for yeast, lubricant, etc. during the past week. This CRF was fill every clinical visit (from V1 to V9). 

For this data, we know if a participants took antibiotics during the past week but we don't know the exact day of antibiotic use.

One participant (068200457) have a empty value ("", and not `NA`). We will consider this value as missing.

```{r}

crf20 <- 
  metadata(mae)$visits_crfs_merged |> 
  left_join(
    colData(mae) |> 
      as_tibble() |> 
      select(pid, ITT, mITT, PP, randomized, site) |> 
      distinct()
    ) |> 
  filter(randomized) |>
  select(pid, visit_code, study_day, PP, site, used_during_past_week) 

crf20 |> 
  dplyr::count(used_during_past_week) |> 
  gt::gt(caption = "Antibiotics/probiotics use during the past week, by visit")

```

We only look at the vaginal/oral antibiotics, yeast medication and probiotics use during the past week.

```{r}

suppl_antibiotics_crf20 <- 
  crf20 |>
  filter(str_detect(used_during_past_week, "antibiotic|yeast|probiotic")) |> 
  mutate(
    used_during_past_week = str_remove(used_during_past_week, "lube_with_sex, "), 
    pid_label = ifelse(!PP, paste("(*) ", pid), pid)
  )
 
suppl_antibiotics_crf20 |> 
  dplyr::count(used_during_past_week) |> 
  gt::gt(caption = "Antibiotics/probiotics use during the past week, by visit")

suppl_antibiotics_crf20 |> 
  mutate(
    used_during_past_week = used_during_past_week |> factor(
      levels = c("oral_antibiotic", "vaginal_antibiotic", "oral_yeast", "vaginal_yeast", "oral_antibiotic, oral_yeast", "oral_probiotic", "vaginal_probiotic")
    )
  ) |> 
  ggplot() +
  aes(x = study_day, y = pid_label, color = used_during_past_week) +
  geom_point() +
  annotate(
    "rect",
    xmin = -Inf, xmax = 35,
    ymin = -Inf, ymax = Inf,
    fill = "pink", alpha = 0.2
  ) +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Medication use during the past week",
    values = c(
      "oral_antibiotic" = "deeppink", 
      "vaginal_antibiotic" = "deeppink3", 
      "oral_yeast" = "orange", 
      "vaginal_yeast" = "orange4", 
      "oral_antibiotic, oral_yeast" = "purple", 
      "oral_probiotic" = "lightblue",
      "vaginal_probiotic" = "lightblue4"
      )
  ) +
   scale_y_discrete(drop = FALSE) +
  labs(
    x = "Study day",
    y = "Participant",
    color = "Medication use during the past week",
    title = "Medication use during the past week (CRF20)", 
    caption = "(*) Randomized participants not included in the per-protocol population"
  ) +
  scale_x_continuous(breaks = seq(0,100, by=7))

```

We want to see if some supplementary antibiotics were not reported in CRF31 but were reported in CRF20. We look at a window of 7 days before the study day of the visit where CRF20 was filled. The following table shows the antibiotics use reported in CRF20 and CRF31 for each participant, for each visit of CRF20. The rows are colored in pink when there is no overlap between the two CRFs, i.e. when the participant reported taking antibiotics during the past week in CRF20 but did not report any antibiotic use in CRF31 during the corresponding 7-day window, or vice versa. The rows are colored in darker pink when there is no overlap and when the visit of CRF20 occurred before day 35. This allows us to identify potential discrepancies between the two CRFs that may have an impact on the primary outcome, which is assessed at day 35.

```{r}

suppl_antibiotics_crf20 <- 
  suppl_antibiotics_crf20 |> 
  mutate(
    window_start = study_day - 7,
    window_end   = study_day
  )

suppl_antibiotics_crf20_crf31 <-
  suppl_antibiotics_crf20 |> 
  select(pid, crf20_weekly_visit = visit_code, crf20_start = window_start, crf20_end = window_end, antibiotics_crf20 = used_during_past_week, PP, pid_label) |>
  left_join(
    suppl_antibiotics_crf31 |> 
      select(pid, crf31_start = med_start_relative_date, crf31_end = med_stop_relative_date, antibiotics_crf31 = antibiotics_cat),
    by = "pid",
    relationship = "many-to-many"
  ) |> 
  mutate(
    overlap = !is.na(crf31_start) & crf31_start <= crf20_end & crf31_end >= crf20_start
  ) 

suppl_antibiotics_crf20_crf31 |>
  select(pid_label, crf20_weekly_visit, crf20_start, crf20_end, antibiotics_crf20, crf31_start, crf31_end, antibiotics_crf31, PP, overlap) |>
  arrange(overlap, is.na(crf31_start), PP) |>  
  gt() |>
  tab_style(
    style = 
      cell_fill(
        color = "coral1", alpha = 0.1),
        locations = cells_body(rows = !overlap)
  ) |> 
  tab_style(
    style = cell_fill(color = "coral1", alpha = 0.3),
    locations = cells_body(rows = crf20_start <= 35)
  ) |> 
    tab_style(
      style = cell_fill(color = "coral1", alpha = 0.7),
      locations = cells_body(rows = !overlap & PP & crf20_start <= 35)
    )

```


We merged the add the information of CRF20 in the `suppl_antibiotics` dataset, incorporing new rows for the visits where antibiotics were reported in CRF20 but not in CRF31. We also created a variable `antibiotics_source` to indicate whether the information on supplementary antibiotics came from CRF20, CRF31, or both.

```{r}

suppl_antibiotics_crf20 <- 
  suppl_antibiotics_crf20 |>
  # Separate multiple entries ("oral_antibiotic, oral_yeast")
  separate_rows(used_during_past_week, sep = ",\\s*") |>
  mutate(
    is_antibiotic = str_detect(used_during_past_week, "antibiotic")
  ) |>
  rowwise() |>
  mutate(study_day = list(window_start:window_end)) |>
  unnest(study_day) |>
  select(pid, study_day, used_during_past_week, is_antibiotic) |>
  ungroup()

suppl_antibiotics_crf20_to_add <-
  suppl_antibiotics_crf20 |>
  left_join(
    suppl_antibiotics |> select(pid, study_day, crf31_med = supplementary_antibiotics),
    by = c("pid", "study_day")
  ) |>
  mutate(
    already_in_crf31 = !is.na(crf31_med) & is_antibiotic,
    # Keep if: non-antibiotic (always) OR antibiotic not in CRF31
    keep = !already_in_crf31
  ) |>
  filter(keep) |>
  select(pid, study_day, supplementary_antibiotics = used_during_past_week)

# combined dataset of supplementary antibiotics from both CRF31 and CRF20
suppl_antibiotics <- 
  suppl_antibiotics |>
  mutate(source = "CRF31") |>
  bind_rows(
    suppl_antibiotics_crf20_to_add |> mutate(source = "CRF20")
  ) |>
  group_by(pid, study_day) |>
  summarise(
    supplementary_antibiotics = paste(supplementary_antibiotics, collapse = "; "),
    source = paste(unique(source), collapse = " & "),
    .groups = "drop"
  )

```

```{r}

tmp <- 
  suppl_antibiotics |>
  separate_rows(supplementary_antibiotics, sep = ";\\s*") |>
  arrange(pid, supplementary_antibiotics, study_day) |>
  group_by(pid, supplementary_antibiotics, source) |>
  mutate(
    gap = study_day - lag(study_day, default = dplyr::first(study_day) - 1) > 1,
    segment_id = cumsum(gap)
  ) |>
  group_by(pid, supplementary_antibiotics, source, segment_id) |>
  summarise(
    start = min(study_day),
    end   = max(study_day),
    .groups = "drop"
  ) |>
  left_join(
    visits_crfs_merged |> select(pid, site, PP) |> distinct(),
    by = "pid"
  ) |> 
  mutate(
    pid_label = ifelse(!PP, paste("(*) ", pid), pid)
  )

# Graphe
tmp |>
  ggplot() +
  aes(
    x     = start,
    xend  = end,
    y     = pid_label,
    color = supplementary_antibiotics,
    linetype = source
  ) +
  annotate(
    "rect",
    xmin = -Inf, xmax = 35,
    ymin = -Inf, ymax = Inf,
    fill = "pink", alpha = 0.2
  ) +
  geom_segment(linewidth = 1.2, lineend = "round") +
  geom_point(
    data = tmp |> filter(start == end),
    aes(x = start),
    size = 1
  ) +
  facet_grid(site ~ ., scales = "free_y", space = "free_y") +
  scale_linetype_manual(
    values = c(
      "CRF31"          = "solid",
      "CRF20"          = "dotted",
      "CRF31 & CRF20"  = "solid"
    )
  ) +
  labs(
    x        = "Study day",
    y        = "Participant",
    color    = "Medication",
    linetype = "Source",
    title    = "Periods of supplementary medication use",
    caption  = "(*) Randomized participants not included in the per-protocol population"
  ) +
  scale_x_continuous(breaks = seq(0, 100, by = 7))

```

We add the summary variables for sexual activity and semen exposure to the `calc_vars_visits` table in the metadata of the `mae` object.

```{r}
mae@metadata$calc_vars_visits <- 
  mae@metadata$calc_vars_visits |> 
  left_join(
    suppl_antibiotics |> 
      select(pid, study_day, supplementary_antibiotics, source_supplementary_antibiotics = source),
    by = join_by(pid, study_day)
  )
```


# Saving the `MultiAssayExperiment` objects

We save the MAE for downstream analyses.

```{r}
#| eval: true

saveRDS(
  mae, 
  str_c(
    get_03_output_dir(),  
    "mae_full_03_", today() |> str_remove_all("-"), ".rds"
  )
)

```



