---
title: "Calculated variables"
author: Laura Symul, Laura Vermeren
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: refresh # true refresh false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| warning: false
#| cache: false

library(tidyverse)
library(magrittr)
library(gt)
library(patchwork)
library(SummarizedExperiment)
library(tidySummarizedExperiment)
library(MultiAssayExperiment)

tmp <- fs::dir_map("R scripts/", source)
tmp <- fs::dir_map("../VIBRANT-99-utils/R/", source)
rm(tmp)
theme_set(theme_light())
```



```{r}

mae_files <- 
  fs::dir_ls(
    get_02_output_dir(),
    regexp = "mae_full_.*\\.rds$"
  ) |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)

mae <- readRDS(mae_files)

rm(mae_files)

```



> TODO: edit this file (introduction, etc.)


The primary purpose of this document is to define and create additional assays to the `MultiAssayExperiment` object that contain the primary and secondary outcomes of the VIBRANT trial.


# Calculated variables

```{r}
mae@metadata$calc_vars_pid <- 
  tibble(pid = mae@colData |> as_tibble() |> pull(pid) |> unique())
```



## Education - can we harmonize between sites ?

```{r}

education <- 
  bind_rows(
    mae@metadata$crf_data_clean$crf1 |> 
      select(uid, pid, visit_code, edu),
    mae@metadata$crf_data_clean$crf101 |> 
      select(uid, pid, visit_code, edu)
  ) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, site, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

```

```{r}
#| fig-width: 8
#| fig-height: 2.5
education |> 
  ggplot() +
  aes(y = edu, fill = randomized) +
  facet_wrap(. ~ site, scales = "free") +
  geom_bar() +
  xlab("Number of participants") +
  ylab("") +
  scale_fill_manual(
    name = "",
    values = c("TRUE" = "dodgerblue1", "FALSE" = "gray70"),
    labels = c("TRUE" = "Randomized", "FALSE" = "Not randomized")
  ) +
  theme(legend.position = "top")



```

```{r}
#| fig-width: 8
#| fig-height: 2

education |> 
  filter(randomized) |> 
  ggplot() +
  aes(y = edu) +
  facet_wrap(. ~ site, scales = "free") +
  geom_bar(fill = "dodgerblue1") +
  xlab("Number of participants") +
  ylab("")
```

## Economic worry

```{r}

worry <- 
   bind_rows(
    mae@metadata$crf_data_clean$crf2 |> 
      select(uid, pid, visit_code, starts_with("worry_")),
    mae@metadata$crf_data_clean$crf102 |> 
      select(uid, pid, visit_code, starts_with("worry_"))
  ) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, site, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

```

```{r}
#| fig-height: 5
#| fig-width: 6
# create an upset plot to see the combinations of worries
library(UpSetR)
# create a binary matrix for the worries
worry_matrix_CAP <-
  worry |>
  filter(randomized, site == "CAP", !is.na(worry_food)) |>
  select(starts_with("worry_")) |>
  # remove "worry_" prefix from column names
  rename_with(~ str_remove_all(., "worry_")) |>
  mutate(across(everything(), ~ ifelse(. == "Checked", 1, 0))) |>
  as.data.frame()

upset(
  worry_matrix_CAP,
  nsets = ncol(worry_matrix_CAP),
  nintersects = NA,
  keep.order = TRUE,
  order.by = "freq",
  #text.scale = c(2, 2, 2, 1.5, 2, 2),
  main.bar.color = "dodgerblue1",
  sets.bar.color = "gray70",
  set_size.show = TRUE
  )

worry_matrix_MGH <-
  worry |>
  filter(randomized, site == "MGH", !is.na(worry_food)) |>
  select(starts_with("worry_")) |>
  # remove "worry_" prefix from column names
  rename_with(~ str_remove_all(., "worry_")) |>
  mutate(across(everything(), ~ ifelse(. == "Checked", 1, 0))) |>
  as.data.frame()

upset(
  worry_matrix_MGH,
  nsets = ncol(worry_matrix_MGH),
  nintersects = NA,
  keep.order = TRUE,
  order.by = "freq",
  #text.scale = c(2, 2, 2, 1.5, 2, 2),
  main.bar.color = "dodgerblue1",
  # sets.bar.color = "gray70",
  set_size.show = TRUE
)

```


```{r}


table1_data <- 
  metadata(mae)$participant_crfs_merged |> 
  dplyr::left_join(
    colData(mae) |> as_tibble() |> select(pid, site, arm, ITT, mITT, PP) |> distinct(), 
    by = join_by(pid, site)
    )  |> 
  filter(mITT) |> 
  select(pid, site, arm, cut_size_meals, eat_less, hungry_did_not_eat) |> 
  mutate(
    food_insecurity = ifelse(
      cut_size_meals == "Yes" | eat_less == "Yes" | hungry_did_not_eat == "Yes",
      "Yes",
      "No"
    ),
    food_insecurity = factor(food_insecurity, levels = c("No", "Yes"))
  ) |> 
  select(-c(eat_less, cut_size_meals, hungry_did_not_eat))

```

```{r}
#| fig-height: 4
#| fig-width: 8

table1_data |> 
  left_join(
    worry |> select(pid, worry_food) |> distinct(),
    by = join_by(pid)
  ) |> 
  dplyr::count(site, food_insecurity, worry_food) |> 
  ggplot() +
  aes(x = food_insecurity, y = worry_food, fill = n) +
  geom_tile() +
  geom_text(aes(label = n)) +
  facet_wrap(. ~ site, scales = "free") +
  scale_fill_gradient(low = "white", high = "dodgerblue1") 



```




## Pre-existing conditions

```{r}

pre_existing_conditions <- 
  mae@metadata$crf_data_clean$crf5 |> 
  select(pid, no_pre_existing_condition, diagnosis) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, site, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

```

```{r}
#| fig-height: 7
#| 
pre_existing_conditions |> 
  filter(randomized) |> 
  dplyr::mutate(
    diagnosis = 
      case_when(
        no_pre_existing_condition ~ "No pre-existing condition",
        TRUE ~ diagnosis
        )
    ) |> 
  select(pid, diagnosis, site) |> 
  distinct() |> 
  dplyr::count(site, diagnosis) |> 
  arrange(-n) |> 
  mutate(diagnosis = diagnosis |> fct_inorder()) |>
  ggplot() +
  aes(x = n, y = diagnosis) +
  facet_grid(. ~ site, scales = "free_x") +
  geom_col()

```
## Psychological wellbeing

```{r}
psych_wellbeing <- 
  mae@metadata$crf_data_clean$crf11 |> 
  select(pid, upset:difficulties_piling_up) |>
  distinct() |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, site, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

psych_wellbeing_long <- 
  psych_wellbeing |> 
  filter(randomized) |>
  filter(!is.na(upset)) |>
  pivot_longer(
    cols = upset:difficulties_piling_up,
    names_to = "question",
    values_to = "answer"
  )



```

```{r}
#| fig-height: 10
#| fig-width: 8
psych_wellbeing_long |> 
  group_by(site, pid) |> 
  mutate(total_score = sum(answer |> as.integer())) |>
  ungroup() |> 
  arrange(total_score) |> 
  mutate(
    pid = pid |> fct_inorder()
  ) |>
  ggplot() +
  aes(y = pid, x = question, fill = answer) +
  facet_grid(site ~ ., scales = "free_y", space = "free") +
  geom_tile() +
  scale_fill_brewer(type = "div", palette = "PRGn") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(angle = 0)
  )
```

## Sexual behavior

### PSA

```{r}

psa <- 
  mae@metadata$crf_data_clean$crf45 |> 
  select(uid, pid, visit_code, psa_results) |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, site, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

```

```{r}
#| fig-height: 8
#| fig-width: 6


psa |> 
  ggplot() +
  aes(x = visit_code, y = pid) +
  geom_tile(fill = "lightseagreen")

```

## Metronidazole use

```{r}

mtz_crf_raw <- 
  mae@metadata$crf_data_raw$crf22 |> 
  left_join(
    mae@colData |> as_tibble() |> 
      select(pid, site, randomized) |> 
      distinct(),
    by = join_by(pid)
  )

mtz_crf_raw |> 
  dplyr::count(dfseq)


mtz_crf_raw |> 
  filter(randomized) |> 
  dplyr::count(dfseq)


mtz_crf_raw |> 
  filter(randomized) |> 
  dplyr::filter(dfseq == 1200) 

mtz_crf_raw |> 
  filter(randomized) |> 
  dplyr::count(pid) |> 
  arrange(-n) |> 
  dplyr::count(n)

mtz_crf_raw |> 
  group_by(pid) |> 
  mutate(min_study_day = min(study_day)) |> 
  ungroup() |> 
  mutate(pid = pid |> fct_reorder(min_study_day)) |>
  ggplot() +
  aes(x = study_day, y = pid, col = dfseq |> as.character()) +
  geom_point()

mae@metadata$exposures |> 
  ggplot() +
  aes(x = study_day, y = pid, col = metronidazole |> factor()) + geom_point() +
  scale_color_manual(
    name = "Metronidazole use",
    na.value = "lightgray",
    values = c("0" = "red", "1" = "purple", "2" = "blue")
  ) +
  geom_point()
  

```

## Study product exposure related variables


### Number of self-reported study product doses taken


```{r}

n_self_reported_doses <- 
  metadata(mae)$exposures |>
  group_by(pid) |>
  summarize(
    n_product_doses = sum(study_product |> as.integer(), na.rm = TRUE),
    n_product_doses_PP = sum(study_product_PP, na.rm = TRUE),
    comments_studyproduct = str_c(comments_studyproduct[!is.na(comments_studyproduct)], collapse = "; "),
    .groups = "drop"
  )


mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(n_self_reported_doses, by = join_by(pid))

rm(n_self_reported_doses)

```


### Time since last self-reported study product doses at Week 2 in-clinic visit


We document the time between the last dose taken prior to the in-person visit at week 2 and the swab collected at that visit. This is relevant because the primary outcome is detection of the LBP strain at the week 2 visit, and the time since last dose may influence the likelihood of detection.

Information on the self-reported time of last study product use comes from `exposures`, while the study day of the week 2 in-person visit is taken from `visits_crfs_merged.`

We define the following categories for the timing of the last study product use:
-   "same day": last dose taken on the same day as the visit (before 6 p.m.)
-   "day before": last dose taken the day before the visit
-   "more than 1 day": last dose taken more than one day before the visit


```{r}

study_product <-
  mae@colData |>
  as_tibble() |>
  select(pid, site, randomized, arm, ITT, mITT, PP) |>
  distinct() |>
  filter(randomized, mITT) |>
  left_join(
    metadata(mae)$exposures |> select(pid, visit_code, study_product),
    by = "pid"
  ) |>
  left_join(
    metadata(mae)$visits_crfs_merged |> as_tibble() |> select(pid, visit_code, study_day, insert_study_product_time),
    by = (c("pid", "visit_code"))) |>
  dplyr::rename(
    study_day_product = study_day,
    visit_code_product = visit_code
  ) 

# study_day visit 2 in person
# ! pid 068100026 appear 2 times ! -> removed using filter(!is.na(study_day_visit2))
visit2_studyday <- 
  metadata(mae)$visits_crfs_merged |> 
  as_tibble() |> 
  select(pid, visit_code, study_day) |> 
  distinct() |> 
  filter(visit_code == "1200") |> 
  dplyr::rename(study_day_v2 = study_day) |> 
  dplyr::filter(!is.na(study_day_v2))

  
study_prod_time <- 
  study_product |> 
  left_join(visit2_studyday |> select(-visit_code), by = "pid") |> 
  filter(arm != "Pl") |> 
  group_by(pid) |> 
  filter(study_product == "1") |> 
  filter(study_day_product <= study_day_v2) |>  # only consider doses taken before v2
  filter(!((study_day_product == study_day_v2) & (insert_study_product_time %in% c("Late evening (9pm-midnight)", "Evening (6pm-9pm)")))) |> 
  arrange(study_day_product) |> 
  slice_tail(n = 1) |>              
  ungroup() 

# create category 
time_since_last_dose_week2 <- 
  study_prod_time |> 
  mutate(
    last_dose = 
      case_when(
        study_day_product == study_day_v2 ~ "same day", 
        study_day_product == study_day_v2-1 ~ "day before", 
        study_day_product < study_day_v2-1~ "more than one day"
      ), 
    last_dose_at_W2 = factor(last_dose, levels = c("same day", "day before", "more than one day"))
  ) |> 
  select(pid, last_dose_at_W2)


mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  left_join(time_since_last_dose_week2, by = join_by(pid))

rm(study_product, visit2_studyday, study_prod_time, time_since_last_dose_week2)

```

```{r}

mae@metadata$calc_vars_pid |> 
  left_join(
    mae@colData |> as_tibble() |> select(pid, arm, mITT) |> distinct(),
    by = join_by(pid)
  ) |>
  filter(mITT) |> 
  dplyr::count(arm, last_dose_at_W2) |> 
  gt::gt()

```



# Saving the `MultiAssayExperiment` objects

We save the MAE for downstream analyses.

```{r}

saveRDS(
  mae, 
  str_c(
    get_03_output_dir(),  
    "mae_full_", today() |> str_remove_all("-"), ".rds"
  )
)

```



