---
title: "MAE augmentation - Calculated variables at the participant and visit code level"
author: Mariam Rahali Masbouri, Khadija Rahali Masbouri, Laura Symul, Laura Vermeren
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: refresh # true refresh false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| warning: false
#| cache: false
#| label: setup

library(tidyverse)
library(magrittr)
library(gt)
library(patchwork)
library(grid)
library(gridExtra)
# library(factoextra)
# library(SummarizedExperiment)
# library(tidySummarizedExperiment)
# library(MultiAssayExperiment)

tmp <- fs::dir_map("R scripts/", source)
tmp <- fs::dir_map("../VIBRANT-99-utils/R/", source)
rm(tmp)
theme_set(theme_light())

```


```{r}

mae_files <- 
  fs::dir_ls(
    get_03_output_dir(),
    regexp = "mae_full_03_.*\\.rds$"
  ) |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)

mae <- readRDS(mae_files)

rm(mae_files)

```


# Calculated variables at the participant and visit code level

The goal of this document is to create calculated variables at the participant and visit code level, and add them to the metadata of the `MultiAssayExperiment` object. Two tables will be created : one table at the weekly visit code level (`calc_vars_visit_code_weekly`) and another at the daily visit code level (`calc_vars_visit_code_daily`).

> TODO : add description and list of variables created at the visit code level, and the rationale for creating them.


# Calculated variables at the weekly visit code level

```{r}

mae@metadata$calc_vars_weekly_visit_code <- 
  mae@colData |> 
  as_tibble() |> 
  select(uid, pid, visit_code, visit_type) |> 
  filter(!is.na(pid), !is.na(visit_code), visit_type == "Clinic") |>
  distinct()

```

## Days since last dose 

> based on the study product exposure data 

We calculate the number of days since last dose of the study product based on the `exposure` table constructed in 01a. We first calculate the number of days since last dose at the participant and study day which is the corresponding study product exposure variable being > 0, and then we add this variable to the `calc_vars_weekly_visit_code` table by merging by `uid`.

Since the rationale behind this variable is to assess the number of days since last dose of the study product before the weekly clinic visit, we calculate this variable at the daily visit code level, and then we will merge it to the weekly visit code level table.


```{r}

tmp <- mae@metadata$calc_vars_study_day |> select(pid, study_day) |> distinct() |> arrange(pid, study_day)

exposure <- 
  tibble(tmp) |>
  left_join(
    mae@metadata$exposures |> 
      select(pid, study_day, daily_self_reported_exposure_sp),
    by = join_by(pid, study_day)
  ) |>
  left_join(
    mae@colData |> as_tibble() |> select(uid, pid, study_day, visit_code, visit_type, site),
    by = join_by(pid, study_day)
  )

exposure <-
  exposure |>
  arrange(pid, study_day) |>
  group_by(pid) |>
  mutate(
    last_sp_dose_day = case_when(
      daily_self_reported_exposure_sp > 0 ~ study_day,
      TRUE ~ NA_integer_
    ),
    last_sp_dose_day = zoo::na.locf(last_sp_dose_day, na.rm = FALSE),
    days_since_last_sp_dose = study_day - last_sp_dose_day
  ) |>
  ungroup() |>
  select(-last_sp_dose_day)


```

We add the variable `days_since_last_sp_dose` to the `calc_vars_weekly_visit_code` table.

```{r}

mae@metadata$calc_vars_weekly_visit_code <- 
  mae@metadata$calc_vars_weekly_visit_code |> 
  left_join(
    exposure |> 
      select(uid, days_since_last_sp_dose),
    by = join_by(uid)
  )

```


## Days since last sex 

We calculate the number of days since last sexual intercourse with and without condom, and with unknown condom use, based on the semen exposure variable calculated in the `calc_vars_study_day` table. We first calculate the number of days since last sexual intercourse with and without semen exposure and with unknown condom use, at the participant and study day which is the corresponding semen exposure variable being > 0, and then we add these variables to the `calc_vars_weekly_visit_code` table by merging by `uid`.

```{r}

tmp <- mae@metadata$calc_vars_weekly_visit_code$uid |> unique() |> sort()

sex <- 
    mae@metadata$calc_vars_study_day |> 
    select(
       pid, study_day, pv_sex_w_semen_exposure, pv_sex_wo_semen_exposure, other_sex_wo_semen_exp
    )

```


```{r}

sex <- 
  sex |>
  arrange(pid, study_day) |>
  group_by(pid) |> 
  mutate(
    # with condom only
    last_sex_day_w_semen_exp = case_when(
      pv_sex_w_semen_exposure > 0 ~ study_day,
      TRUE ~ NA_integer_
    ),
    last_sex_day_w_semen_exp = zoo::na.locf(last_sex_day_w_semen_exp, na.rm = FALSE), 
    days_since_last_sex_w_semen_exp = study_day - last_sex_day_w_semen_exp,
    
    # without condom only
    last_sex_day_wo_semen_exp = case_when(
      pv_sex_wo_semen_exposure > 0 ~ study_day,
      TRUE ~ NA_integer_
    ),
    last_sex_day_wo_semen_exp = zoo::na.locf(last_sex_day_wo_semen_exp, na.rm = FALSE),
    days_since_last_sex_wo_semen_exp = study_day - last_sex_day_wo_semen_exp, 
    # unknown condom use
    last_other_sex_day = case_when(
      other_sex_wo_semen_exp > 0 ~ study_day,
      TRUE ~ NA_integer_
    ),
    last_other_sex_day = zoo::na.locf(last_other_sex_day, na.rm = FALSE),
    days_since_last_other_sex = study_day - last_other_sex_day
  ) |> 
  ungroup() |> 
  select(-c(last_sex_day_wo_semen_exp, last_sex_day_w_semen_exp, last_other_sex_day)) |>
  left_join(
    mae@colData |> as_tibble() |> select(uid, pid, study_day, visit_code, visit_type, site),
    by = join_by(pid, study_day)
  )

```

Add the variable `days_since_last_sex_wo_semen_exp`, `days_since_last_sex_w_semen_exp` and `days_since_last_other_sex` to the `calc_vars_weekly_visit_code` table.

```{r}

mae@metadata$calc_vars_weekly_visit_code <- 
  mae@metadata$calc_vars_weekly_visit_code |> 
  left_join(
    sex |> 
      select(uid, days_since_last_sex_wo_semen_exp, days_since_last_sex_w_semen_exp, days_since_last_other_sex),
    by = join_by(uid)
  )

```

# Calculated variables at the daily visit code level

> TODO

# Saving the `MultiAssayExperiment` objects

We save the MAE for downstream analyses.

```{r}
#| eval: true

saveRDS(
  mae, 
  str_c(
    get_03_output_dir(),  
    "mae_full_04_", today() |> str_remove_all("-"), ".rds"
  )
)

```

