---
title: "Post-unblinding augmentation trial data"
author: Laura Symul, Laura Vermeren, Khadija Rahali Masbouri, Mariam Rahali Masbouri
date: today
format: 
   html:
     page-layout: full
     code-fold: true
     toc: true
     toc-location: left
     toc-depth: 5
     embed-resources: true
execute:
  cache: refresh # true refresh false
  warning: false
knitr:
  opts_chunk:
    out.width: "100%"
editor: source
---

```{r}
#| warning: false
#| cache: false

library(tidyverse)
library(magrittr)
library(gt)
library(patchwork)
library(SummarizedExperiment)
library(tidySummarizedExperiment)
library(MultiAssayExperiment)

tmp <- fs::dir_map("R scripts/", source)
tmp <- fs::dir_map("../VIBRANT-99-utils/R/", source)
rm(tmp)
theme_set(theme_light())

```


The purpose of this document is to compute additional "calculated variables" that require unblinded data, and to save the updated `MultiAssayExperiment` object for downstream analyses.


# Loading the `MultiAssayExperiment` object

```{r}

mae_files <- 
  fs::dir_ls(
    get_04_output_dir(),
    regexp = "mae_full_.*\\.rds$"
  ) |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)

mae <- readRDS(mae_files)

rm(mae_files)

```


# Calculated variables

## Study product exposure related variables



### Exposure by qPCR on dosing days


After Unblinding, qPCR assays were performed on daily self-collected swabs to inform effective exposure (see `30_qPCR_daily.qmd`). 

In this section, this daily qPCR data is analysed to describe patterns of LBP detection across the dosing period and subsequent follow-up.


```{r}

df <- 
  mae[["qPCR_daily"]] |> 
  as_tibble() |> 
  select(uid, .feature, copies_per_swab_med, taxon_label, LBP, strain_origin) |> 
  left_join( 
    mae@colData |> 
      as_tibble() |> 
      select(
        uid, 
        pid, 
        visit_code, PIPV, visit, study_day, 
        site, randomized, arm, ITT, mITT, PP
      ),
    by = join_by(uid)
  ) |> 
  arrange(pid, study_day, LBP, strain_origin) |>
  mutate(.feature = .feature |> fct_inorder())


```

##### 16S total copies per swabs

First, to assess general quality of collected swabs, we visualize the qPCR 16S results across visits and participants. For each participant, the median number of 16S copies per swab (log₁₀-transformed) is displayed as a heatmap, with participants on the y-axis and visit codes on the x-axis. The plot is stratified by study arm and site using nested facets.

```{r}
#| fig-width: 10
#| fig-height: 10

df |> 
  filter(.feature == "16S", randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = copies_per_swab_med |> log10()) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free", space = "free") +
  geom_tile() +
  scale_fill_viridis_c(option = "F", na.value = "blue") +
  theme(
    axis.text = element_text(size = 6),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6)
  )

```

##### LBP strains

We generate heatmaps of qPCR results for each study arm. For every participant and visit, the median number of copies per swab for each LBP strain is shown, with strain identity on the y-axis and visit code on the x-axis.

```{r}
#| fig-width: 10
#| fig-height: 15

plots <- 
  map(
  .x = df$arm |> unique() |> sort(),
  ~ df |> 
    filter(.feature != "16S", arm == .x) |> 
    ggplot() +
    aes(x = visit_code, y = .feature, fill = LBP, alpha = copies_per_swab_med |> asinh()) +
    ggh4x::facet_nested(site + pid ~ ., scales = "free", space = "free") +
    geom_tile() +
    scale_alpha_continuous(range = c(0.1, 1), limits = c(-1/2, 20)) +
    ggtitle(.x) +
    theme(
      axis.text = element_text(size = 4),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6)
    )
)

```

```{r}
names(plots) <- df$arm |> unique() |> sort()
```


::: panel-tabset
```{r}
#| results: asis
#| fig-width: 10
#| fig-height: 15

plot_tabs(plots)
```
:::


We plot longitudinal qPCR measurements for each participant and strain, stratified by study arm, site, and participant. The x-axis shows study days and the y-axis the median copies per swab (log₁₀ scale). Each line represents one LBP strain, with color indicating strain identity. Shaded background areas indicate treatment periods: red for oral metronidazole and yellow for LBP dosing. A dashed horizontal line marks the 10⁵ copies/swab threshold.

```{r}
#| fig-width: 10
#| fig-height: 15

plots <- 
  map(
  .x = df$arm |> unique() |> sort(),
  ~ df |> 
    filter(.feature != "16S", arm == .x, study_day >= 0) |> 
    ggplot() +
    aes(x = study_day, y = copies_per_swab_med, col = LBP ) + #  |> asinh()
    ggh4x::facet_nested(site + pid ~ .) + # scales = "free", space = "free"
    geom_hline(yintercept = 10^5, lty = 2, col = "gray40") +
    geom_rect(
      data = tibble(study_day = 0, copies_per_swab_med = NA), 
      xmin = 0.5, xmax = 7.5, ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "red", col = NA
    ) +
    geom_rect(
      data = tibble(study_day = 0, copies_per_swab_med = NA), 
      xmin = 7.5, 
      xmax = case_when(
        .x == "LC106-3" ~ 10.5,
        .x == "LC106-o" ~ 10.5,
        TRUE ~ 14.5
      ), 
      ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "gold", col = NA
    ) +
    # Rectangle jaune supplémentaire pour LC106-o uniquement
    {if(.x == "LC106-o") geom_rect(
      data = tibble(study_day = 0, copies_per_swab_med = NA), 
      xmin = 3.5, xmax = 7.5, ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "gold", col = NA
    )} +
    geom_line(aes(group = .feature)) +
    geom_point() +
    scale_y_log10() +
    expand_limits(y = 10^10) +
    scale_x_continuous(breaks = seq(0, 35, by = 7)) +
    ggtitle(.x)
)
```

```{r}
names(plots) <- df$arm |> unique() |> sort()
```


::: panel-tabset
```{r}
#| results: asis
#| fig-width: 10
#| fig-height: 15

plot_tabs(plots)
```
:::


##### LBP exposure

We summarize the qPCR results to compute total LBP copies per swab and the number of detected LBP strains for each participant at each visit. We also calculate the proportion of detected LBP strains among those expected based on the study arm.

```{r}

df_summ <- 
  df |> 
  filter(.feature != "16S") |> 
  group_by(uid, pid, randomized, arm, ITT, mITT, PP, site, visit_code, PIPV, visit, study_day) |> 
  summarise(
    LC106_total = sum(copies_per_swab_med[str_detect(LBP, "LC106")], na.rm = TRUE),
    LC115_excl_total = sum(copies_per_swab_med[!str_detect(LBP, "LC106")], na.rm = TRUE),
    total_LBP = sum(copies_per_swab_med[str_detect(LBP, "LC")], na.rm = TRUE),
    total_16S = sum(copies_per_swab_med[.feature == "16S"], na.rm = TRUE),
    n_LC106 = sum(str_detect(LBP, "LC106") & (copies_per_swab_med != 0), na.rm = TRUE),
    n_LC115_excl = sum(!str_detect(LBP, "LC106") & (copies_per_swab_med != 0), na.rm = TRUE),
    n_LBP = sum(str_detect(LBP, "LC") & (copies_per_swab_med != 0), na.rm = TRUE),
    .groups = "drop"
  ) |> 
  mutate(
    n_expected = 
      case_when(
        arm == "Pl" ~ 0,
        str_detect(arm, "LC106") ~ 6,
        str_detect(arm, "LC115") ~ 15,
        TRUE ~ NA_real_
      ),
    n_detected_among_expected = 
      case_when(
        arm == "Pl" ~ n_LBP,
        str_detect(arm, "LC106") ~ n_LC106,
        str_detect(arm, "LC115") ~ n_LBP,
        TRUE ~ NA_real_
      ),
    p_detected = n_detected_among_expected/n_expected,
  )

```



```{r}
#| fig-height: 11
#| fig-width: 7
#| eval: false

df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = total_LBP |> log10()) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() +
  geom_text(aes(label = n_LBP, col = total_LBP |> log10()), size = 2) +
  scale_fill_viridis_c(option = "F") + guides(col = "none") +
  scale_color_viridis_c(option = "F", direction = -1, na.value = "white") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) +
  labs(caption = "Text = number of detected (>0) LBP strains by qPCR.")


df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = n_LBP) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() +
  scale_fill_viridis_c("") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) +
  labs(caption = "Number of detected (>0) LBP strains by qPCR.")


df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = p_detected) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() +
  # scale_fill_gradient2(low = "gray80", mid = "purple", high = "blue", midpoint = 7.5) +
  scale_fill_viridis_c("proportion of LBP strains\ndetected among expected", option = "G") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) +
  labs(caption = ".")


```


**Comparison with self-declared exposure.**

Dots indicate whether participants reported inserting the study product on each day.

```{r}

df_summ <- 
  df_summ |>
  full_join( 
    mae@colData |> 
      as_tibble() |> 
      select(
        uid, 
        pid, 
        visit_code, PIPV, visit, study_day, 
        site, randomized, arm, ITT, mITT, PP)) |>
  arrange(pid, study_day) |>
  select(-ends_with("sp")) |> 
  left_join(
    mae@metadata$exposures |> 
      select(uid, pid, visit_code, study_day, ends_with("sp"))) |>
  filter(randomized)


```

```{r}
#| fig-height: 15
#| fig-width: 10

df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = total_LBP |> log10()) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() +
  geom_point(aes(shape = daily_self_reported_adherence_sp |> factor()), size = 2, fill = "black", col = "white") +
  scale_shape_manual("self-declared\nexposure", values = c(NA, 21)) +
  scale_fill_viridis_c(option = "F") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) 


df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = p_detected) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() + # 
  geom_point(aes(shape = daily_self_reported_adherence_sp |> factor()), size = 2, fill = "black", col = "white") +
  scale_shape_manual("self-declared\nexposure", values = c(NA, 21)) +
  scale_fill_viridis_c(
    "proportion of LBP strains\ndetected among expected", option = "G"
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  )


```

Here we adjust for reported product losses, representing actual exposure (i.e., tablets that remained in place).

```{r}
#| fig-height: 15
#| fig-width: 10

df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = total_LBP |> log10()) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() +
  geom_point(aes(shape = daily_self_reported_exposure_sp |> round() |> factor()), size = 2, fill = "black", col = "white") +
  scale_shape_manual("self-declared\nexposure", values = c(NA, 21)) +
  scale_fill_viridis_c(option = "F") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  )


df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = p_detected) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() + # 
  geom_point(aes(shape = daily_self_reported_exposure_sp |> round() |> factor()), size = 2, fill = "black", col = "white") +
  scale_shape_manual("self-declared\nexposure", values = c(NA, 21)) +
  scale_fill_viridis_c(
    "proportion of LBP strains\ndetected among expected", option = "G"
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) 


```


**Comparison with applicator staining**


```{r}

tmp <- 
   metadata(mae)$participant_crfs_merged |> 
   as_tibble() |> 
   select(pid, used_applicators, number_applicators, applicator_stain_positive, applicator_stain_negative, applicator_stain_indeterminant, total_applicators_stained, proportion_positive_applicators, total) |> 
   dplyr::rename(
     returned_used_applicators = used_applicators,
     number_of_returned_applicators = number_applicators
   )

applicator <- 
  mae@colData |> as_tibble() |>
  select(pid, site, randomized, arm, ITT, mITT, PP) |>
  distinct() |>
  filter(randomized, mITT) |>
  left_join(tmp, by = c("pid")) 


applicator <- 
  applicator |> 
  mutate(
    returned_complete_set = (number_of_returned_applicators >= 7),
    return_category = case_when(
      (returned_used_applicators == "Yes") & returned_complete_set ~ "Returned complete set",
      (returned_used_applicators == "Yes") & !returned_complete_set ~ "Returned incomplete set",
      returned_used_applicators == "No" ~ "Did not return applicators",
      TRUE ~ "Missing"
    ) |> fct_infreq()
  ) 

applicator |> 
  dplyr::count(return_category) |> 
  mutate(`%` = round(n / sum(n), 3) * 100) |> 
  gt() |> 
  cols_label(return_category = "")

applicator <- 
  applicator |> 
  mutate(
    staining_done_category = 
      case_when(
        (number_of_returned_applicators == 0) | is.na(number_of_returned_applicators) ~ "No applicators returned",
        is.na(total_applicators_stained) ~ "No staining done",
        total_applicators_stained == number_of_returned_applicators ~ "All returned applicators stained",
        total_applicators_stained < number_of_returned_applicators ~ "Partial staining",
        total_applicators_stained > number_of_returned_applicators ~ "???",
        TRUE ~ "???"
      )
  )


```


```{r}

df_summ <- 
  df_summ |> 
  select(-contains("applicator")) |> 
  left_join(
    applicator |> 
      select(
        pid, returned_used_applicators, number_of_returned_applicators, 
             starts_with("applicator_stain")
        )
  ) |> 
  mutate(
    number_of_returned_applicators = number_of_returned_applicators |> replace_na(0),
    n_pos_applicator = applicator_stain_positive |> replace_na(0),
    n_undetermined_applicator = applicator_stain_indeterminant |> replace_na(0),
    p_pos_applicator = ifelse(number_of_returned_applicators == 0, 0.4, n_pos_applicator/number_of_returned_applicators),
    p_pos_or_undet_applicator = 
      ifelse(
        number_of_returned_applicators == 0, 0.4, 
        (n_pos_applicator + n_undetermined_applicator)/number_of_returned_applicators
      )
  )


```




```{r}
#| fig-height: 15
#| fig-width: 10
#| eval: false

df_summ |>
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = total_LBP |> log10()) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() +
  scale_fill_viridis_c(option = "F") + 
  ggnewscale::new_scale_fill() +
  geom_point(
    data = df_summ |> filter(randomized, daily_self_reported_adherence_sp == 1),
    aes(shape = returned_used_applicators, fill = p_pos_applicator), 
    size = 2, col = "white"
    ) +
  scale_fill_gradient2(
    "Prop. pos.\napplicators",
    high = "black", mid = "green", low = "yellow", 
    midpoint = 0.5, breaks = c(0, 1/2, 1)
    )  +
  scale_shape_manual(
    "Returned\napplicators", values = c(23, 21), 
    guide = guide_legend(override.aes = list(color = "black", fill = "black"))
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) 


df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = p_detected) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() + 
  scale_fill_viridis_c(
    "proportion of LBP strains\ndetected among expected", option = "G"
    ) +
  ggnewscale::new_scale_fill() +
  geom_point(
    data = df_summ |> filter(randomized, daily_self_reported_adherence_sp == 1),
    aes(shape = returned_used_applicators, fill = p_pos_applicator), 
    size = 2, col = "white"
  ) +
  scale_fill_gradient2(
    "Prop. pos.\napplicators",
    high = "black", mid = "green", low = "yellow", 
    midpoint = 0.5, breaks = c(0, 1/2, 1)
    ) +
  scale_shape_manual(
    "Returned\napplicators", values = c(23, 21), 
    guide = guide_legend(override.aes = list(color = "black", fill = "black"))
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) 


```




```{r}
#| fig-height: 15
#| fig-width: 10

df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = total_LBP |> log10()) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() +
  scale_fill_viridis_c(option = "F") + 
  ggnewscale::new_scale_fill() +
  geom_point(
    data = df_summ |> filter(randomized, daily_self_reported_adherence_sp == 1),
    aes(shape = returned_used_applicators, fill = p_pos_or_undet_applicator), 
    size = 2, col = "white"
    ) +
  scale_fill_gradient2(
    "Prop. pos. or ?\napplicators",
    high = "black", mid = "green", low = "yellow", 
    midpoint = 0.5, breaks = c(0, 1/2, 1)
    )  +
  scale_shape_manual(
    "Returned\napplicators", values = c(23, 21), 
    guide = guide_legend(override.aes = list(color = "black", fill = "black"))
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) 


df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = p_detected) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() + 
  scale_fill_viridis_c(
    "proportion of LBP strains\ndetected among expected", option = "G"
    ) +
  ggnewscale::new_scale_fill() +
  geom_point(
    data = df_summ |> filter(randomized, daily_self_reported_adherence_sp == 1),
    aes(shape = returned_used_applicators, fill = p_pos_or_undet_applicator), 
    size = 2, col = "white"
  ) +
  scale_fill_gradient2(
    "Prop. pos. or ?\napplicators",
    high = "black", mid = "green", low = "yellow", 
    midpoint = 0.5, breaks = c(0, 1/2, 1)
    ) +
  scale_shape_manual(
    "Returned\napplicators", values = c(23, 21), 
    guide = guide_legend(override.aes = list(color = "black", fill = "black"))
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) 


```


##### Exposed by qPCR

For each study day, we define "effective exposure by qPCR" as having at least 50% of expected LBP strains detected by qPCR and a total LBP load exceeding 10⁷ copies per swab.

```{r}

df_summ <- 
  df_summ |> 
  mutate(
    rel_ab = pmin(total_LBP/total_16S/2, 1),
    exposed = (p_detected >= 0.5) & (total_LBP > 10^7)
  )

```

```{r}
#| fig-height: 15
#| fig-width: 10

df_summ |> 
  filter(randomized) |> 
  ggplot() +
  aes(x = visit_code, y = pid, fill = exposed) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  geom_tile() +
  scale_fill_manual(
    "Exposed by qPCR",
    values = c("TRUE" = "blue", "FALSE" = "gray80"),
    labels = c("TRUE" = "Yes", "FALSE" = "No")
    ) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "bottom", legend.direction = "horizontal"
  ) +
  labs(caption = ".")
```


We summarize, for each participant, the number and proportion of days exposed by qPCR (daily home swabs) among expected dosing days.

```{r}
#| warning: false

df_summ <-
  df_summ |>
  filter(randomized) |>
  group_by(pid) |>
  mutate(
    first_day_of_self_report_sp = case_when(
      pid == "068100026" ~ 8,
      pid == "068100045" ~ NA_real_,
      TRUE ~ na_if(min(study_day[daily_self_reported_adherence_sp == 1], na.rm = TRUE), Inf)
    )
  ) |> 
  ungroup() |>
  filter(!(is.na(LC106_total) & is.na(daily_self_reported_adherence_sp))) |>
  mutate(
    dosing_days_sp = case_when(
      arm %in% c("LC106-7", "LC115", "LC106-o") ~
        study_day >= first_day_of_self_report_sp &
        study_day <= first_day_of_self_report_sp + 6,
      
      arm %in% "LC106-3" ~
        study_day >= first_day_of_self_report_sp &
        study_day <= first_day_of_self_report_sp + 2,
      
      arm %in% "Pl" ~ FALSE
    ),
    dosing_days_placebo = case_when(
      arm %in% c("LC106-7", "LC115", "LC106-o") ~ FALSE,
      
      arm %in% "LC106-3" ~
        study_day >= first_day_of_self_report_sp + 3 &
        study_day <= first_day_of_self_report_sp + 6,
      
      arm %in% "Pl" ~
        study_day >= first_day_of_self_report_sp &
        study_day <= first_day_of_self_report_sp + 6
    ),
    exposed_tmp = exposed & dosing_days_sp
  ) 

df_summ_pid <- df_summ |>
  group_by(arm, site, pid, ITT, mITT, PP) |>
  summarise(
    pid_exposed = any(exposed_tmp, na.rm = TRUE),
    n_days_exposed = sum(exposed_tmp, na.rm = TRUE),
    n_days_dosed = sum(dosing_days_sp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(prop_days_exposed = ifelse(n_days_exposed == 0, 0, n_days_exposed /
                                      n_days_dosed))

```

```{r}
df_summ_pid <- 
  df_summ_pid |> 
  mutate(
    pid_exposed_by_qPCR = pid_exposed,
    exposed_by_qPCR_lc106 = pid_exposed & str_detect(arm, "LC106"),
    exposed_by_qPCR_lc115 = pid_exposed & str_detect(arm, "LC115"),
    n_days_exposed_by_qPCR = n_days_exposed,
    n_days_exp_dosed = n_days_dosed
  )
```

```{r}
#| fig-height: 12
#| fig-width: 10

df_summ_pid |>
  mutate(pid_label = fct_reorder(pid, prop_days_exposed)) |>
  ggplot(aes(x = prop_days_exposed, y = pid_label)) +
  geom_segment(aes(xend = 0, yend = pid_label, col = pid_exposed_by_qPCR), linewidth = 0.8) +
  geom_point(aes(col = pid_exposed_by_qPCR), size = 1.5) +
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey80")) +
  labs(
    x     = "Proportion of dosing days exposed",
    y     = "",
    color = "Exposed by qPCR",
    title = "CTV05 exposure by qPCR — participant level"
  ) +
  theme(axis.text.y = element_text(size = 8))

```

```{r}

df_summ_pid |>
  dplyr::count(arm, site, pid_exposed_by_qPCR) |>
  ggplot(aes(x = arm, y = n, fill = pid_exposed_by_qPCR)) +
  geom_col(position = "fill") +
  facet_wrap(~ site) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey80")) +
  labs(
    x = "Arm", y = "Proportion of participants",
    fill = "Exposed by qPCR",
    title = "Proportion of participants exposed to CTV05 (by qPCR)"
  )
```
We also do a distinction between `exposed_by_lc106` and `exposed_by_lc115`.

```{r}
df_summ_pid |>
  mutate(
    exposed_by_qPCR_all = case_when(
      exposed_by_qPCR_lc106 == TRUE ~ "exposed_by_lc106",
      exposed_by_qPCR_lc115 == TRUE ~ "exposed_by_lc115",
      TRUE ~ NA_character_
    )
  ) |>
  dplyr::count(arm, site, exposed_by_qPCR_all) |>
  ggplot(aes(x = arm, y = n, fill = exposed_by_qPCR_all)) +
  geom_col(position = "fill") +
  facet_wrap(~ site) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_discrete(na.value = "lightgray") +
  labs(
    x = "Arm", y = "Proportion of participants",
    fill = "Exposed by qPCR",
    title = "Proportion of participants exposed to CTV05 (by qPCR)"
  )
```
# Documenting participants with delayed self-declared study product intake (for LC106-3 arm)

Before computing daily level variables, we want to assess whether there are participants that delayed their study product intake in LC106-3 arm. This will allow us to see if we can use `dosing_days_sp` and `dosing_days_placebo` to differentiate between placebo and study product doses in LC106-3 arm. If we see that there are some participants with delayed doses, we will define `dosing_days_sp` and `dosing_days_placebo` by counting the number of doses.

```{r}
#| fig-height: 4
#| fig-width: 6

df_summ |>
  filter(arm == "LC106-3") |>
  filter(study_day < 20) |>
  ggplot() +
  aes(x = study_day |> factor(), y = pid, color = daily_self_reported_adherence_sp |> as.character()) +
  geom_point() +
  scale_color_manual(
    "Self-declared study product",
    values = c("1" = "#F8766D"),
    na.value = "transparent"
  ) +
  labs(y = "Participant", x = "Study day")
  

```
From the figure above, we see that no participants have delayed study product intake. We can thus use `dosing_days_sp` and `dosing_days_placebo` to differentiate between placebo and study product doses in LC106-3 arm.

# Creating daily-level variables

Finally, we create daily variables, which represents daily exposure to the study product : `daily_exposure_to_lc106` represents the daily exposure to LC-106, `daily_exposure_to_lc115` represents the daily exposure to LC-115, and `daily_exposure_to_placebo` represents the daily exposure to the placebo product.

```{r}

calc_var <- 
mae@metadata$calc_vars_study_day |>
  select(pid, study_day, daily_self_reported_exposure_sp) |>
  left_join(df_summ_pid |> select(pid, exposed_by_qPCR_lc106, exposed_by_qPCR_lc115)) |>
  left_join(df_summ |> select(pid, study_day, arm, site, dosing_days_sp, dosing_days_placebo)) |>
  mutate(
         daily_exposure_to_lc106 = case_when(
            arm == "LC106-3" ~ daily_self_reported_exposure_sp * exposed_by_qPCR_lc106 * dosing_days_sp,
            TRUE ~ daily_self_reported_exposure_sp * exposed_by_qPCR_lc106),
         
         daily_exposure_to_lc115 = daily_self_reported_exposure_sp * exposed_by_qPCR_lc115,
         
         daily_exposure_to_placebo = case_when(
            arm == "Pl" ~ daily_self_reported_exposure_sp,
            arm == "LC106-3" ~ daily_self_reported_exposure_sp * dosing_days_placebo,
            TRUE ~ NA
         ))
```


```{r}
#| fig-height: 10
#| fig-width: 8


calc_var |>
  filter(!is.na(study_day), !is.na(site)) |>
  mutate(
    daily_exposure_to = case_when(
      daily_exposure_to_lc106 > 0 ~ "lc106",
      daily_exposure_to_lc115 > 0 ~ "lc115",
      daily_exposure_to_placebo > 0 ~ "placebo"
    ),
    study_day = factor(study_day)
  ) |>
  ggplot() +
  aes(x = study_day, y = pid, color = daily_exposure_to) +
  geom_point() + 
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Daily exposure to",
    values = c("lc106" = "#F8766D", "lc115" = "#00BA38", "placebo" = "#619CFF"),
    na.value = "transparent"
  ) +
  labs(y = "Participant", x = "Study day")


```

In the plot above, we can see that, for participant `068200387` in arm LC106-3, it is indicated that the placebo product was taken, but not the LBP product. 

In this case, we will consider that, if the participant said they have taken all their doses, but no LBP was detected, we will consider that no product was taken at all, either LBP or placebo product.

```{r}

calc_var <- 
  calc_var |>
  mutate(daily_exposure_to_placebo = ifelse(pid == "068200387", 0, daily_exposure_to_placebo))

```

```{r}
#| fig-height: 10
#| fig-width: 8


calc_var |>
  filter(!is.na(study_day), !is.na(site)) |>
  mutate(
    daily_exposure_to = case_when(
      daily_exposure_to_lc106 > 0 ~ "lc106",
      daily_exposure_to_lc115 > 0 ~ "lc115",
      daily_exposure_to_placebo > 0 ~ "placebo"
    ),
    study_day = factor(study_day)
  ) |>
  ggplot() +
  aes(x = study_day, y = pid, color = daily_exposure_to) +
  geom_point() + 
  ggh4x::facet_nested(arm + site ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(
    "Daily exposure to",
    values = c("lc106" = "#F8766D", "lc115" = "#00BA38", "placebo" = "#619CFF"),
    na.value = "transparent"
  ) +
  labs(y = "Participant", x = "Study day")


```

# Saving pid-level and daily-level variables

We add these variables to the calculated variables table in the MAE metadata.

```{r}

selected_vars_pid <- c("pid_exposed_by_qPCR", "n_days_exposed_by_qPCR", "n_days_exp_dosed")
selected_vars_study_day <- c("daily_exposure_to_lc106", "daily_exposure_to_lc115", "daily_exposure_to_placebo")

mae@metadata$calc_vars_pid <- 
  mae@metadata$calc_vars_pid |> 
  select(-any_of(selected_vars_pid)) |> 
  left_join(
    df_summ_pid |> 
      select(pid, all_of(selected_vars_pid)),
    by = join_by(pid)
  )

mae@metadata$calc_vars_study_day <- 
  mae@metadata$calc_vars_study_day |> 
   select(-any_of(selected_vars_study_day)) |>
  left_join(
    calc_var |> 
      select(pid, study_day, all_of(selected_vars_study_day)),
    by = join_by(pid, study_day)
  )
  
```

# Saving the `MultiAssayExperiment` objects

The unblinded and augmented `MAE` object is saved to disk as an `.rds` file that can be loaded for downstream analyses.

```{r}

saveRDS(
  mae,
  str_c(
    get_05_output_dir(),
    "mae_full_", today() |> str_remove_all("-"), ".rds"
  )
)

```



