---
title: "VIBRANT subsetting MAE"
author: Laura Symul, Laura Vermeren, Lara Wautier, Céline Bugli
date: today
format: 
  html:
    code-fold: true
    toc: true
    toc-location: left
    theme: journal
    embed-resources: true
execute:
  cache: true # refresh true false
project:
  execute-dir: project
editor: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
#| warning: false
#| cache: false

library(tidyverse)
library(magrittr)
library(gt)
library(patchwork)
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(tidySummarizedExperiment)
library(brms)
library(gtsummary)
library(labelled)
library(scales)
# library(gridExtra)
library(ggpubr)
library(ggh4x)
# library(mia) # BiocManager::install("mia")

#theme_set(theme_light())
tmp <- fs::dir_map("R scripts", source)
rm(tmp)

```

```{r source-utils}
#| warning: false
#| cache: false
tmp <- fs::dir_map("../VIBRANT-99-utils/R/", source)
rm(tmp)
```


:::callout-note
The purpose of this document is to subset the master `MAE` object into smaller, analysis-specific `MAE` objects that contain only the data required for the respective analyses.  

These curated `MAE` objects are intended for public release, and therefore do not include sensitive or unnecessary information.

At the end of this document, several `MAE` objects are created:

- `MAE_1`: contains all necessary data for the first VIBRANT paper 1 (primary outcome) [cite paper].  
- `MAE_2`: contains all necessary data for [describe analysis or paper].  
- ...
:::


```{r loading-mae}
#| eval: true

mae_file <- 
  fs::dir_ls(str_c(data_dir(), "04 unblinded MAEs/"), regexp = "mae_full_.*\\.rds$") |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1)

cat(str_c("Loading the RDS file containing the unblinded VIBRANT MAE: /", mae_file |> str_remove(data_dir()), "\n"))

mae <- readRDS(mae_file)

rm(mae_file)

fs::dir_ls(str_c(data_dir(), "01 Preprocessed and QCed/"), regexp = "exposures_.*\\.Rdata$") |> 
  sort(decreasing = TRUE) |>
  magrittr::extract(1) |> 
  load()


mae

```


## MAE 1 

`MAE_1` contains all necessary information for the primary outcome analyses and contains only the randomized participants. 

-   `participant_crf_merged` table stored in the metadata of the MAE with the following variable : site, age, race, ethn, cut_size_meals, eat_less, hungry_did_not_eat, sexual_partners_lifetime, sexual_partners_past_month, contraception_method, used_applicators, number_applicators, applicator_stain_positive, applicator_stain_negative, applicator_stain_indeterminant, total_applicators_stained, proportion_positive_applicators, total

-   `visits_crfs_merged` table stored in the metadata of the MAE contains : visit_code, study_day, the participant's partner's gender (past_week_sex_partner), insert_study_product_time, nugent_total_score

-   `coldata` contains the population flag (`ITT`, `mITT`, `PP`), `PIPV`, if participants is randomized or not (`randomized`), the site, location, arm, visit_code, visit_type, study_week

The `mae_1`will contains the following assay with the following informations: 

-   `col_LBP_mg` SE contains : .feature, .sample, outcome
-   `col_LBP_qPCR` SE contains : .feature, .sample,  outcome
-   `col_crispatus_mg` SE contains : .feature, .sample, outcome
-   `mg` SE contains the following : .feature, .sample, uid, LBP, poor_quality_mg_data, rel_ab, rel_ab_bact and the taxonomic table 
-   `qPCR` SE contains : uid, .feature, copies_per_swab_med, taxon_label, LBP, strain_origin
-   `qPCR_daily` SE contains : uid, .feature, copies_per_swab_med, taxon_label, LBP, strain_origin


```{r}

# Randomized participants 
mae_randomized <- mae[, !is.na(colData(mae)$randomized) & colData(mae)$randomized == TRUE]

# subsetting only necessary SE
se_names <- c("col_LBP_mg", "col_LBP_qPCR", "col_crispatus_mg", "mg", "qPCR", "qPCR_daily")
selected_experiments <- experiments(mae_randomized)[se_names]

mae_1 <- MultiAssayExperiment(
  experiments = selected_experiments,
  colData = colData(mae_randomized),
  metadata = metadata(mae_randomized)
)

# Meta data 

## filtering partcipant_crfs_merged
vars_participant <- c("pid", "site", "age", "race", "ethn", "cut_size_meals", "eat_less", "hungry_did_not_eat",
                      "sexual_partners_lifetime", "sexual_partners_past_month", "contraception_method",
                      "used_applicators", "number_applicators", "applicator_stain_positive",
                      "applicator_stain_negative", "applicator_stain_indeterminant",
                      "total_applicators_stained", "proportion_positive_applicators", "total")

## filtering visits_crfs_merged
vars_visits <- c("pid", "visit_code", "study_day", 
                 "past_week_sex_partner", "insert_study_product_time", "nugent_total_score")

## new metadata
new_metadata <- 
  list(
    participant_crfs_merged = metadata(mae_1)$participant_crfs_merged[, vars_participant],
    visits_crfs_merged = metadata(mae_1)$visits_crfs_merged[, vars_visits],
    exposures = metadata(mae_1)$exposures  
)

metadata(mae_1) <- new_metadata


# filtering coldata
vars_coldata <- c("ITT", "mITT", "PP", "PIPV", "randomized", 
                  "site", "location", "arm", 
                  "visit_code", "visit_type", "study_week")
colData(mae_randomized) <- colData(mae_randomized)[, vars_coldata]


# filtering the SE 

# 1. mg 
assays(mae_1[["mg"]]) <- assays(mae_1[["mg"]])[c("rel_ab_all", "rel_ab_bact", "rel_ab")]
colData(mae_1[["mg"]]) <- colData(mae_1[["mg"]])[, c("uid", "poor_quality_mg_data")]

# 2. qPCR
assays(mae[["qPCR"]]) <- assays(mae[["qPCR"]])[c("copies_per_swab_med")]
colData(mae[["qPCR"]]) <- colData(mae[["qPCR"]])[, c("uid", "qpcr_sample_type")]
rowData(mae[["qPCR"]]) <- rowData(mae[["qPCR"]])[, c("taxon_label", "LBP", "strain_origin")]

# 3. qPCR_daily
assays(mae[["qPCR_daily"]]) <- assays(mae[["qPCR_daily"]])[c("copies_per_swab_med")]
colData(mae[["qPCR_daily"]]) <- colData(mae[["qPCR_daily"]])[, c("uid", "qpcr_sample_type")]
rowData(mae[["qPCR_daily"]]) <- rowData(mae[["qPCR_daily"]])[, c("taxon_label", "LBP", "strain_origin")]

# 4. col_LBP_mg, col_LBP_qPCR, col_crispatus_mg : no changes

```


## Save MAEs 

```{r}

saveRDS(
  mae_1, 
  str_c(
    get_05_output_dir(),  
    "mae_1_", today() |> str_remove_all("-"), ".rds"
  )
)


```

